---
title: "Olink_2_cohorts"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Olink_2_cohorts"

# Load libraries
library(EnhancedVolcano)
library(here)
library(enrichplot)
library(readr)
library(readxl)
library(Matrix)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(pheatmap)
library(png)
library(gridExtra)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(qs2)
library(statmod)
library(rtracklayer)
library(png)
library(reactable)
library(UpSetR)
library(limma)
library(igraph)
library(gridExtra) 
library(ggpubr)
library(DESeq2)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(UpSetR)
library(ComplexHeatmap)
library(survival)
library(broom)
library(circlize)
library(variancePartition)

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

## Import data

Olink data Zurich and Poland

```{r read data}
data_z <- qs_read(here::here("data", "o37198_2025-02-27_olink_SummarizedExperiment.qs2"))

clin_z <- read_excel(here::here("data", "Metadata_Olink_analysis.xlsx"))

meta_z <- colData(data_z) %>% data.frame()
isOutlierColumn <- meta_z$Plate == "P0194" & meta_z$Col %in% c("3", "4")
table(isOutlierColumn)
use <- meta_z$SampleType %in% c("Serum") & meta_z$Group %in% c("SSc-PAH", "SSc-noPAH") & !isOutlierColumn
table(use)
data_z <- data_z[ , use]
meta_z <- meta_z[use, ]
rownames(meta_z) <- gsub("_P0194", "", rownames(meta_z))
meta_z$Barcode <- rownames(meta_z)
rownames(data_z) <- data_z@elementMetadata@listData[["Gene"]]
rownames(data_z@assays@data@listData[["npx"]]) <- data_z@elementMetadata@listData[["Gene"]]
data_z <- assay(data_z, "npx")
data_z <- as.data.frame(data_z[-c(1:3), ])
colnames(data_z) <- rownames(meta_z)

clin_z <- clin_z %>%
  dplyr::filter(Barcode %in% colnames(data_z))
dim(data_z)
table(clin_z$Group)


data_p <- read.delim(here::here("data", "OLINK_PL", "quant.xls"), sep = "\t", check.names = FALSE)

clin_p <- read_excel(here::here("data", "OLINK_PL", "Metadata_Olink_PL.xlsx"))

rownames(data_p) <- sub("_.*", "", data_p$SampleID)
data_p <- data_p[, -1]
keep <- intersect(colnames(data_p), clin_p$Barcode)
data_p <- data_p[, keep, drop = FALSE]
dim(data_p)
table(clin_p$Group)
```

```{r, fig.height=8, fig.width=10.8}
barcodes <- colnames(data_z)

grp <- clin_z$Group[match(barcodes, clin_z$Barcode)]
cmp <- clin_z$COMPERA_4strata[match(barcodes, clin_z$Barcode)]
cmp <- factor(cmp, levels = c("low","intermediate-low","intermediate-high","high"))
sex <- clin_z$Sex[match(barcodes, clin_z$Barcode)]

# --- colors ---
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "low"               = "#9CA3AF",   # grey
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)
cols_sex <- c("F"="green4", "M"="yellow2")

# --- z-score per protein (row) ---
matz <- t(scale(t(data_z)))
matz[is.nan(matz)] <- 0

# heatmap colors
hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

# column annotations (COMPERA NAs shown as light grey)
ha <- HeatmapAnnotation(
  Group   = grp,
  COMPERA = cmp,
  Sex = sex,
  col = list(Group = cols_group, COMPERA = cols_compera, Sex = cols_sex),
  na_col = "#E5E7EB",
  annotation_legend_param = list(
    Group   = list(title = "Group"),
    COMPERA = list(title = "Risk (COMPERA)"),
    Sex   = list(title = "Sex")
  )
)

# --- draw (all proteins, no row names, default clustering, no group split) ---
Heatmap(
  matz,
  name = "Z-score",
  col = hm_cols,
  top_annotation = ha,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 9),
  row_title = NULL,
  column_title = NULL
) |>
  draw(heatmap_legend_side = "right", annotation_legend_side = "right")
```

```{r, fig.height=8, fig.width=10.8}
barcodes <- colnames(data_p)

grp <- clin_p$Group[match(barcodes, clin_p$Barcode)]
cmp <- clin_p$Compera_4strata[match(barcodes, clin_p$Barcode)]
cmp <- factor(cmp, levels = c("Low","Intermediate-low","Intermediate-high","High"))
ori <- clin_p$Origin[match(barcodes, clin_p$Barcode)]
sex <- clin_p$Sex[match(barcodes, clin_p$Barcode)]

# --- colors ---
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "Low"               = "#9CA3AF",   # grey
  "Intermediate-low"  = "#DDB7B2",
  "Intermediate-high" = "#C7685F",
  "High"              = "#A63A2A"
)
cols_origin <- c("Krakow"="orange", "Lublin"="purple")
cols_sex <- c("F"="green4", "M"="yellow2")

# --- z-score per protein (row) ---
matz <- t(scale(t(data_p)))
matz[is.nan(matz)] <- 0

# heatmap colors
hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

# column annotations (COMPERA NAs shown as light grey)
ha <- HeatmapAnnotation(
  Group   = grp,
  COMPERA = cmp,
  Origin = ori,
  Sex = sex,
  col = list(Group = cols_group, COMPERA = cols_compera, Origin = cols_origin, Sex = cols_sex),
  na_col = "#E5E7EB",
  annotation_legend_param = list(
    Group   = list(title = "Group"),
    COMPERA = list(title = "Risk (COMPERA)"),
    Origin = list(title = "Origin"),
    Sex   = list(title = "Sex")
  )
)

# --- draw (all proteins, no row names, default clustering, no group split) ---
Heatmap(
  matz,
  name = "Z-score",
  col = hm_cols,
  top_annotation = ha,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 9),
  row_title = NULL,
  column_title = NULL
) |>
  draw(heatmap_legend_side = "right", annotation_legend_side = "right")
```

```{r heatmap_z_top100var, fig.height=10.8, fig.width=10.8}
barcodes <- colnames(data_z)

grp <- clin_z$Group[match(barcodes, clin_z$Barcode)]
cmp <- clin_z$COMPERA_4strata[match(barcodes, clin_z$Barcode)]
cmp <- factor(cmp, levels = c("low","intermediate-low","intermediate-high","high"))
sex <- clin_z$Sex[match(barcodes, clin_z$Barcode)]

# --- colors ---
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "low"               = "#9CA3AF",
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)
cols_sex <- c("F"="green4", "M"="yellow2")

# --- pick top 100 variable proteins (rows) ---
vars <- apply(data_z, 1, var, na.rm = TRUE)
top100 <- names(sort(vars, decreasing = TRUE))[1:100]
mat_top <- as.matrix(data_z[top100, , drop = FALSE])

# --- z-score per protein (row) ---
matz <- t(scale(t(mat_top)))
matz[is.nan(matz)] <- 0

hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

ha <- HeatmapAnnotation(
  Group   = grp,
  COMPERA = cmp,
  Sex     = sex,
  col = list(Group = cols_group, COMPERA = cols_compera, Sex = cols_sex),
  na_col = "#E5E7EB",
  annotation_legend_param = list(
    Group   = list(title = "Group"),
    COMPERA = list(title = "Risk (COMPERA)"),
    Sex     = list(title = "Sex")
  )
)

Heatmap(
  matz,
  name = "Z-score",
  col = hm_cols,
  top_annotation = ha,
  show_row_names = TRUE,              
  row_names_gp = gpar(fontsize = 7),
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 8.5),
  row_title = "Top 100 variable proteins",
  column_title = NULL
) |>
  draw(heatmap_legend_side = "right", annotation_legend_side = "right")
```

```{r heatmap_p_top100var, fig.height=10.88, fig.width=10.8}
stopifnot(exists("data_p"), exists("clin_p"))

barcodes <- colnames(data_p)

grp <- clin_p$Group[match(barcodes, clin_p$Barcode)]
cmp <- clin_p$Compera_4strata[match(barcodes, clin_p$Barcode)]
cmp <- factor(cmp, levels = c("Low","Intermediate-low","Intermediate-high","High"))
ori <- clin_p$Origin[match(barcodes, clin_p$Barcode)]
sex <- clin_p$Sex[match(barcodes, clin_p$Barcode)]

# --- colors ---
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "Low"               = "#9CA3AF",
  "Intermediate-low"  = "#DDB7B2",
  "Intermediate-high" = "#C7685F",
  "High"              = "#A63A2A"
)
cols_origin <- c("Krakow"="orange", "Lublin"="purple")
cols_sex <- c("F"="green4", "M"="yellow2")

# --- pick top 100 variable proteins (rows) ---
vars <- apply(data_p, 1, var, na.rm = TRUE)
top100 <- names(sort(vars, decreasing = TRUE))[1:100]
mat_top <- as.matrix(data_p[top100, , drop = FALSE])

# --- z-score per protein (row) ---
matz <- t(scale(t(mat_top)))
matz[is.nan(matz)] <- 0

hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

ha <- HeatmapAnnotation(
  Group   = grp,
  COMPERA = cmp,
  Origin  = ori,
  Sex     = sex,
  col = list(Group = cols_group, COMPERA = cols_compera, Origin = cols_origin, Sex = cols_sex),
  na_col = "#E5E7EB",
  annotation_legend_param = list(
    Group   = list(title = "Group"),
    COMPERA = list(title = "Risk (COMPERA)"),
    Origin  = list(title = "Origin"),
    Sex     = list(title = "Sex")
  )
)

Heatmap(
  matz,
  name = "Z-score",
  col = hm_cols,
  top_annotation = ha,
  show_row_names = TRUE,              # set FALSE if you want to hide protein names
  row_names_gp = gpar(fontsize = 7),
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 8.5),
  row_title = "Top 100 variable proteins",
  column_title = NULL
) |>
  draw(heatmap_legend_side = "right", annotation_legend_side = "right")
```
```{r pca_z_group, fig.height=3.5, fig.width=5.4}
# PCA (samples as rows)
pca <- prcomp(t(as.matrix(data_z)), scale. = TRUE)
pvar <- (pca$sdev^2) / sum(pca$sdev^2) * 100

df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  Group = clin_z$Group[match(rownames(pca$x), clin_z$Barcode)]
)

cols_group <- c("SSc-noPAH" = "#004C8C", "SSc-PAH" = "#A63A2A")

ggplot(df, aes(PC1, PC2, color = Group)) +
  stat_ellipse(aes(fill = Group), type = "norm", level = 0.95,
               geom = "polygon", alpha = 0.10, color = NA) +
  stat_ellipse(type = "norm", level = 0.95, linewidth = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = cols_group) +
  scale_fill_manual(values = cols_group) +
  labs(
    x = paste0("PC1 (", round(pvar[1], 1), "%)"),
    y = paste0("PC2 (", round(pvar[2], 1), "%)")
  ) +
  theme_minimal()
```

```{r pca_z_compera, fig.height=3.5, fig.width=5.4}
# --- subset to SSc-PAH only ---
keep <- clin_z$Group == "SSc-PAH"
barcodes_keep <- clin_z$Barcode[keep]

data_sub <- data_z[, colnames(data_z) %in% barcodes_keep, drop = FALSE]
clin_sub <- clin_z[match(colnames(data_sub), clin_z$Barcode), ]

# PCA
pca <- prcomp(t(as.matrix(data_sub)), scale. = TRUE)
pvar <- (pca$sdev^2) / sum(pca$sdev^2) * 100

df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  COMPERA = factor(
    clin_sub$COMPERA_4strata,
    levels = c("low","intermediate-low","intermediate-high","high")
  )
)

cols_compera_z <- c(
  "low"               = "#9CA3AF",
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)

ggplot(df, aes(PC1, PC2, color = COMPERA)) +
  stat_ellipse(aes(fill = COMPERA), type = "norm", level = 0.90,
               geom = "polygon", alpha = 0.10, color = NA) +
  stat_ellipse(type = "norm", level = 0.90, linewidth = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = cols_compera_z, name = "Risk (COMPERA)") +
  scale_fill_manual(values = cols_compera_z,  name = "Risk (COMPERA)") +
  labs(
    x = paste0("PC1 (", round(pvar[1], 1), "%)"),
    y = paste0("PC2 (", round(pvar[2], 1), "%)")
  ) +
  theme_minimal() +
  guides(fill = "none")
```

```{r pca_p_group, fig.height=3.5, fig.width=5.4}
pca <- prcomp(t(as.matrix(data_p)), scale. = TRUE)
pvar <- (pca$sdev^2) / sum(pca$sdev^2) * 100

df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  Group = clin_p$Group[match(rownames(pca$x), clin_p$Barcode)]
)

cols_group <- c("SSc-noPAH" = "#004C8C", "SSc-PAH" = "#A63A2A")

ggplot(df, aes(PC1, PC2, color = Group)) +
  stat_ellipse(aes(fill = Group), type = "norm", level = 0.95,
               geom = "polygon", alpha = 0.10, color = NA) +
  stat_ellipse(type = "norm", level = 0.95, linewidth = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = cols_group) +
  scale_fill_manual(values = cols_group) +
  labs(
    x = paste0("PC1 (", round(pvar[1], 1), "%)"),
    y = paste0("PC2 (", round(pvar[2], 1), "%)")
  ) +
  theme_minimal()
```

```{r pca_p_compera, fig.height=3.5, fig.width=5.4}
# --- subset to SSc-PAH only ---
keep <- clin_p$Group == "SSc-PAH"
barcodes_keep <- clin_p$Barcode[keep]

data_sub <- data_p[, colnames(data_p) %in% barcodes_keep, drop = FALSE]
clin_sub <- clin_p[match(colnames(data_sub), clin_p$Barcode), ]

# PCA
pca <- prcomp(t(as.matrix(data_sub)), scale. = TRUE)
pvar <- (pca$sdev^2) / sum(pca$sdev^2) * 100

df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  COMPERA = factor(
    clin_sub$Compera_4strata,
    levels = c("Low","Intermediate-low","Intermediate-high","High")
  )
)

cols_compera_p <- c(
  "Low"               = "#9CA3AF",
  "Intermediate-low"  = "#DDB7B2",
  "Intermediate-high" = "#C7685F",
  "High"              = "#A63A2A"
)

ggplot(df, aes(PC1, PC2, color = COMPERA)) +
  stat_ellipse(aes(fill = COMPERA), type = "norm", level = 0.90,
               geom = "polygon", alpha = 0.10, color = NA) +
  stat_ellipse(type = "norm", level = 0.90, linewidth = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = cols_compera_p, name = "Risk (COMPERA)") +
  scale_fill_manual(values = cols_compera_p,  name = "Risk (COMPERA)") +
  labs(
    x = paste0("PC1 (", round(pvar[1], 1), "%)"),
    y = paste0("PC2 (", round(pvar[2], 1), "%)")
  ) +
  theme_minimal() +
  guides(fill = "none")
```

## Variance partitioning

```{r vp_z_all, fig.height=4.5, fig.width=6.5}
# expression: proteins x samples
expr <- as.matrix(data_z)

# metadata aligned to columns of expr
meta <- clin_z |>
  dplyr::distinct(Barcode, .keep_all = TRUE) |>
  dplyr::filter(Barcode %in% colnames(expr)) |>
  tibble::column_to_rownames("Barcode")

expr <- expr[, rownames(meta), drop = FALSE]

# variables for this model (Zurich: NO Origin in all-samples model)
vars <- c("Group", "Age", "Sex", "Sample_age_days")
meta <- meta |>
  dplyr::mutate(
    Group = factor(Group),
    Sex   = factor(Sex)
  )

# keep complete cases for model variables
n_before <- nrow(meta)

meta <- meta[complete.cases(meta[, vars]), , drop = FALSE]
expr <- expr[, rownames(meta), drop = FALSE]

n_after <- nrow(meta)
message("Zurich (all): ", n_after, " / ", n_before, " samples used")

form <- ~ Group + Age + Sex + Sample_age_days

vp <- fitExtractVarPartModel(expr, form, meta)
colnames(vp) <- gsub("`", "", colnames(vp))

# quick summaries
vp_mean <- sort(colMeans(vp), decreasing = TRUE)
df <- data.frame(variable = names(vp_mean), value = vp_mean)

ggplot(df, aes(x = reorder(variable, -value), y = value)) +
  geom_col() +
  labs(x = NULL, y = "Mean fraction of variance explained",
       title = "Zurich: all samples") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotVarPart(vp) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vp_z_pah, fig.height=4.5, fig.width=6.5}
expr <- as.matrix(data_z)

meta <- clin_z |>
  dplyr::distinct(Barcode, .keep_all = TRUE) |>
  dplyr::filter(Barcode %in% colnames(expr)) |>
  dplyr::filter(Group == "SSc-PAH") |>
  tibble::column_to_rownames("Barcode")

expr <- expr[, rownames(meta), drop = FALSE]

# requested covariates (Origin not available in Zurich -> not included)
vars <- c("Age", "Sex", "TAPSE_sPAP", "Creatinine", "ILDonHRCT")
meta <- meta |>
  dplyr::mutate(
    Sex = factor(Sex),
    ILDonHRCT = factor(ILDonHRCT)
  )

n_before <- nrow(meta)

meta <- meta[complete.cases(meta[, vars]), , drop = FALSE]
expr <- expr[, rownames(meta), drop = FALSE]

n_after <- nrow(meta)
message("Zurich (all): ", n_after, " / ", n_before, " samples used")

form <- ~ Age + Sex  + TAPSE_sPAP + Creatinine + ILDonHRCT + mPAP_mmHg 

vp <- fitExtractVarPartModel(expr, form, meta)
colnames(vp) <- gsub("`", "", colnames(vp))

vp_mean <- sort(colMeans(vp), decreasing = TRUE)
df <- data.frame(variable = names(vp_mean), value = vp_mean)

ggplot(df, aes(x = reorder(variable, -value), y = value)) +
  geom_col() +
  labs(x = NULL, y = "Mean fraction of variance explained",
       title = "Zurich: SSc-PAH only") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotVarPart(vp) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vp_p_all, fig.height=4.5, fig.width=6.5}
library(variancePartition)

expr <- as.matrix(data_p)

meta <- clin_p |>
  dplyr::distinct(Barcode, .keep_all = TRUE) |>
  dplyr::filter(Barcode %in% colnames(expr)) |>
  tibble::column_to_rownames("Barcode")

expr <- expr[, rownames(meta), drop = FALSE]

# requested covariates
vars <- c("Group", "Age", "Sex", "Sample_age", "Origin")
meta <- meta |>
  dplyr::mutate(
    Group  = factor(Group),
    Sex    = factor(Sex),
    Origin = factor(Origin)
  )

n_before <- nrow(meta)

meta <- meta[complete.cases(meta[, vars]), , drop = FALSE]
expr <- expr[, rownames(meta), drop = FALSE]

n_after <- nrow(meta)
message("Zurich (all): ", n_after, " / ", n_before, " samples used")


form <- ~ Group + Age + Sex + Sample_age + Origin

vp <- fitExtractVarPartModel(expr, form, meta)
colnames(vp) <- gsub("`", "", colnames(vp))

vp_mean <- sort(colMeans(vp), decreasing = TRUE)
df <- data.frame(variable = names(vp_mean), value = vp_mean)

ggplot(df, aes(x = reorder(variable, -value), y = value)) +
  geom_col() +
  labs(x = NULL, y = "Mean fraction of variance explained",
       title = "Poland: all samples") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotVarPart(vp) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vp_p_pah, fig.height=4.5, fig.width=6.5}
library(variancePartition)

expr <- as.matrix(data_p)

meta <- clin_p |>
  dplyr::distinct(Barcode, .keep_all = TRUE) |>
  dplyr::filter(Barcode %in% colnames(expr)) |>
  dplyr::filter(Group == "SSc-PAH") |>
  tibble::column_to_rownames("Barcode")

expr <- expr[, rownames(meta), drop = FALSE]

# requested covariates
vars <- c("Age", "Sex", "TAPSE_sPAP", "Creatinine", "ILD")
meta <- meta |>
  dplyr::mutate(
    Sex    = factor(Sex),
    ILD    = factor(ILD)
  )

n_before <- nrow(meta)

meta <- meta[complete.cases(meta[, vars]), , drop = FALSE]
expr <- expr[, rownames(meta), drop = FALSE]

n_after <- nrow(meta)
message("Zurich (all): ", n_after, " / ", n_before, " samples used")

form <- ~ Age + Sex + Creatinine + ILD + TAPSE_sPAP

vp <- fitExtractVarPartModel(expr, form, meta)
colnames(vp) <- gsub("`", "", colnames(vp))

vp_mean <- sort(colMeans(vp), decreasing = TRUE)
df <- data.frame(variable = names(vp_mean), value = vp_mean)

ggplot(df, aes(x = reorder(variable, -value), y = value)) +
  geom_col() +
  labs(x = NULL, y = "Mean fraction of variance explained",
       title = "Poland: SSc-PAH only") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotVarPart(vp) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



## Differential expression analysis

```{r de_group_zurich}
# --- prepare metadata aligned to expression ---
meta_z_de <- clin_z %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::mutate(
    Group = factor(Group, levels = c("SSc-noPAH", "SSc-PAH")),
    Sex   = factor(Sex),
    Age   = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

# keep only complete cases for model variables
meta_z_de <- meta_z_de[complete.cases(meta_z_de[, c("Group", "Age", "Sex")]), , drop = FALSE]

expr_z_de <- as.matrix(data_z[, rownames(meta_z_de), drop = FALSE])

# --- design: Group + Age + Sex ---
design_z <- model.matrix(~ Group + Age + Sex, data = meta_z_de)

# rescale numeric covariate(s) like in your code
if ("Age" %in% colnames(design_z)) design_z[, "Age"] <- rescale(design_z[, "Age"])

# --- fit limma ---
fit_z <- lmFit(expr_z_de, design_z)
fit_z <- eBayes(fit_z)

# coefficient name for GroupSSc-PAH (vs reference SSc-noPAH)
coef_group_z <- grep("^Group", colnames(design_z), value = TRUE)[1]

DE_z <- topTable(fit_z, coef = coef_group_z, number = Inf, adjust.method = "BH")
DE_z$Gene <- rownames(DE_z)

DE_z_signif <- DE_z %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

# write outputs (same style as your loop, but single comparison)
write.csv(DE_z,
          file = here::here(output_dir_data, "DE_Results_Group_SScPAH_vs_noPAH_Zurich.csv"),
          quote = FALSE, row.names = FALSE)
write.csv(DE_z_signif,
          file = here::here(output_dir_data, "DE_Significant_Group_SScPAH_vs_noPAH_Zurich.csv"),
          quote = FALSE, row.names = FALSE)

# store for volcano chunk
n_z <- ncol(expr_z_de)
```

```{r de_group_poland}
meta_p_de <- clin_p %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::mutate(
    Group = factor(Group, levels = c("SSc-noPAH", "SSc-PAH")),
    Sex   = factor(Sex),
    Age   = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_p_de <- meta_p_de[complete.cases(meta_p_de[, c("Group", "Age", "Sex")]), , drop = FALSE]

expr_p_de <- as.matrix(data_p[, rownames(meta_p_de), drop = FALSE])

design_p <- model.matrix(~ Group + Age + Sex, data = meta_p_de)
if ("Age" %in% colnames(design_p)) design_p[, "Age"] <- rescale(design_p[, "Age"])

fit_p <- lmFit(expr_p_de, design_p)
fit_p <- eBayes(fit_p)

coef_group_p <- grep("^Group", colnames(design_p), value = TRUE)[1]

DE_p <- topTable(fit_p, coef = coef_group_p, number = Inf, adjust.method = "BH")
DE_p$Gene <- rownames(DE_p)

DE_p_signif <- DE_p %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

write.csv(DE_p,
          file = here::here(output_dir_data, "DE_Results_Group_SScPAH_vs_noPAH_Poland.csv"),
          quote = FALSE, row.names = FALSE)
write.csv(DE_p_signif,
          file = here::here(output_dir_data, "DE_Significant_Group_SScPAH_vs_noPAH_Poland.csv"),
          quote = FALSE, row.names = FALSE)

n_p <- ncol(expr_p_de)
```

```{r, fig.height=4, fig.width=3, warning=FALSE}
keyvals <- rep("grey", nrow(DE_z))
names(keyvals) <- rep("Not significant", nrow(DE_z))

keyvals[DE_z$logFC > 0.5 & DE_z$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Increased"

keyvals[DE_z$logFC < -0.5 & DE_z$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

volcano_z <- EnhancedVolcano(
  DE_z,
  lab = rownames(DE_z),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Zurich cohort\nSSc-PAH vs SSc-noPAH\nn = ", n_z)
) +
  theme(legend.position = "none")

volcano_z
```

```{r, fig.height=4, fig.width=3, warning=FALSE}
keyvals <- rep("grey", nrow(DE_p))
names(keyvals) <- rep("Not significant", nrow(DE_p))

keyvals[DE_p$logFC > 0.5 & DE_p$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Increased"

keyvals[DE_p$logFC < -0.5 & DE_p$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

volcano_p <- EnhancedVolcano(
  DE_p,
  lab = rownames(DE_p),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Poland cohort\nSSc-PAH vs SSc-noPAH\nn = ", n_p)
) +
  theme(legend.position = "none")

volcano_p
```

```{r de_tapse_zurich_pah}
meta_z_de <- clin_z %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    Sex = factor(Sex),
    Age = as.numeric(Age),
    TAPSE_sPAP = as.numeric(TAPSE_sPAP)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_z_de <- meta_z_de[
  complete.cases(meta_z_de[, c("TAPSE_sPAP", "Age", "Sex")]),
  , drop = FALSE
]

expr_z_de <- as.matrix(data_z[, rownames(meta_z_de), drop = FALSE])

design_z <- model.matrix(~ TAPSE_sPAP + Age + Sex, data = meta_z_de)

design_z[, "TAPSE_sPAP"] <- rescale(design_z[, "TAPSE_sPAP"])
design_z[, "Age"]        <- rescale(design_z[, "Age"])

fit_z <- lmFit(expr_z_de, design_z)
fit_z <- eBayes(fit_z)

DE_z_TAPSE <- topTable(
  fit_z,
  coef = "TAPSE_sPAP",
  number = Inf,
  adjust.method = "BH"
)

DE_z_TAPSE$Gene <- rownames(DE_z_TAPSE)

DE_z_TAPSE_signif <- DE_z_TAPSE %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

write.csv(
  DE_z_TAPSE,
  file = here::here(output_dir_data, "DE_TAPSE_sPAP_SScPAH_Zurich.csv"),
  quote = FALSE, row.names = FALSE
)

write.csv(
  DE_z_TAPSE_signif,
  file = here::here(output_dir_data, "DE_TAPSE_sPAP_SScPAH_Zurich_significant.csv"),
  quote = FALSE, row.names = FALSE
)

n_z_TAPSE <- ncol(expr_z_de)
```

```{r de_tapse_poland_pah}
meta_p_de <- clin_p %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    Sex = factor(Sex),
    Age = as.numeric(Age),
    TAPSE_sPAP = as.numeric(TAPSE_sPAP)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_p_de <- meta_p_de[
  complete.cases(meta_p_de[, c("TAPSE_sPAP", "Age", "Sex")]),
  , drop = FALSE
]

expr_p_de <- as.matrix(data_p[, rownames(meta_p_de), drop = FALSE])

design_p <- model.matrix(~ TAPSE_sPAP + Age + Sex, data = meta_p_de)

design_p[, "TAPSE_sPAP"] <- rescale(design_p[, "TAPSE_sPAP"])
design_p[, "Age"]        <- rescale(design_p[, "Age"])

fit_p <- lmFit(expr_p_de, design_p)
fit_p <- eBayes(fit_p)

DE_p_TAPSE <- topTable(
  fit_p,
  coef = "TAPSE_sPAP",
  number = Inf,
  adjust.method = "BH"
)

DE_p_TAPSE$Gene <- rownames(DE_p_TAPSE)

DE_p_TAPSE_signif <- DE_p_TAPSE %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

write.csv(
  DE_p_TAPSE,
  file = here::here(output_dir_data, "DE_TAPSE_sPAP_SScPAH_Poland.csv"),
  quote = FALSE, row.names = FALSE
)

write.csv(
  DE_p_TAPSE_signif,
  file = here::here(output_dir_data, "DE_TAPSE_sPAP_SScPAH_Poland_significant.csv"),
  quote = FALSE, row.names = FALSE
)

n_p_TAPSE <- ncol(expr_p_de)
```

```{r, fig.height=4, fig.width=3, warning=FALSE}
keyvals <- rep("grey", nrow(DE_z_TAPSE))
names(keyvals) <- rep("Not significant", nrow(DE_z_TAPSE))

keyvals[DE_z_TAPSE$logFC > 0.5 & DE_z_TAPSE$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Increased"

keyvals[DE_z_TAPSE$logFC < -0.5 & DE_z_TAPSE$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

volcano_z_TAPSE <- EnhancedVolcano(
  DE_z_TAPSE,
  lab = rownames(DE_z_TAPSE),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Zurich cohort SSc-PAH\nTAPSE_sPAP\nn = ", n_z_TAPSE)
) +
  theme(legend.position = "none")

volcano_z_TAPSE
```

```{r, fig.height=4, fig.width=3, warning=FALSE}
keyvals <- rep("grey", nrow(DE_p_TAPSE))
names(keyvals) <- rep("Not significant", nrow(DE_p_TAPSE))

keyvals[DE_p_TAPSE$logFC > 0.5 & DE_p_TAPSE$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Increased"

keyvals[DE_p_TAPSE$logFC < -0.5 & DE_p_TAPSE$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

volcano_p_TAPSE <- EnhancedVolcano(
  DE_p_TAPSE,
  lab = rownames(DE_p_TAPSE),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Poland cohort SSc-PAH\nTAPSE_sPAP\nn = ", n_p_TAPSE)
) +
  theme(legend.position = "none")

volcano_p_TAPSE
```

```{r}
meta_z_bin <- clin_z %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = ifelse(TAPSE_sPAP < 0.32, "Low", "High"),
    TAPSE_group = factor(TAPSE_group, levels = c("High", "Low")),
    Sex = factor(Sex),
    Age = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_z_bin <- meta_z_bin[
  complete.cases(meta_z_bin[, c("TAPSE_group", "Age", "Sex")]),
  , drop = FALSE
]

expr_z_bin <- as.matrix(data_z[, rownames(meta_z_bin), drop = FALSE])

design_z_bin <- model.matrix(~ TAPSE_group + Age + Sex, data = meta_z_bin)
design_z_bin[, "Age"] <- rescale(design_z_bin[, "Age"])

fit_z_bin <- lmFit(expr_z_bin, design_z_bin)
fit_z_bin <- eBayes(fit_z_bin)

coef_bin_z <- grep("^TAPSE_group", colnames(design_z_bin), value = TRUE)[1]

DE_z_bin <- topTable(fit_z_bin, coef = coef_bin_z, number = Inf, adjust.method = "BH")
DE_z_bin$Gene <- rownames(DE_z_bin)

DE_z_bin_signif <- DE_z_bin %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

n_z_bin <- ncol(expr_z_bin)
```

```{r, fig.height=4, fig.width=3, warning=FALSE}
keyvals <- rep("grey", nrow(DE_z_bin))
names(keyvals) <- rep("Not significant", nrow(DE_z_bin))

keyvals[DE_z_bin$logFC > 0.5 & DE_z_bin$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Higher in Low TAPSE_sPAP"

keyvals[DE_z_bin$logFC < -0.5 & DE_z_bin$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Higher in High TAPSE_sPAP"

volcano_z_bin <- EnhancedVolcano(
  DE_z_bin,
  lab = rownames(DE_z_bin),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Zurich cohort SSc-PAH\nTAPSE/sPAP<0.32 vs ≥0.32\nn = ", n_z_bin)
) +
  theme(legend.position = "none")

volcano_z_bin
```

```{r}
meta_p_bin <- clin_p %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = ifelse(TAPSE_sPAP < 0.32, "Low", "High"),
    TAPSE_group = factor(TAPSE_group, levels = c("High", "Low")),
    Sex = factor(Sex),
    Age = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_p_bin <- meta_p_bin[
  complete.cases(meta_p_bin[, c("TAPSE_group", "Age", "Sex")]),
  , drop = FALSE
]

expr_p_bin <- as.matrix(data_p[, rownames(meta_p_bin), drop = FALSE])

design_p_bin <- model.matrix(~ TAPSE_group + Age + Sex, data = meta_p_bin)
design_p_bin[, "Age"] <- rescale(design_p_bin[, "Age"])

fit_p_bin <- lmFit(expr_p_bin, design_p_bin)
fit_p_bin <- eBayes(fit_p_bin)

coef_bin_p <- grep("^TAPSE_group", colnames(design_p_bin), value = TRUE)[1]

DE_p_bin <- topTable(fit_p_bin, coef = coef_bin_p, number = Inf, adjust.method = "BH")
DE_p_bin$Gene <- rownames(DE_p_bin)

DE_p_bin_signif <- DE_p_bin %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

n_p_bin <- ncol(expr_p_bin)
```

```{r, fig.height=4, fig.width=3, warning=FALSE}
keyvals <- rep("grey", nrow(DE_p_bin))
names(keyvals) <- rep("Not significant", nrow(DE_p_bin))

keyvals[DE_p_bin$logFC > 0.5 & DE_p_bin$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Higher in Low TAPSE_sPAP"

keyvals[DE_p_bin$logFC < -0.5 & DE_p_bin$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Higher in High TAPSE_sPAP"

volcano_p_bin <- EnhancedVolcano(
  DE_p_bin,
  lab = rownames(DE_p_bin),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Poland cohort SSc-PAH\nTAPSE/sPAP<0.32 vs ≥0.32\nn = ", n_p_bin)
) +
  theme(legend.position = "none")

volcano_p_bin
```

```{r}
meta_z_comp <- clin_z %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    COMPERA_number = as.numeric(COMPERA_number),
    Sex = factor(Sex),
    Age = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_z_comp <- meta_z_comp[
  complete.cases(meta_z_comp[, c("COMPERA_number", "Age", "Sex")]),
  , drop = FALSE
]

expr_z_comp <- as.matrix(data_z[, rownames(meta_z_comp), drop = FALSE])

design_z_comp <- model.matrix(~ COMPERA_number + Age + Sex, data = meta_z_comp)

design_z_comp[, "COMPERA_number"] <- rescale(design_z_comp[, "COMPERA_number"])
design_z_comp[, "Age"]            <- rescale(design_z_comp[, "Age"])

fit_z_comp <- lmFit(expr_z_comp, design_z_comp)
fit_z_comp <- eBayes(fit_z_comp)

DE_z_comp <- topTable(
  fit_z_comp,
  coef = "COMPERA_number",
  number = Inf,
  adjust.method = "BH"
)

DE_z_comp$Gene <- rownames(DE_z_comp)

DE_z_comp_signif <- DE_z_comp %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

n_z_comp <- ncol(expr_z_comp)
```

```{r, fig.height=4, fig.width=3, warning=FALSE}
keyvals <- rep("grey", nrow(DE_z_comp))
names(keyvals) <- rep("Not significant", nrow(DE_z_comp))

keyvals[DE_z_comp$logFC > 0.5 & DE_z_comp$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Positive association"

keyvals[DE_z_comp$logFC < -0.5 & DE_z_comp$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Negative association"

volcano_z_comp <- EnhancedVolcano(
  DE_z_comp,
  lab = rownames(DE_z_comp),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Zurich cohort SSc-PAH\nCOMPERA_number\nn = ", n_z_comp)
) +
  theme(legend.position = "none")

volcano_z_comp
```

```{r}
meta_p_comp <- clin_p %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    Compera_number = as.numeric(Compera_number),
    Sex = factor(Sex),
    Age = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_p_comp <- meta_p_comp[
  complete.cases(meta_p_comp[, c("Compera_number", "Age", "Sex")]),
  , drop = FALSE
]

expr_p_comp <- as.matrix(data_p[, rownames(meta_p_comp), drop = FALSE])

design_p_comp <- model.matrix(~ Compera_number + Age + Sex, data = meta_p_comp)

design_p_comp[, "Compera_number"] <- rescale(design_p_comp[, "Compera_number"])
design_p_comp[, "Age"]            <- rescale(design_p_comp[, "Age"])

fit_p_comp <- lmFit(expr_p_comp, design_p_comp)
fit_p_comp <- eBayes(fit_p_comp)

DE_p_comp <- topTable(
  fit_p_comp,
  coef = "Compera_number",
  number = Inf,
  adjust.method = "BH"
)

DE_p_comp$Gene <- rownames(DE_p_comp)

DE_p_comp_signif <- DE_p_comp %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

n_p_comp <- ncol(expr_p_comp)
```

```{r, fig.height=4, fig.width=3, warning=FALSE}
keyvals <- rep("grey", nrow(DE_p_comp))
names(keyvals) <- rep("Not significant", nrow(DE_p_comp))

keyvals[DE_p_comp$logFC > 0.5 & DE_p_comp$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Positive association"

keyvals[DE_p_comp$logFC < -0.5 & DE_p_comp$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Negative association"

volcano_p_comp <- EnhancedVolcano(
  DE_p_comp,
  lab = rownames(DE_p_comp),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Poland cohort\nSSc-PAH only\nCOMPERA_number\nn = ", n_p_comp)
) +
  theme(legend.position = "none")

volcano_p_comp
```

```{r cor_group, fig.height=4, fig.width=4}
df_group <- dplyr::inner_join(
  DE_z[, c("Gene", "logFC")],
  DE_p[, c("Gene", "logFC")],
  by = "Gene",
  suffix = c("_z", "_p")
)

cor_group <- cor(df_group$logFC_z, df_group$logFC_p)

ggplot(df_group, aes(logFC_z, logFC_p)) +
  geom_point(size = 1, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  labs(
    x = "Zurich logFC",
    y = "Poland logFC",
    title = paste0("SSc-PAH vs SSc-noPAH\nr = ", round(cor_group, 2))
  ) +
  theme_classic()
```

```{r cor_tapse_cont, fig.height=4, fig.width=4}
df_tapse <- dplyr::inner_join(
  DE_z_TAPSE[, c("Gene", "logFC")],
  DE_p_TAPSE[, c("Gene", "logFC")],
  by = "Gene",
  suffix = c("_z", "_p")
)

cor_tapse <- cor(df_tapse$logFC_z, df_tapse$logFC_p)

ggplot(df_tapse, aes(logFC_z, logFC_p)) +
  geom_point(size = 1, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  labs(
    x = "Zurich logFC",
    y = "Poland logFC",
    title = paste0("TAPSE_sPAP (continuous)\nr = ", round(cor_tapse, 2))
  ) +
  theme_classic()
```

```{r cor_tapse_binary, fig.height=4, fig.width=4}
df_tapse_bin <- dplyr::inner_join(
  DE_z_bin[, c("Gene", "logFC")],
  DE_p_bin[, c("Gene", "logFC")],
  by = "Gene",
  suffix = c("_z", "_p")
)

cor_tapse_bin <- cor(df_tapse_bin$logFC_z, df_tapse_bin$logFC_p)

ggplot(df_tapse_bin, aes(logFC_z, logFC_p)) +
  geom_point(size = 1, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  labs(
    x = "Zurich logFC",
    y = "Poland logFC",
    title = paste0("TAPSE groups (<0.32 vs ≥0.32)\nr = ", round(cor_tapse_bin, 2))
  ) +
  theme_classic()
```

```{r cor_compera, fig.height=4, fig.width=4}
df_comp <- dplyr::inner_join(
  DE_z_comp[, c("Gene", "logFC")],
  DE_p_comp[, c("Gene", "logFC")],
  by = "Gene",
  suffix = c("_z", "_p")
)

cor_comp <- cor(df_comp$logFC_z, df_comp$logFC_p)

ggplot(df_comp, aes(logFC_z, logFC_p)) +
  geom_point(size = 1, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  labs(
    x = "Zurich logFC",
    y = "Poland logFC",
    title = paste0("COMPERA_number\nr = ", round(cor_comp, 2))
  ) +
  theme_classic()
```

```{r score_heatmap_helpers, message=FALSE, warning=FALSE}
# score for ranking (for GSEA later)
add_score <- function(de) {
  de$score <- de$logFC * (-log10(de$P.Value))
  de
}

# heatmap: top20 up/down by score
plot_top_score_heatmap <- function(expr, clin, de, samples, ha, title_text) {
  de <- add_score(de)

  de_sub <- de[match(rownames(expr), de$Gene), ]
  rownames(de_sub) <- de_sub$Gene

  up   <- de_sub[order(de_sub$score, decreasing = TRUE), , drop = FALSE]
  down <- de_sub[order(de_sub$score, decreasing = FALSE), , drop = FALSE]

  top_up   <- rownames(up)[1:20]
  top_down <- rownames(down)[1:20]

  genes_use <- c(top_up, top_down)

  mat <- as.matrix(expr[genes_use, samples, drop = FALSE])

  # z-score per protein (row), same as your heatmaps
  matz <- t(scale(t(mat)))
  matz[is.nan(matz)] <- 0

  hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

  Heatmap(
    matz,
    name = "Z-score",
    col = hm_cols,
    top_annotation = ha,
    cluster_rows = FALSE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    row_names_gp = gpar(fontsize = 13),
    show_column_names = TRUE,
    column_names_gp = gpar(fontsize = 8),
    row_title = title_text,
    column_title = NULL,
    height = unit(nrow(matz) * 5, "mm")
  ) |>
    draw(heatmap_legend_side = "right", annotation_legend_side = "right")
}
```

```{r add_tapse_group}
clin_z <- clin_z %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = dplyr::case_when(
      is.na(TAPSE_sPAP)      ~ NA_character_,
      TAPSE_sPAP < 0.32      ~ "<0.32",
      TAPSE_sPAP >= 0.32     ~ "≥0.32"
    ),
    TAPSE_group = factor(TAPSE_group, levels = c("≥0.32", "<0.32"))
  )

clin_p <- clin_p %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = dplyr::case_when(
      is.na(TAPSE_sPAP)      ~ NA_character_,
      TAPSE_sPAP < 0.32      ~ "<0.32",
      TAPSE_sPAP >= 0.32     ~ "≥0.32"
    ),
    TAPSE_group = factor(TAPSE_group, levels = c("≥0.32", "<0.32"))
  )
```

```{r heatmaps_scores_zurich_tapseGroup, fig.height=10, fig.width=10.8}
# ---- add TAPSE/sPAP group to clin_z once ----
clin_z <- clin_z %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    `TAPSE/sPAP` = dplyr::case_when(
      is.na(TAPSE_sPAP)  ~ NA_character_,
      TAPSE_sPAP < 0.32  ~ "<0.32",
      TRUE               ~ "≥0.32"
    ),
    `TAPSE/sPAP` = factor(`TAPSE/sPAP`, levels = c("≥0.32", "<0.32"))
  )

# ---- colors (same palette logic) ----
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "low"               = "#9CA3AF",
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)
cols_tapse_spap <- c("≥0.32" = "#9CA3AF", "<0.32" = "#A63A2A")

# 1) SSc-PAH vs SSc-noPAH
samples <- intersect(colnames(data_z), clin_z$Barcode)
meta_hm <- clin_z[match(samples, clin_z$Barcode), ]

ha <- HeatmapAnnotation(
  Group        = meta_hm$Group,
  COMPERA      = factor(meta_hm$COMPERA_4strata, levels = c("low","intermediate-low","intermediate-high","high")),
  `TAPSE/sPAP` = meta_hm$`TAPSE/sPAP`,
  col = list(Group = cols_group, COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_z,
  clin = clin_z,
  de = DE_z,
  samples = samples,
  ha = ha,
  title_text = "Zurich: top20 up/down by score\nSSc-PAH vs SSc-noPAH"
)

# 2) TAPSE_sPAP (continuous), SSc-PAH only
meta2 <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

samples2 <- meta2$Barcode

ha2 <- HeatmapAnnotation(
  COMPERA      = factor(meta2$COMPERA_4strata, levels = c("low","intermediate-low","intermediate-high","high")),
  `TAPSE/sPAP` = meta2$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_z,
  clin = clin_z,
  de = DE_z_TAPSE,
  samples = samples2,
  ha = ha2,
  title_text = "Zurich: top20 up/down by score\nTAPSE_sPAP (continuous), SSc-PAH"
)

# 3) TAPSE_sPAP groups (<0.32 vs ≥0.32), SSc-PAH only
meta3 <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

samples3 <- meta3$Barcode

ha3 <- HeatmapAnnotation(
  COMPERA      = factor(meta3$COMPERA_4strata, levels = c("low","intermediate-low","intermediate-high","high")),
  `TAPSE/sPAP` = meta3$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_z,
  clin = clin_z,
  de = DE_z_bin,
  samples = samples3,
  ha = ha3,
  title_text = "Zurich: top20 up/down by score\nTAPSE_sPAP <0.32 vs ≥0.32, SSc-PAH"
)

# 4) COMPERA_number (continuous), SSc-PAH only
meta4 <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(COMPERA_number), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

samples4 <- meta4$Barcode

ha4 <- HeatmapAnnotation(
  COMPERA      = factor(meta4$COMPERA_4strata, levels = c("low","intermediate-low","intermediate-high","high")),
  `TAPSE/sPAP` = meta4$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_z,
  clin = clin_z,
  de = DE_z_comp,
  samples = samples4,
  ha = ha4,
  title_text = "Zurich: top20 up/down by score\nCOMPERA_number (continuous), SSc-PAH"
)
```

```{r heatmaps_scores_poland_tapseGroup, fig.height=10, fig.width=10.8}
# ---- add TAPSE/sPAP group to clin_p once ----
clin_p <- clin_p %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    `TAPSE/sPAP` = dplyr::case_when(
      is.na(TAPSE_sPAP)  ~ NA_character_,
      TAPSE_sPAP < 0.32  ~ "<0.32",
      TRUE               ~ "≥0.32"
    ),
    `TAPSE/sPAP` = factor(`TAPSE/sPAP`, levels = c("≥0.32", "<0.32"))
  )

# ---- colors ----
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "Low"               = "#9CA3AF",
  "Intermediate-low"  = "#DDB7B2",
  "Intermediate-high" = "#C7685F",
  "High"              = "#A63A2A"
)
cols_origin <- c("Krakow"="orange", "Lublin"="purple")
cols_tapse_spap <- c("≥0.32" = "#9CA3AF", "<0.32" = "#A63A2A")

# 1) SSc-PAH vs SSc-noPAH
samples <- intersect(colnames(data_p), clin_p$Barcode)
meta_hm <- clin_p[match(samples, clin_p$Barcode), ]

ha <- HeatmapAnnotation(
  Group        = meta_hm$Group,
  COMPERA      = factor(meta_hm$Compera_4strata, levels = c("Low","Intermediate-low","Intermediate-high","High")),
  Origin       = meta_hm$Origin,
  `TAPSE/sPAP` = meta_hm$`TAPSE/sPAP`,
  col = list(Group = cols_group, COMPERA = cols_compera, Origin = cols_origin, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_p,
  clin = clin_p,
  de = DE_p,
  samples = samples,
  ha = ha,
  title_text = "Poland: top20 up/down by score\nSSc-PAH vs SSc-noPAH"
)

# 2) TAPSE_sPAP (continuous), SSc-PAH only
meta2 <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

samples2 <- meta2$Barcode

ha2 <- HeatmapAnnotation(
  COMPERA      = factor(meta2$Compera_4strata, levels = c("Low","Intermediate-low","Intermediate-high","High")),
  Origin       = meta2$Origin,
  `TAPSE/sPAP` = meta2$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, Origin = cols_origin, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_p,
  clin = clin_p,
  de = DE_p_TAPSE,
  samples = samples2,
  ha = ha2,
  title_text = "Poland: top20 up/down by score\nTAPSE_sPAP (continuous), SSc-PAH"
)

# 3) TAPSE_sPAP groups (<0.32 vs ≥0.32), SSc-PAH only
meta3 <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

samples3 <- meta3$Barcode

ha3 <- HeatmapAnnotation(
  COMPERA      = factor(meta3$Compera_4strata, levels = c("Low","Intermediate-low","Intermediate-high","High")),
  Origin       = meta3$Origin,
  `TAPSE/sPAP` = meta3$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, Origin = cols_origin, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_p,
  clin = clin_p,
  de = DE_p_bin,
  samples = samples3,
  ha = ha3,
  title_text = "Poland: top20 up/down by score\nTAPSE_sPAP <0.32 vs ≥0.32, SSc-PAH"
)

# 4) Compera_number (continuous), SSc-PAH only
meta4 <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(Compera_number), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

samples4 <- meta4$Barcode

ha4 <- HeatmapAnnotation(
  COMPERA      = factor(meta4$Compera_4strata, levels = c("Low","Intermediate-low","Intermediate-high","High")),
  Origin       = meta4$Origin,
  `TAPSE/sPAP` = meta4$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, Origin = cols_origin, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_p,
  clin = clin_p,
  de = DE_p_comp,
  samples = samples4,
  ha = ha4,
  title_text = "Poland: top20 up/down by score\nCOMPERA_number (continuous), SSc-PAH"
)
```

## Check Age and Sex

```{r}
# Wilcoxon test (robust, no normality assumption)
age_test_p <- wilcox.test(Age ~ `TAPSE/sPAP`, data = clin_p)

age_test_p

ggplot(clin_p, aes(x = `TAPSE/sPAP`, y = Age, fill = `TAPSE/sPAP`)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.7) +
  theme_classic() +
  labs(title = "Poland cohort: Age by TAPSE/sPAP group",
       y = "Age (years)",
       x = "TAPSE/sPAP group") +
  theme(legend.position = "none")
```

```{r}
# Contingency table
tab_sex_p <- table(clin_p$`TAPSE/sPAP`, clin_p$Sex)
tab_sex_p

# If expected counts are small use Fisher
sex_test_p <- fisher.test(tab_sex_p)
sex_test_p

ggplot(clin_p, aes(x = `TAPSE/sPAP`, fill = Sex)) +
  geom_bar(position = "fill") +
  theme_classic() +
  labs(title = "Poland cohort: Sex distribution by TAPSE/sPAP",
       y = "Proportion",
       x = "TAPSE/sPAP group")
```

```{r}
age_test_z <- wilcox.test(Age ~ `TAPSE/sPAP`, data = clin_z)

age_test_z

ggplot(clin_z, aes(x = `TAPSE/sPAP`, y = Age, fill = `TAPSE/sPAP`)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.7) +
  theme_classic() +
  labs(title = "Zurich cohort: Age by TAPSE/sPAP group",
       y = "Age (years)",
       x = "TAPSE/sPAP group") +
  theme(legend.position = "none")
```

```{r}
tab_sex_z <- table(clin_z$`TAPSE/sPAP`, clin_z$Sex)
tab_sex_z

sex_test_z <- fisher.test(tab_sex_z)
sex_test_z
ggplot(clin_z, aes(x = `TAPSE/sPAP`, fill = Sex)) +
  geom_bar(position = "fill") +
  theme_classic() +
  labs(title = "Zurich cohort: Sex distribution by TAPSE/sPAP",
       y = "Proportion",
       x = "TAPSE/sPAP group")
```

# Lasso

```{r panel_lasso_functions, message=FALSE, warning=FALSE}
library(glmnet)
library(pROC)
library(ggplot2)
library(dplyr)

fit_panel_lasso <- function(
  expr_train, clin_train,
  expr_test,  clin_test,
  cutoff = 0.32,
  use_covariates = FALSE,
  alpha = 1
){
  # --- outcome (SSc-PAH only assumed upstream) ---
  y_train <- ifelse(as.numeric(clin_train$TAPSE_sPAP) < cutoff, 1, 0)
  y_test  <- ifelse(as.numeric(clin_test$TAPSE_sPAP)  < cutoff, 1, 0)

  # --- X matrices: samples x proteins ---
  Xtr_prot <- t(as.matrix(expr_train[, clin_train$Barcode, drop = FALSE]))
  Xte_prot <- t(as.matrix(expr_test[,  clin_test$Barcode,  drop = FALSE]))

  if (!use_covariates) {
    Xtr <- Xtr_prot
    Xte <- Xte_prot
    pf  <- rep(1, ncol(Xtr))
  } else {
    cov_tr <- data.frame(
      Age = as.numeric(clin_train$Age),
      Sex = factor(clin_train$Sex)
    )
    cov_te <- data.frame(
      Age = as.numeric(clin_test$Age),
      Sex = factor(clin_test$Sex, levels = levels(cov_tr$Sex))
    )

    Xtr_cov <- model.matrix(~ Age + Sex, data = cov_tr)[, -1, drop = FALSE]
    Xte_cov <- model.matrix(~ Age + Sex, data = cov_te)[, -1, drop = FALSE]

    Xtr <- cbind(Xtr_cov, Xtr_prot)
    Xte <- cbind(Xte_cov, Xte_prot)

    # Age/Sex unpenalized, proteins penalized
    pf <- c(rep(0, ncol(Xtr_cov)), rep(1, ncol(Xtr_prot)))
  }

  # --- CV LASSO ---
  set.seed(1)
  cvfit <- cv.glmnet(
    x = Xtr, y = y_train,
    family = "binomial",
    alpha = alpha,
    penalty.factor = pf,
    standardize = TRUE,
    nfolds = 3,
    type.measure = "auc"
  )

  # --- choose lambda giving 3–10 proteins (closest if none) ---
  lambdas <- cvfit$glmnet.fit$lambda
  beta_mat <- as.matrix(coef(cvfit$glmnet.fit, s = lambdas))  # rows = features+intercept, cols = lambdas

  # count proteins only
  feat_names <- rownames(beta_mat)
  prot_idx <- if (!use_covariates) {
    which(feat_names != "(Intercept)")
  } else {
    which(feat_names != "(Intercept)" & !(feat_names %in% colnames(Xtr)[1:(ncol(Xtr) - ncol(Xtr_prot))]))
    # easier: proteins are the columns of Xtr_prot
    which(feat_names %in% colnames(Xtr_prot))
  }

  nprot <- apply(beta_mat[prot_idx, , drop = FALSE] != 0, 2, sum)

  ok <- which(nprot >= 3 & nprot <= 10)
  if (length(ok) > 0) {
    # pick best CV AUC among those (closest to min cvm)
    j <- ok[which.min(cvfit$cvm[match(lambdas[ok], cvfit$lambda)])]
    lambda_use <- lambdas[j]
  } else {
    # pick lambda with nprot closest to 5
    j <- which.min(abs(nprot - 5))
    lambda_use <- lambdas[j]
  }

  # --- final coefficients at chosen lambda ---
  beta <- as.matrix(coef(cvfit, s = lambda_use))
  beta <- beta[beta[,1] != 0, , drop = FALSE]

  # proteins selected (exclude intercept + covariates)
  selected <- rownames(beta)
  selected <- selected[selected != "(Intercept)"]
  if (use_covariates) {
    selected_proteins <- selected[selected %in% colnames(Xtr_prot)]
  } else {
    selected_proteins <- selected
  }

  # --- predict in Poland using locked Zurich model ---
  p_train <- as.numeric(predict(cvfit, newx = Xtr, s = lambda_use, type = "response"))
  p_test  <- as.numeric(predict(cvfit, newx = Xte, s = lambda_use, type = "response"))

  # AUCs
  auc_train <- as.numeric(pROC::auc(pROC::roc(y_train, p_train, quiet = TRUE)))
  auc_test  <- as.numeric(pROC::auc(pROC::roc(y_test,  p_test,  quiet = TRUE)))

  list(
    cvfit = cvfit,
    lambda_use = lambda_use,
    selected_proteins = selected_proteins,
    beta_nonzero = beta,
    train = data.frame(Barcode = clin_train$Barcode, y = y_train, p = p_train),
    test  = data.frame(Barcode = clin_test$Barcode,  y = y_test,  p = p_test),
    auc_train = auc_train,
    auc_test = auc_test
  )
}

plot_panel_results <- function(res, title_prefix = ""){
  roc_train <- pROC::roc(res$train$y, res$train$p, quiet = TRUE)
  roc_test  <- pROC::roc(res$test$y,  res$test$p,  quiet = TRUE)

  p1 <- ggplot() +
    geom_line(data = data.frame(
      tpr = rev(roc_train$sensitivities),
      fpr = rev(1 - roc_train$specificities),
      cohort = "Zurich"
    ), aes(fpr, tpr)) +
    geom_line(data = data.frame(
      tpr = rev(roc_test$sensitivities),
      fpr = rev(1 - roc_test$specificities),
      cohort = "Poland"
    ), aes(fpr, tpr)) +
    geom_abline(linetype = 2) +
    labs(
      x = "False positive rate",
      y = "True positive rate",
      title = paste0(
        title_prefix,
        "\nAUC Zurich = ", round(res$auc_train, 2),
        " | AUC Poland = ", round(res$auc_test, 2)
      )
    ) +
    theme_classic()

  df_box <- bind_rows(
    transform(res$train, cohort = "Zurich"),
    transform(res$test,  cohort = "Poland")
  )
  df_box$Group <- factor(ifelse(df_box$y == 1, "<0.32", "≥0.32"), levels = c("≥0.32", "<0.32"))

  p2 <- ggplot(df_box, aes(Group, p)) +
    geom_boxplot(outlier.size = 0.7) +
    geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
    facet_wrap(~ cohort) +
    labs(x = "TAPSE/sPAP group", y = "Predicted probability (Low group)") +
    theme_classic()

  list(roc = p1, score_box = p2)
}
```

```{r panel_proteins_only, fig.height=4, fig.width=6.5}
# --- SSc-PAH only + complete cases (TAPSE_sPAP needed for outcome) ---
z_sub <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

p_sub <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

res_prot <- fit_panel_lasso(
  expr_train = data_z,
  clin_train = z_sub,
  expr_test  = data_p,
  clin_test  = p_sub,
  cutoff = 0.32,
  use_covariates = FALSE
)

res_prot$selected_proteins
plots <- plot_panel_results(res_prot, title_prefix = "Panel (proteins only)")
plots$roc
```

```{r panel_proteins_only_box, fig.height=3.8, fig.width=6.5}
plots$score_box
```

```{r panel_proteins_age_sex, fig.height=4, fig.width=6.5}
z_sub2 <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

p_sub2 <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

res_cov <- fit_panel_lasso(
  expr_train = data_z,
  clin_train = z_sub2,
  expr_test  = data_p,
  clin_test  = p_sub2,
  cutoff = 0.32,
  use_covariates = TRUE
)

res_cov$selected_proteins
plots2 <- plot_panel_results(res_cov, title_prefix = "Panel (proteins + Age + Sex)")
plots2$roc
```

```{r panel_proteins_age_sex_box, fig.height=3.8, fig.width=6.5}
plots2$score_box
```

```{r de_zurich_lowTAPSE_vs_noPAH}
# Zurich only: compare SSc-PAH with low TAPSE/sPAP (<0.32) vs SSc-noPAH
meta_z_low_vs_no <- clin_z %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    Group2 = dplyr::case_when(
      Group == "SSc-noPAH" ~ "SSc-noPAH",
      Group == "SSc-PAH" & !is.na(TAPSE_sPAP) & TAPSE_sPAP < 0.32 ~ "SSc-PAH_low",
      TRUE ~ NA_character_
    ),
    Group2 = factor(Group2, levels = c("SSc-noPAH", "SSc-PAH_low")),
    Sex = factor(Sex),
    Age = as.numeric(Age)
  ) %>%
  dplyr::filter(!is.na(Group2)) %>%
  tibble::column_to_rownames("Barcode")

# keep complete cases for covariates you want to adjust for
meta_z_low_vs_no <- meta_z_low_vs_no[
  complete.cases(meta_z_low_vs_no[, c("Group2", "Age", "Sex")]),
  , drop = FALSE
]

expr_z_low_vs_no <- as.matrix(data_z[, rownames(meta_z_low_vs_no), drop = FALSE])

design <- model.matrix(~ Group2 + Age + Sex, data = meta_z_low_vs_no)
design[, "Age"] <- rescale(design[, "Age"])

fit <- lmFit(expr_z_low_vs_no, design)
fit <- eBayes(fit)

coef_name <- grep("^Group2", colnames(design), value = TRUE)[1]

DE_z_low_vs_no <- topTable(fit, coef = coef_name, number = Inf, adjust.method = "BH")
DE_z_low_vs_no$Gene <- rownames(DE_z_low_vs_no)

DE_z_low_vs_no_signif <- DE_z_low_vs_no %>%
  dplyr::filter(adj.P.Val < 0.2, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

write.csv(DE_z_low_vs_no,
          file = here::here(output_dir_data, "DE_Zurich_SScPAH_lowTAPSE_vs_SScnoPAH.csv"),
          quote = FALSE, row.names = FALSE)

write.csv(DE_z_low_vs_no_signif,
          file = here::here(output_dir_data, "DE_Zurich_SScPAH_lowTAPSE_vs_SScnoPAH_significant.csv"),
          quote = FALSE, row.names = FALSE)

n_z_low_vs_no <- ncol(expr_z_low_vs_no)
```

```{r}

# ---- 1. Use Zurich TAPSE binary DE results ----
de_z <- DE_z_bin
de_z$score <- de_z$logFC * (-log10(de_z$P.Value))

# order by absolute score
de_z_ord <- de_z[order(abs(de_z$score), decreasing = TRUE), ]

top5 <- de_z_ord$Gene[1:50]
test <- DE_z_bin_signif %>% 
    dplyr::filter(Gene %in% dplyr::filter(DE_p_bin, P.Value < 0.05 & abs(logFC) > 0.5)$Gene) %>%
  dplyr::filter(Gene %in% DE_z_low_vs_no_signif$Gene)
test2 <- DE_p_bin %>% 
    dplyr::filter(Gene %in% test$Gene)
top5 <- test$Gene[1:3]
```

```{r build_additive_score}
# ---- helper to compute signed score ----
compute_panel_score <- function(expr, clin, genes, cutoff = 0.32) {

  # outcome
  y <- ifelse(as.numeric(clin$TAPSE_sPAP) < cutoff, 1, 0)

  # matrix samples x genes
  X <- t(as.matrix(expr[genes, clin$Barcode, drop = FALSE]))

  # z-score per protein within cohort
  Xz <- scale(X)

  # direction from Zurich logFC
  directions <- sign(de_z$logFC[match(genes, de_z$Gene)])

  # additive signed score
  score <- as.numeric(Xz %*% directions)

  data.frame(
    Barcode = clin$Barcode,
    y = y,
    score = score
  )
}
```

```{r evaluate_panel, fig.height=4, fig.width=6.5}
# ---- Zurich SSc-PAH ----
z_sub <- clin_z %>%
  filter(Group == "SSc-PAH") %>%
  filter(!is.na(TAPSE_sPAP)) %>%
  filter(Barcode %in% colnames(data_z))

res_z <- compute_panel_score(data_z, z_sub, top5)

# ---- Poland SSc-PAH ----
p_sub <- clin_p %>%
  filter(Group == "SSc-PAH") %>%
  filter(!is.na(TAPSE_sPAP)) %>%
  filter(Barcode %in% colnames(data_p))

res_p <- compute_panel_score(data_p, p_sub, top5)

# ---- AUCs ----
auc_z <- as.numeric(auc(roc(res_z$y, res_z$score, quiet=TRUE)))
auc_p <- as.numeric(auc(roc(res_p$y, res_p$score, quiet=TRUE)))

# ---- ROC plot ----
roc_z <- roc(res_z$y, res_z$score, quiet=TRUE)
roc_p <- roc(res_p$y, res_p$score, quiet=TRUE)

ggplot() +
  geom_line(data=data.frame(
    tpr=rev(roc_z$sensitivities),
    fpr=rev(1-roc_z$specificities)
  ), aes(fpr,tpr), color="black") +
  geom_line(data=data.frame(
    tpr=rev(roc_p$sensitivities),
    fpr=rev(1-roc_p$specificities)
  ), aes(fpr,tpr), color="red") +
  geom_abline(linetype=2) +
  labs(
    x="False positive rate",
    y="True positive rate",
    title=paste0(
      "5-protein panel (by Zurich score)\n",
      "AUC Zurich=", round(auc_z,2),
      " | AUC Poland=", round(auc_p,2)
    )
  ) +
  theme_classic()
```

```{r boxplot_panel, fig.height=4, fig.width=6.5}
df_box <- bind_rows(
  transform(res_z, cohort="Zurich"),
  transform(res_p, cohort="Poland")
)

df_box$Group <- factor(ifelse(df_box$y==1,"<0.32","≥0.32"),
                       levels=c("≥0.32","<0.32"))

ggplot(df_box, aes(Group, score)) +
  geom_boxplot() +
  geom_jitter(width=0.15, alpha=0.6) +
  facet_wrap(~cohort) +
  theme_classic() +
  labs(x="TAPSE/sPAP group", y="Panel score")
```

```{r panel_poland_train_zurich_test, message=FALSE, warning=FALSE}
library(glmnet)
library(pROC)
library(ggplot2)
library(dplyr)

# -------------------------
# 1) Subset to SSc-PAH only
# -------------------------
z_sub <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

p_sub <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

# -------------------------
# 2) Outcome: TAPSE/sPAP < 0.32
# -------------------------
cutoff <- 0.32
y_tr <- ifelse(as.numeric(p_sub$TAPSE_sPAP) < cutoff, 1, 0)  # train = Poland
y_te <- ifelse(as.numeric(z_sub$TAPSE_sPAP) < cutoff, 1, 0)  # test  = Zurich

# -------------------------
# 3) X matrices (samples x proteins)
# -------------------------
Xtr_prot <- t(as.matrix(data_p[, p_sub$Barcode, drop = FALSE]))
Xte_prot <- t(as.matrix(data_z[, z_sub$Barcode, drop = FALSE]))

# -------------------------
# 4) Option A: proteins only
# -------------------------
set.seed(1)
cvfit_prot <- cv.glmnet(
  x = Xtr_prot, y = y_tr,
  family = "binomial",
  alpha = 1,
  standardize = TRUE,
  nfolds = 5,
  type.measure = "auc"
)

# choose lambda that gives 3–10 proteins (otherwise closest to 5)
lambdas <- cvfit_prot$glmnet.fit$lambda
beta_mat <- as.matrix(coef(cvfit_prot$glmnet.fit, s = lambdas))
feat_names <- rownames(beta_mat)
prot_idx <- which(feat_names != "(Intercept)")
nprot <- apply(beta_mat[prot_idx, , drop = FALSE] != 0, 2, sum)

ok <- which(nprot >= 3 & nprot <= 20)
if (length(ok) > 0) {
  j <- ok[which.min(cvfit_prot$cvm[match(lambdas[ok], cvfit_prot$lambda)])]
  lambda_use_prot <- lambdas[j]
} else {
  j <- which.min(abs(nprot - 5))
  lambda_use_prot <- lambdas[j]
}

# selected proteins
beta_use <- as.matrix(coef(cvfit_prot, s = lambda_use_prot))
sel <- rownames(beta_use)[beta_use[,1] != 0]
sel_proteins <- sel[sel != "(Intercept)"]
sel_proteins

# predictions
p_tr <- as.numeric(predict(cvfit_prot, newx = Xtr_prot, s = lambda_use_prot, type = "response"))
p_te <- as.numeric(predict(cvfit_prot, newx = Xte_prot, s = lambda_use_prot, type = "response"))

auc_tr <- as.numeric(auc(roc(y_tr, p_tr, quiet=TRUE)))
auc_te <- as.numeric(auc(roc(y_te, p_te, quiet=TRUE)))

# ROC plot
roc_tr <- roc(y_tr, p_tr, quiet=TRUE)
roc_te <- roc(y_te, p_te, quiet=TRUE)

p_roc_prot <- ggplot() +
  geom_line(data=data.frame(tpr=rev(roc_tr$sensitivities), fpr=rev(1-roc_tr$specificities)),
            aes(fpr,tpr), color="red") +
  geom_line(data=data.frame(tpr=rev(roc_te$sensitivities), fpr=rev(1-roc_te$specificities)),
            aes(fpr,tpr), color="black") +
  geom_abline(linetype=2) +
  labs(
    x="False positive rate", y="True positive rate",
    title=paste0("LASSO panel (proteins only)\nTrain Poland AUC=", round(auc_tr,2),
                 " | Test Zurich AUC=", round(auc_te,2))
  ) +
  theme_classic()

p_roc_prot
```

```{r panel_poland_train_zurich_test_boxplot, fig.height=4, fig.width=6.5}
df_box <- bind_rows(
  data.frame(score = p_tr, y = y_tr, cohort="Poland"),
  data.frame(score = p_te, y = y_te, cohort="Zurich")
)
df_box$Group <- factor(ifelse(df_box$y==1,"<0.32","≥0.32"), levels=c("≥0.32","<0.32"))

ggplot(df_box, aes(Group, score)) +
  geom_boxplot() +
  geom_jitter(width=0.15, alpha=0.6) +
  facet_wrap(~cohort) +
  theme_classic() +
  labs(x="TAPSE/sPAP group", y="Predicted probability (Low group)")
```

```{r panel_poland_train_zurich_test_covariates, message=FALSE, warning=FALSE}
# -------------------------
# 5) Option B: proteins + Age + Sex (unpenalized)
# -------------------------
z_sub2 <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

p_sub2 <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

y_tr2 <- ifelse(as.numeric(p_sub2$TAPSE_sPAP) < cutoff, 1, 0)
y_te2 <- ifelse(as.numeric(z_sub2$TAPSE_sPAP) < cutoff, 1, 0)

Xtr_prot2 <- t(as.matrix(data_p[, p_sub2$Barcode, drop = FALSE]))
Xte_prot2 <- t(as.matrix(data_z[, z_sub2$Barcode, drop = FALSE]))

cov_tr <- data.frame(Age = as.numeric(p_sub2$Age), Sex = factor(p_sub2$Sex))
cov_te <- data.frame(Age = as.numeric(z_sub2$Age), Sex = factor(z_sub2$Sex, levels = levels(cov_tr$Sex)))

Xtr_cov <- model.matrix(~ Age + Sex, data = cov_tr)[, -1, drop = FALSE]
Xte_cov <- model.matrix(~ Age + Sex, data = cov_te)[, -1, drop = FALSE]

Xtr <- cbind(Xtr_cov, Xtr_prot2)
Xte <- cbind(Xte_cov, Xte_prot2)

pf <- c(rep(0, ncol(Xtr_cov)), rep(1, ncol(Xtr_prot2)))

set.seed(1)
cvfit_cov <- cv.glmnet(
  x = Xtr, y = y_tr2,
  family = "binomial",
  alpha = 1,
  penalty.factor = pf,
  standardize = TRUE,
  nfolds = 5,
  type.measure = "auc"
)

# choose lambda giving 3–10 proteins (otherwise closest to 5)
lambdas <- cvfit_cov$glmnet.fit$lambda
beta_mat <- as.matrix(coef(cvfit_cov$glmnet.fit, s = lambdas))
feat_names <- rownames(beta_mat)

prot_names <- colnames(Xtr_prot2)
prot_idx <- which(feat_names %in% prot_names)

nprot <- apply(beta_mat[prot_idx, , drop = FALSE] != 0, 2, sum)

ok <- which(nprot >= 3 & nprot <= 20)
if (length(ok) > 0) {
  j <- ok[which.min(cvfit_cov$cvm[match(lambdas[ok], cvfit_cov$lambda)])]
  lambda_use_cov <- lambdas[j]
} else {
  j <- which.min(abs(nprot - 5))
  lambda_use_cov <- lambdas[j]
}

beta_use <- as.matrix(coef(cvfit_cov, s = lambda_use_cov))
sel <- rownames(beta_use)[beta_use[,1] != 0]
sel_proteins_cov <- sel[sel %in% prot_names]
sel_proteins_cov

p_tr2 <- as.numeric(predict(cvfit_cov, newx = Xtr, s = lambda_use_cov, type = "response"))
p_te2 <- as.numeric(predict(cvfit_cov, newx = Xte, s = lambda_use_cov, type = "response"))

auc_tr2 <- as.numeric(auc(roc(y_tr2, p_tr2, quiet=TRUE)))
auc_te2 <- as.numeric(auc(roc(y_te2, p_te2, quiet=TRUE)))

roc_tr2 <- roc(y_tr2, p_tr2, quiet=TRUE)
roc_te2 <- roc(y_te2, p_te2, quiet=TRUE)

p_roc_cov <- ggplot() +
  geom_line(data=data.frame(tpr=rev(roc_tr2$sensitivities), fpr=rev(1-roc_tr2$specificities)),
            aes(fpr,tpr), color="red") +
  geom_line(data=data.frame(tpr=rev(roc_te2$sensitivities), fpr=rev(1-roc_te2$specificities)),
            aes(fpr,tpr), color="black") +
  geom_abline(linetype=2) +
  labs(
    x="False positive rate", y="True positive rate",
    title=paste0("LASSO panel (proteins + Age + Sex)\nTrain Poland AUC=", round(auc_tr2,2),
                 " | Test Zurich AUC=", round(auc_te2,2))
  ) +
  theme_classic()

p_roc_cov
```

```{r panel_poland_train_zurich_test_covariates_boxplot, fig.height=4, fig.width=6.5}
df_box2 <- bind_rows(
  data.frame(score = p_tr2, y = y_tr2, cohort="Poland"),
  data.frame(score = p_te2, y = y_te2, cohort="Zurich")
)

df_box2$Group <- factor(ifelse(df_box2$y==1,"<0.32","≥0.32"), levels=c("≥0.32","<0.32"))

ggplot(df_box2, aes(Group, score)) +
  geom_boxplot() +
  geom_jitter(width=0.15, alpha=0.6) +
  facet_wrap(~cohort) +
  stat_compare_means(
    comparisons = list(c("≥0.32","<0.32")),
    method = "wilcox.test",
    label = "p.format"
  ) +
  theme_classic() +
  labs(x="TAPSE/sPAP group", y="Predicted probability (Low group)")
```

```{r panel_poland_train_zurich_preselect50, message=FALSE, warning=FALSE}
library(glmnet)
library(pROC)
library(dplyr)
library(ggplot2)

# --------------------------------------------
# 1) Preselect top 50 proteins (Poland only)
# --------------------------------------------
de_train <- DE_p_bin   # TAPSE group DE in Poland
de_train$score <- de_train$logFC * (-log10(de_train$P.Value))
DE_p_signif_2 <- DE_p %>%
  dplyr::filter(adj.P.Val<0.1)
de_train_ord <- de_train[order(abs(de_train$score), decreasing = TRUE), ]
top50 <- de_train_ord %>%
  dplyr::filter(Gene %in% DE_p_signif_2$Gene) %>%
  slice_head(n=50) %>%
  pull(Gene)


length(top50)

# --------------------------------------------
# 2) Subset to SSc-PAH only
# --------------------------------------------
cutoff <- 0.32

p_sub <- clin_p %>%
  filter(Group == "SSc-PAH") %>%
  filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  filter(Barcode %in% colnames(data_p))

z_sub <- clin_z %>%
  filter(Group == "SSc-PAH") %>%
  filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  filter(Barcode %in% colnames(data_z))

y_tr <- ifelse(as.numeric(p_sub$TAPSE_sPAP) < cutoff, 1, 0)
y_te <- ifelse(as.numeric(z_sub$TAPSE_sPAP) < cutoff, 1, 0)

# --------------------------------------------
# 3) Build X matrices (only top50 proteins)
# --------------------------------------------
Xtr_prot <- t(as.matrix(data_p[top50, p_sub$Barcode, drop = FALSE]))
Xte_prot <- t(as.matrix(data_z[top50, z_sub$Barcode, drop = FALSE]))

# Age + Sex covariates (unpenalized)
cov_tr <- data.frame(
  Age = as.numeric(p_sub$Age),
  Sex = factor(p_sub$Sex)
)

cov_te <- data.frame(
  Age = as.numeric(z_sub$Age),
  Sex = factor(z_sub$Sex, levels = levels(cov_tr$Sex))
)

Xtr_cov <- model.matrix(~ Age + Sex, data = cov_tr)[, -1, drop = FALSE]
Xte_cov <- model.matrix(~ Age + Sex, data = cov_te)[, -1, drop = FALSE]

Xtr <- cbind(Xtr_cov, Xtr_prot)
Xte <- cbind(Xte_cov, Xte_prot)

# penalty factor: 0 for covariates, 1 for proteins
pf <- c(rep(0, ncol(Xtr_cov)), rep(1, ncol(Xtr_prot)))

# --------------------------------------------
# 4) LASSO (5-fold CV)
# --------------------------------------------
set.seed(1)

cvfit <- cv.glmnet(
  x = Xtr,
  y = y_tr,
  family = "binomial",
  alpha = 1,
  penalty.factor = pf,
  standardize = TRUE,
  nfolds = 5,
  type.measure = "auc"
)

# --------------------------------------------
# 5) Choose lambda giving 3–10 proteins
# --------------------------------------------
lambdas <- cvfit$glmnet.fit$lambda
beta_mat <- as.matrix(coef(cvfit$glmnet.fit, s = lambdas))
feat_names <- rownames(beta_mat)

prot_names <- colnames(Xtr_prot)
prot_idx <- which(feat_names %in% prot_names)

nprot <- apply(beta_mat[prot_idx, , drop = FALSE] != 0, 2, sum)

ok <- which(nprot >= 3 & nprot <= 10)

if (length(ok) > 0) {
  j <- ok[which.min(cvfit$cvm[match(lambdas[ok], cvfit$lambda)])]
  lambda_use <- lambdas[j]
} else {
  j <- which.min(abs(nprot - 5))
  lambda_use <- lambdas[j]
}

# selected proteins
beta_use <- as.matrix(coef(cvfit, s = lambda_use))
sel <- rownames(beta_use)[beta_use[,1] != 0]
sel_proteins <- sel[sel %in% prot_names]
sel_proteins

# --------------------------------------------
# 6) Predictions
# --------------------------------------------
p_tr <- as.numeric(predict(cvfit, newx = Xtr, s = lambda_use, type = "response"))
p_te <- as.numeric(predict(cvfit, newx = Xte, s = lambda_use, type = "response"))

auc_tr <- as.numeric(auc(roc(y_tr, p_tr, quiet=TRUE)))
auc_te <- as.numeric(auc(roc(y_te, p_te, quiet=TRUE)))

# --------------------------------------------
# 7) ROC plot
# --------------------------------------------
roc_tr <- roc(y_tr, p_tr, quiet=TRUE)
roc_te <- roc(y_te, p_te, quiet=TRUE)

ggplot() +
  geom_line(data=data.frame(tpr=rev(roc_tr$sensitivities),
                            fpr=rev(1-roc_tr$specificities)),
            aes(fpr,tpr), color="red") +
  geom_line(data=data.frame(tpr=rev(roc_te$sensitivities),
                            fpr=rev(1-roc_te$specificities)),
            aes(fpr,tpr), color="black") +
  geom_abline(linetype=2) +
  labs(
    x="False positive rate",
    y="True positive rate",
    title=paste0(
      "LASSO (top50 preselected, Age+Sex)\n",
      "Train Poland AUC=", round(auc_tr,2),
      " | Test Zurich AUC=", round(auc_te,2)
    )
  ) +
  theme_classic()
```


```{r panel_score_3groups_both_cohorts, fig.height=4.2, fig.width=8.2, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

# ---- build function to compute model matrix like in training ----
make_X_cov_prot <- function(clin_df, expr_mat, proteins, cov_levels_sex) {
  # covariates
  cov_df <- data.frame(
    Age = as.numeric(clin_df$Age),
    Sex = factor(clin_df$Sex, levels = cov_levels_sex)
  )
  X_cov <- model.matrix(~ Age + Sex, data = cov_df)[, -1, drop = FALSE]

  # proteins: samples x proteins
  X_prot <- t(as.matrix(expr_mat[proteins, clin_df$Barcode, drop = FALSE]))

  cbind(X_cov, X_prot)
}

# proteins used in the model input (top50 preselected)
proteins_use <- top50

# sex levels must match training (Poland) to get same design matrix columns
sex_levels_train <- levels(factor(p_sub2$Sex))  # from your Poland training subset (Age+Sex model)

# ---- make "all patients" clinical frames for each cohort ----
# Zurich: include SSc-noPAH and SSc-PAH, keep only those with Age/Sex available
z_all <- clin_z %>%
  dplyr::filter(Group %in% c("SSc-noPAH", "SSc-PAH")) %>%
  dplyr::filter(!is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    group3 = dplyr::case_when(
      Group == "SSc-noPAH" ~ "SSc-noPAH",
      Group == "SSc-PAH" & !is.na(TAPSE_sPAP) & TAPSE_sPAP < 0.32 ~ "SSc-PAH <0.32",
      Group == "SSc-PAH" & !is.na(TAPSE_sPAP) & TAPSE_sPAP >= 0.32 ~ "SSc-PAH ≥0.32",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(group3))

# Poland
p_all <- clin_p %>%
  dplyr::filter(Group %in% c("SSc-noPAH", "SSc-PAH")) %>%
  dplyr::filter(!is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    group3 = dplyr::case_when(
      Group == "SSc-noPAH" ~ "SSc-noPAH",
      Group == "SSc-PAH" & !is.na(TAPSE_sPAP) & TAPSE_sPAP < 0.32 ~ "SSc-PAH <0.32",
      Group == "SSc-PAH" & !is.na(TAPSE_sPAP) & TAPSE_sPAP >= 0.32 ~ "SSc-PAH ≥0.32",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(group3))

# order groups for plotting
group_levels <- c("SSc-noPAH", "SSc-PAH ≥0.32", "SSc-PAH <0.32")
z_all$group3 <- factor(z_all$group3, levels = group_levels)
p_all$group3 <- factor(p_all$group3, levels = group_levels)

# ---- compute scores using the locked Poland-trained cvfit at lambda_use ----
X_z_all <- make_X_cov_prot(z_all, data_z, proteins_use, cov_levels_sex = sex_levels_train)
X_p_all <- make_X_cov_prot(p_all, data_p, proteins_use, cov_levels_sex = sex_levels_train)

# probability and linear predictor (eta)
z_prob <- as.numeric(predict(cvfit, newx = X_z_all, s = lambda_use, type = "response"))
p_prob <- as.numeric(predict(cvfit, newx = X_p_all, s = lambda_use, type = "response"))

z_eta  <- as.numeric(predict(cvfit, newx = X_z_all, s = lambda_use, type = "link"))
p_eta  <- as.numeric(predict(cvfit, newx = X_p_all, s = lambda_use, type = "link"))

df_score <- bind_rows(
  data.frame(cohort = "Zurich", Barcode = z_all$Barcode, group3 = z_all$group3,
             prob_low = z_prob, eta = z_eta),
  data.frame(cohort = "Poland", Barcode = p_all$Barcode, group3 = p_all$group3,
             prob_low = p_prob, eta = p_eta)
)


comparisons_list <- list(
  c("SSc-noPAH", "SSc-PAH ≥0.32"),
  c("SSc-noPAH", "SSc-PAH <0.32"),
  c("SSc-PAH ≥0.32", "SSc-PAH <0.32")
)

ggplot(df_score, aes(group3, prob_low)) +
  geom_boxplot(outlier.size = 0.7) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ cohort) +
  stat_compare_means(
    comparisons = comparisons_list,
    method = "wilcox.test",
    label = "p.format",
    step.increase = 0.08
  ) +
  theme_classic() +
  labs(x = NULL,
       y = "LASSO score (predicted P[Low TAPSE/sPAP])")
```

```{r cox_zurich_compera_plus_lasso, message=FALSE, warning=FALSE}
library(survival)
library(broom)
library(dplyr)

# ---- 1) Zurich SSc-PAH: assemble survival dataframe ----
# df_score must exist and contain: cohort, Barcode, eta (or prob_low)
score_z <- df_score %>%
  dplyr::filter(cohort == "Zurich") %>%
  dplyr::select(Barcode, eta, prob_low)

df_surv_z <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::select(Barcode, COMPERA_number, DeathDAYS, Death) %>%
  dplyr::left_join(score_z, by = "Barcode") %>%
  dplyr::mutate(
    COMPERA_number = as.numeric(COMPERA_number),
    Death_event = ifelse(Death %in% c("yes", "Yes", "TRUE", "1", 1), 1, 0),
    DeathDAYS = as.numeric(DeathDAYS)
  ) %>%
  dplyr::filter(!is.na(DeathDAYS), !is.na(Death_event), !is.na(COMPERA_number), !is.na(eta))

# ---- 2) Fit Cox models ----
m0 <- coxph(Surv(DeathDAYS, Death_event) ~ COMPERA_number, data = df_surv_z)
m1 <- coxph(Surv(DeathDAYS, Death_event) ~ eta, data = df_surv_z)
m2 <- coxph(Surv(DeathDAYS, Death_event) ~ COMPERA_number + eta, data = df_surv_z)

# ---- 3) Summaries (HR, CI, p) ----
tab0 <- broom::tidy(m0, exponentiate = TRUE, conf.int = TRUE)
tab1 <- broom::tidy(m1, exponentiate = TRUE, conf.int = TRUE)
tab2 <- broom::tidy(m2, exponentiate = TRUE, conf.int = TRUE)

tab0$model <- "COMPERA"
tab1$model <- "LASSO score (eta)"
tab2$model <- "COMPERA + LASSO score"

cox_tables <- dplyr::bind_rows(tab0, tab1, tab2)
cox_tables

# ---- 4) Likelihood ratio tests: does score add info beyond COMPERA? ----
anova_m0_m2 <- anova(m0, m2, test = "LRT")
anova_m0_m2

# ---- 5) C-index (concordance) ----
c0 <- summary(m0)$concordance[1]
c1 <- summary(m1)$concordance[1]
c2 <- summary(m2)$concordance[1]
data.frame(
  model = c("COMPERA", "LASSO score", "COMPERA + LASSO score"),
  c_index = c(c0, c1, c2)
)
```


```{r lasso_poland_train_zurich_preselect50_fit_and_score, message=FALSE, warning=FALSE}
library(glmnet)
library(pROC)
library(dplyr)

# ---- 1) Preselect proteins (training cohort only: Poland) ----
de_train <- DE_p_bin
de_train$score <- de_train$logFC * (-log10(de_train$P.Value))

DE_p_signif_2 <- DE_p %>% dplyr::filter(adj.P.Val < 0.05)

de_train_ord <- de_train[order(abs(de_train$score), decreasing = TRUE), ]

# keep your intersection logic; if fewer than 50, it's fine
top50 <- de_train_ord %>%
  dplyr::filter(Gene %in% DE_p_signif_2$Gene) %>%
  dplyr::slice_head(n = 25) %>%
  dplyr::pull(Gene)

length(top50)

# ---- 2) SSc-PAH only for classification ----
cutoff <- 0.32

p_sub <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

z_sub <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

y_tr <- ifelse(as.numeric(p_sub$TAPSE_sPAP) < cutoff, 1, 0)
y_te <- ifelse(as.numeric(z_sub$TAPSE_sPAP) < cutoff, 1, 0)

# ---- 3) Build X matrices (Age+Sex unpenalized + proteins penalized) ----
Xtr_prot <- t(as.matrix(data_p[top50, p_sub$Barcode, drop = FALSE]))
Xte_prot <- t(as.matrix(data_z[top50, z_sub$Barcode, drop = FALSE]))

cov_tr <- data.frame(Age = as.numeric(p_sub$Age), Sex = factor(p_sub$Sex))
cov_te <- data.frame(Age = as.numeric(z_sub$Age), Sex = factor(z_sub$Sex, levels = levels(cov_tr$Sex)))

Xtr_cov <- model.matrix(~ Age + Sex, data = cov_tr)[, -1, drop = FALSE]
Xte_cov <- model.matrix(~ Age + Sex, data = cov_te)[, -1, drop = FALSE]

Xtr <- cbind(Xtr_cov, Xtr_prot)
Xte <- cbind(Xte_cov, Xte_prot)

pf <- c(rep(0, ncol(Xtr_cov)), rep(1, ncol(Xtr_prot)))

# ---- 4) CV LASSO ----
set.seed(1)
cvfit_bin50 <- cv.glmnet(
  x = Xtr, y = y_tr,
  family = "binomial",
  alpha = 1,
  penalty.factor = pf,
  standardize = TRUE,
  nfolds = 5,
  type.measure = "auc"
)

# ---- 5) Choose lambda giving 3–10 proteins ----
lambdas <- cvfit_bin50$glmnet.fit$lambda
beta_mat <- as.matrix(coef(cvfit_bin50$glmnet.fit, s = lambdas))
feat_names <- rownames(beta_mat)

prot_names <- colnames(Xtr_prot)
prot_idx <- which(feat_names %in% prot_names)

nprot <- apply(beta_mat[prot_idx, , drop = FALSE] != 0, 2, sum)

ok <- which(nprot >= 3 & nprot <= 10)
if (length(ok) > 0) {
  j <- ok[which.min(cvfit_bin50$cvm[match(lambdas[ok], cvfit_bin50$lambda)])]
  lambda_bin50 <- lambdas[j]
} else {
  j <- which.min(abs(nprot - 5))
  lambda_bin50 <- lambdas[j]
}

beta_use <- as.matrix(coef(cvfit_bin50, s = lambda_bin50))
sel <- rownames(beta_use)[beta_use[,1] != 0]
sel_proteins_bin50 <- sel[sel %in% prot_names]
sel_proteins_bin50

# ---- 6) Predictions for ROC (SSc-PAH only) ----
p_tr_prob <- as.numeric(predict(cvfit_bin50, newx = Xtr, s = lambda_bin50, type = "response"))
p_te_prob <- as.numeric(predict(cvfit_bin50, newx = Xte, s = lambda_bin50, type = "response"))

auc_tr <- as.numeric(pROC::auc(pROC::roc(y_tr, p_tr_prob, quiet = TRUE)))
auc_te <- as.numeric(pROC::auc(pROC::roc(y_te, p_te_prob, quiet = TRUE)))

roc_tr <- pROC::roc(y_tr, p_tr_prob, quiet = TRUE)
roc_te <- pROC::roc(y_te, p_te_prob, quiet = TRUE)

ggplot() +
  geom_line(data = data.frame(tpr = rev(roc_tr$sensitivities), fpr = rev(1 - roc_tr$specificities)),
            aes(fpr, tpr), color = "red") +
  geom_line(data = data.frame(tpr = rev(roc_te$sensitivities), fpr = rev(1 - roc_te$specificities)),
            aes(fpr, tpr), color = "black") +
  geom_abline(linetype = 2) +
  labs(
    x = "False positive rate",
    y = "True positive rate",
    title = paste0(
      "Top50 preselected LASSO (Age+Sex)\n",
      "Train Poland AUC=", round(auc_tr, 2),
      " | Test Zurich AUC=", round(auc_te, 2)
    )
  ) +
  theme_classic()
```


---

## 2) Compute scores for **ALL patients** (add SSc-noPAH) + 3-group boxplot with p-values

This uses the *same fitted model* (`cvfit_bin50`, `lambda_bin50`, `top50`) and predicts scores for:

- Zurich: SSc-noPAH + SSc-PAH (with TAPSE/sPAP available for SSc-PAH to be grouped)
- Poland: same

```{r lasso_score_allpatients_3groups_boxplot, fig.height=4.8, fig.width=9, warning=FALSE, message=FALSE}
library(ggpubr)

# helper: build X with same sex levels as training
make_X <- function(clin_df, expr_mat, proteins, sex_levels_train) {
  cov_df <- data.frame(
    Age = as.numeric(clin_df$Age),
    Sex = factor(clin_df$Sex, levels = sex_levels_train)
  )
  X_cov <- model.matrix(~ Age + Sex, data = cov_df)[, -1, drop = FALSE]
  X_prot <- t(as.matrix(expr_mat[proteins, clin_df$Barcode, drop = FALSE]))
  cbind(X_cov, X_prot)
}

sex_levels_train <- levels(factor(p_sub$Sex))  # from training subset

# Zurich all (SSc-noPAH + SSc-PAH)
z_all <- clin_z %>%
  dplyr::filter(Group %in% c("SSc-noPAH", "SSc-PAH")) %>%
  dplyr::filter(!is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    group3 = dplyr::case_when(
      Group == "SSc-noPAH" ~ "SSc-noPAH",
      Group == "SSc-PAH" & !is.na(TAPSE_sPAP) & TAPSE_sPAP < 0.32 ~ "SSc-PAH <0.32",
      Group == "SSc-PAH" & !is.na(TAPSE_sPAP) & TAPSE_sPAP >= 0.32 ~ "SSc-PAH ≥0.32",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(group3))

# Poland all
p_all <- clin_p %>%
  dplyr::filter(Group %in% c("SSc-noPAH", "SSc-PAH")) %>%
  dplyr::filter(!is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    group3 = dplyr::case_when(
      Group == "SSc-noPAH" ~ "SSc-noPAH",
      Group == "SSc-PAH" & !is.na(TAPSE_sPAP) & TAPSE_sPAP < 0.32 ~ "SSc-PAH <0.32",
      Group == "SSc-PAH" & !is.na(TAPSE_sPAP) & TAPSE_sPAP >= 0.32 ~ "SSc-PAH ≥0.32",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(group3))

group_levels <- c("SSc-noPAH", "SSc-PAH ≥0.32", "SSc-PAH <0.32")
z_all$group3 <- factor(z_all$group3, levels = group_levels)
p_all$group3 <- factor(p_all$group3, levels = group_levels)

# predict scores
X_z_all <- make_X(z_all, data_z, top50, sex_levels_train)
X_p_all <- make_X(p_all, data_p, top50, sex_levels_train)

z_prob <- as.numeric(predict(cvfit_bin50, newx = X_z_all, s = lambda_bin50, type = "response"))
p_prob <- as.numeric(predict(cvfit_bin50, newx = X_p_all, s = lambda_bin50, type = "response"))

z_eta  <- as.numeric(predict(cvfit_bin50, newx = X_z_all, s = lambda_bin50, type = "link"))
p_eta  <- as.numeric(predict(cvfit_bin50, newx = X_p_all, s = lambda_bin50, type = "link"))

df_score3 <- dplyr::bind_rows(
  data.frame(cohort="Zurich", Barcode=z_all$Barcode, group3=z_all$group3, prob_low=z_prob, eta=z_eta),
  data.frame(cohort="Poland", Barcode=p_all$Barcode, group3=p_all$group3, prob_low=p_prob, eta=p_eta)
)

comparisons_list <- list(
  c("SSc-noPAH", "SSc-PAH ≥0.32"),
  c("SSc-noPAH", "SSc-PAH <0.32"),
  c("SSc-PAH ≥0.32", "SSc-PAH <0.32")
)

ggplot(df_score3, aes(group3, prob_low)) +
  geom_boxplot(outlier.size = 0.7) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ cohort) +
  stat_compare_means(
    comparisons = comparisons_list,
    method = "wilcox.test",
    label = "p.format",
    step.increase = 0.09
  ) +
  theme_classic() +
  labs(x = NULL, y = "LASSO score (predicted P[Low TAPSE/sPAP])")
```

---

## 3) Cox regression in Zurich SSc-PAH + C-index table

This produces exactly the table you want.

```{r , message=FALSE, warning=FALSE}
library(survival)
library(broom)
library(dplyr)

# Zurich SSc-PAH only, join LASSO eta score
score_z <- df_score3 %>%
  dplyr::filter(cohort == "Zurich") %>%
  dplyr::select(Barcode, eta)

df_surv_z <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::select(Barcode, COMPERA_number, DeathDAYS, Death) %>%
  dplyr::left_join(score_z, by = "Barcode") %>%
  dplyr::mutate(
    COMPERA_number = as.numeric(COMPERA_number),
    Death_event = ifelse(Death %in% c("yes","Yes","TRUE","1",1), 1, 0),
    DeathDAYS = as.numeric(DeathDAYS)
  ) %>%
  dplyr::filter(!is.na(DeathDAYS), !is.na(Death_event), !is.na(COMPERA_number), !is.na(eta))

m0 <- coxph(Surv(DeathDAYS, Death_event) ~ COMPERA_number, data = df_surv_z)
m1 <- coxph(Surv(DeathDAYS, Death_event) ~ eta, data = df_surv_z)
m2 <- coxph(Surv(DeathDAYS, Death_event) ~ COMPERA_number + eta, data = df_surv_z)

c0 <- summary(m0)$concordance[1]
c1 <- summary(m1)$concordance[1]
c2 <- summary(m2)$concordance[1]

data.frame(
  model = c("COMPERA", "LASSO score", "COMPERA + LASSO score"),
  c_index = c(c0, c1, c2)
)

# (optional) LRT improvement p-value
anova(m0, m2, test = "LRT")

cox_tables <- bind_rows(
  broom::tidy(m0, exponentiate = TRUE, conf.int = TRUE) %>% mutate(model="COMPERA"),
  broom::tidy(m1, exponentiate = TRUE, conf.int = TRUE) %>% mutate(model="LASSO score"),
  broom::tidy(m2, exponentiate = TRUE, conf.int = TRUE) %>% mutate(model="COMPERA + LASSO score")
)
cox_tables
```

## 

```{r panel_stability_prepare_data, message=FALSE, warning=FALSE}
library(dplyr)
library(glmnet)
library(pROC)
library(ggplot2)
library(qs2)

# 0) Make sure TAPSE groups exist (you already do this earlier, but safe)
clin_z <- clin_z %>%
  mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = case_when(
      is.na(TAPSE_sPAP) ~ NA_character_,
      TAPSE_sPAP < 0.32 ~ "<0.32",
      TRUE              ~ "≥0.32"
    ),
    TAPSE_group = factor(TAPSE_group, levels = c("≥0.32", "<0.32"))
  )

clin_p <- clin_p %>%
  mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = case_when(
      is.na(TAPSE_sPAP) ~ NA_character_,
      TAPSE_sPAP < 0.32 ~ "<0.32",
      TRUE              ~ "≥0.32"
    ),
    TAPSE_group = factor(TAPSE_group, levels = c("≥0.32", "<0.32"))
  )

# 1) Restrict to SSc-PAH for training/testing sets (as in your LASSO code)
train_p <- clin_p %>%
  filter(Group == "SSc-PAH") %>%
  filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  filter(Barcode %in% colnames(data_p))

test_z <- clin_z %>%
  filter(Group == "SSc-PAH") %>%
  filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  filter(Barcode %in% colnames(data_z))

# 2) Align proteins between cohorts
common_proteins <- intersect(rownames(data_p), rownames(data_z))
expr_p <- as.matrix(data_p[common_proteins, , drop = FALSE])
expr_z <- as.matrix(data_z[common_proteins, , drop = FALSE])

# 3) Keep only samples we use for model fit/test
expr_p_tr <- expr_p[, train_p$Barcode, drop = FALSE]
expr_z_te <- expr_z[, test_z$Barcode, drop = FALSE]

dim(expr_p_tr)
dim(expr_z_te)

# Define outcome (1 = low TAPSE/sPAP)
y_p <- ifelse(train_p$TAPSE_sPAP < 0.32, 1, 0)
y_z <- ifelse(test_z$TAPSE_sPAP < 0.32, 1, 0)
table(y_p); table(y_z)
```


---

## Chunk 2 — stability-selection bootstrap LASSO (Age/Sex unpenalized)

This runs **B=500** bootstraps in **Poland only** and returns:
- selection frequency
- sign stability
- coefficient summaries
- OOB AUC per bootstrap (so you get a “typical performance” distribution without using Zurich)

```{r panel_stability_bootstrap_lasso_functions, message=FALSE, warning=FALSE}
library(Matrix)

make_design <- function(expr_mat, clin_df, include_age = TRUE, include_sex = TRUE){
  # expr_mat: proteins x samples, columns match clin_df$Barcode (in same order)
  X_prot <- t(expr_mat[, clin_df$Barcode, drop = FALSE])  # samples x proteins

  cov_parts <- list()
  if (include_age) {
    cov_parts$Age <- as.numeric(clin_df$Age)
  }
  if (include_sex) {
    cov_parts$Sex <- factor(clin_df$Sex)
  }

  if (length(cov_parts) == 0) {
    X <- X_prot
    pf <- rep(1, ncol(X))
    cov_names <- character(0)
  } else {
    cov_df <- as.data.frame(cov_parts)
    # model.matrix handles Sex dummy coding automatically
    X_cov <- model.matrix(~ ., data = cov_df)[, -1, drop = FALSE]  # remove intercept
    X <- cbind(X_cov, X_prot)

    # unpenalize covariates, penalize proteins
    pf <- c(rep(0, ncol(X_cov)), rep(1, ncol(X_prot)))
    cov_names <- colnames(X_cov)
  }

  list(X = X, pf = pf, cov_names = cov_names, prot_names = colnames(X_prot))
}

choose_lambda_by_protein_count <- function(cvfit, prot_names, cov_names,
                                          min_prot = 3, max_prot = 15){
  # choose a lambda that yields between min_prot and max_prot proteins (closest if none)
  lambdas <- cvfit$glmnet.fit$lambda
  beta_mat <- as.matrix(coef(cvfit$glmnet.fit, s = lambdas))

  feat_names <- rownames(beta_mat)
  prot_idx <- which(feat_names %in% prot_names)

  nprot <- apply(beta_mat[prot_idx, , drop = FALSE] != 0, 2, sum)

  ok <- which(nprot >= min_prot & nprot <= max_prot)
  if (length(ok) > 0) {
    # among ok, pick lambda with best CV AUC (min cvm because cv.glmnet stores -AUC? actually for AUC it stores 1-AUC)
    # safer: pick ok lambda that minimizes cvfit$cvm
    j <- ok[which.min(cvfit$cvm[match(lambdas[ok], cvfit$lambda)])]
    return(lambdas[j])
  }

  # if none in range, choose the lambda whose protein count is closest to target mid
  target <- round((min_prot + max_prot)/2)
  j <- which.min(abs(nprot - target))
  lambdas[j]
}

bootstrap_stability_lasso <- function(expr_mat, clin_df, y,
                                     B = 500,
                                     include_age = TRUE,
                                     include_sex = TRUE,
                                     alpha = 1,
                                     nfolds = 5,
                                     min_prot = 3,
                                     max_prot = 15,
                                     seed = 1){
  set.seed(seed)

  # Ensure order consistency
  stopifnot(all(clin_df$Barcode %in% colnames(expr_mat)))
  expr_mat <- expr_mat[, clin_df$Barcode, drop = FALSE]
  y <- as.numeric(y)

  des <- make_design(expr_mat, clin_df, include_age, include_sex)
  X_full <- des$X
  pf <- des$pf
  prot_names <- des$prot_names
  cov_names  <- des$cov_names

  n <- nrow(X_full)
  all_prots <- prot_names

  # storage
  sel_list <- vector("list", B)
  coef_list <- vector("list", B)
  auc_oob <- rep(NA_real_, B)
  lambda_used <- rep(NA_real_, B)

  for (b in seq_len(B)) {
    idx_boot <- sample(seq_len(n), size = n, replace = TRUE)
    idx_oob  <- setdiff(seq_len(n), unique(idx_boot))

    Xb <- X_full[idx_boot, , drop = FALSE]
    yb <- y[idx_boot]

    # CV LASSO on bootstrap sample
    cvfit <- cv.glmnet(
      x = Xb, y = yb,
      family = "binomial",
      alpha = alpha,
      penalty.factor = pf,
      standardize = TRUE,
      nfolds = nfolds,
      type.measure = "auc"
    )

    lam <- choose_lambda_by_protein_count(cvfit, prot_names = all_prots, cov_names = cov_names,
                                          min_prot = min_prot, max_prot = max_prot)
    lambda_used[b] <- lam

    beta <- as.matrix(coef(cvfit$glmnet.fit, s = lam))
    beta <- beta[, 1]
    names(beta) <- rownames(as.matrix(coef(cvfit$glmnet.fit, s = lam)))

    # selected proteins only
    sel_prot <- intersect(names(beta)[beta != 0], all_prots)
    sel_list[[b]] <- sel_prot
    coef_list[[b]] <- beta

    # OOB AUC (if enough OOB samples + both classes)
    if (length(idx_oob) >= 8 && length(unique(y[idx_oob])) == 2) {
      # fit final model on bootstrap sample at chosen lambda, predict on OOB
      fitb <- glmnet(
        x = Xb, y = yb,
        family = "binomial",
        alpha = alpha,
        penalty.factor = pf,
        standardize = TRUE,
        lambda = lam
      )
      p_oob <- as.numeric(predict(fitb, newx = X_full[idx_oob, , drop = FALSE], type = "response"))
      auc_oob[b] <- as.numeric(pROC::auc(pROC::roc(y[idx_oob], p_oob, quiet = TRUE)))
    }
  }

  # summarise selection frequency + sign stability
  sel_counts <- table(unlist(sel_list))
  freq <- as.numeric(sel_counts[all_prots])
  freq[is.na(freq)] <- 0
  freq <- freq / B

  # sign stability: among bootstraps where selected, fraction with positive coefficient
  sign_pos <- rep(NA_real_, length(all_prots))
  names(sign_pos) <- all_prots

  for (p in all_prots) {
    signs <- c()
    for (b in seq_len(B)) {
      beta <- coef_list[[b]]
      if (!is.null(beta) && p %in% names(beta) && beta[p] != 0) {
        signs <- c(signs, sign(beta[p]))
      }
    }
    if (length(signs) > 0) {
      sign_pos[p] <- mean(signs > 0)
    }
  }

  res <- data.frame(
    protein = all_prots,
    select_freq = freq,
    sign_pos_frac = sign_pos,
    stringsAsFactors = FALSE
  ) %>%
    arrange(desc(select_freq))

  list(
    summary = res,
    sel_list = sel_list,
    coef_list = coef_list,
    auc_oob = auc_oob,
    lambda_used = lambda_used,
    settings = list(B=B, include_age=include_age, include_sex=include_sex,
                    alpha=alpha, nfolds=nfolds, min_prot=min_prot, max_prot=max_prot)
  )
}
```


---

## Chunk 3 — run bootstrap (Poland only) + save objects

```{r panel_stability_bootstrap_run, message=FALSE, warning=FALSE}
B <- 500

stab_age_sex <- bootstrap_stability_lasso(
  expr_mat = expr_p_tr,
  clin_df  = train_p,
  y        = y_p,
  B = B,
  include_age = TRUE,
  include_sex = TRUE,     # set FALSE if you want "Age only"
  alpha = 1,
  nfolds = 5,
  min_prot = 3,
  max_prot = 15,
  seed = 1
)

qs2::qs_save(stab_age_sex, here::here(output_dir_data, paste0("stability_lasso_B", B, "_AgeSex.qs2")))
head(stab_age_sex$summary, 20)
summary(stab_age_sex$auc_oob)
```


---

## Chunk 4 — plots: selection frequency + OOB AUC distribution

```{r panel_stability_plots, fig.height=4.5, fig.width=9, message=FALSE, warning=FALSE}
stab_df <- stab_age_sex$summary

# Top selected proteins
topN <- 25
p1 <- stab_df %>%
  slice_head(n = topN) %>%
  ggplot(aes(x = reorder(protein, select_freq), y = select_freq)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Selection frequency (bootstrap)",
    title = paste0("Stability selection (Poland, B=", B, "): top ", topN, " proteins")
  ) +
  theme_classic(base_size = 12)

# AUC distribution (OOB)
auc_df <- data.frame(auc_oob = stab_age_sex$auc_oob) %>%
  filter(!is.na(auc_oob))

p2 <- ggplot(auc_df, aes(x = auc_oob)) +
  geom_histogram(bins = 25) +
  labs(
    x = "OOB AUC (Poland bootstrap)",
    y = "Count",
    title = paste0("Bootstrap OOB AUC distribution (B=", B, ")")
  ) +
  theme_classic(base_size = 12)

p1
ggsave(here::here(output_dir_figs, paste0("stability_top", topN, "_B", B, "_AgeSex.png")),
       plot = p1, width = 9, height = 4.5, dpi = 300)

p2
ggsave(here::here(output_dir_figs, paste0("stability_auc_oob_B", B, "_AgeSex.png")),
       plot = p2, width = 7, height = 4.2, dpi = 300)

# Print helpful summary numbers
cat("OOB AUC mean:", round(mean(auc_df$auc_oob), 3), "\n")
cat("OOB AUC median:", round(median(auc_df$auc_oob), 3), "\n")
cat("OOB AUC 2.5%:", round(quantile(auc_df$auc_oob, 0.025), 3), "\n")
cat("OOB AUC 97.5%:", round(quantile(auc_df$auc_oob, 0.975), 3), "\n")
```


Pick one strategy. I’m giving both:

```{r panel_stability_choose_panel, message=FALSE, warning=FALSE}
stab_df <- stab_age_sex$summary

# Option A: threshold on selection frequency
freq_thresh <- 0.20  # 40% is strict; try 0.30 too
panel_A <- stab_df %>%
  filter(select_freq >= freq_thresh) %>%
  pull(protein)

# Option B: fixed-size panel
panel_size <- 10
panel_B <- stab_df %>%
  slice_head(n = panel_size) %>%
  pull(protein)

length(panel_A)
panel_A

panel_B
```


---

## Chunk 6 — build **Option 2** (refit logistic regression) + **Option 3** (signed mean z-score)

Key points implemented correctly:
- **Standardize proteins using Poland mean/SD** and apply the same scaling to Zurich
- Provide a function to compute:
  - logistic refit score (eta or probability)
  - signed mean z-score

```{r panel_scores_refit_and_signed, message=FALSE, warning=FALSE}
# Choose which panel to use
panel_prots <- panel_A   # or panel_A

# Helper: Poland-based scaling for a set of proteins
get_scaler <- function(expr_mat, proteins){
  m <- rowMeans(expr_mat[proteins, , drop = FALSE], na.rm = TRUE)
  s <- apply(expr_mat[proteins, , drop = FALSE], 1, sd, na.rm = TRUE)
  s[s == 0 | is.na(s)] <- 1
  list(mu = m, sd = s)
}

apply_scaler <- function(expr_mat, proteins, scaler){
  Z <- expr_mat[proteins, , drop = FALSE]
  Z <- sweep(Z, 1, scaler$mu, "-")
  Z <- sweep(Z, 1, scaler$sd, "/")
  Z[is.na(Z)] <- 0
  Z
}

# ---- Option 2: refit logistic regression on Poland ----
fit_logistic_refit <- function(expr_mat, clin_df, proteins, cutoff = 0.32,
                               include_age = TRUE, include_sex = TRUE){
  # expr_mat: proteins x samples
  # clin_df order must match columns used
  df <- clin_df %>% mutate(y = ifelse(TAPSE_sPAP < cutoff, 1, 0))
  stopifnot(all(df$Barcode %in% colnames(expr_mat)))

  # Poland scaling inside this cohort (to be re-used later)
  scaler <- get_scaler(expr_mat[, df$Barcode, drop = FALSE], proteins)
  Z <- t(apply_scaler(expr_mat[, df$Barcode, drop = FALSE], proteins, scaler)) # samples x proteins
  Z <- as.data.frame(Z)
  Z$Barcode <- df$Barcode

  dd <- df %>% dplyr::select(Barcode, y, Age, Sex) %>% left_join(Z, by = "Barcode")

  # formula
  rhs <- c()
  if (include_age) rhs <- c(rhs, "Age")
  if (include_sex) rhs <- c(rhs, "Sex")
  rhs <- c(rhs, proteins)
  fml <- as.formula(paste("y ~", paste(rhs, collapse = " + ")))

  m <- glm(fml, data = dd, family = binomial())
  list(model = m, scaler = scaler, proteins = proteins,
       include_age = include_age, include_sex = include_sex)
}

predict_logistic_refit <- function(fit, expr_mat, clin_df){
  df <- clin_df
  stopifnot(all(df$Barcode %in% colnames(expr_mat)))

  Z <- t(apply_scaler(expr_mat[, df$Barcode, drop = FALSE], fit$proteins, fit$scaler))
  Z <- as.data.frame(Z)
  Z$Barcode <- df$Barcode

  dd <- df %>% dplyr::select(Barcode, Age, Sex, TAPSE_sPAP, TAPSE_group, Group) %>%
    left_join(Z, by = "Barcode")

  eta <- as.numeric(predict(fit$model, newdata = dd, type = "link"))
  p   <- as.numeric(predict(fit$model, newdata = dd, type = "response"))

  dd$score_eta <- eta
  dd$score_p   <- p
  dd
}

# ---- Option 3: signed mean z-score ----
signed_mean_score <- function(expr_mat, clin_df, proteins, sign_map, scaler){
  # sign_map: named vector (protein -> +1/-1)
  stopifnot(all(names(sign_map) %in% proteins))

  Z <- apply_scaler(expr_mat[, clin_df$Barcode, drop = FALSE], proteins, scaler) # proteins x samples
  s <- sign_map[proteins]
  Zs <- sweep(Z, 1, s, "*")
  score <- colMeans(Zs, na.rm = TRUE)

  out <- clin_df %>% dplyr::select(Barcode, Age, Sex, TAPSE_sPAP, TAPSE_group, Group)
  out$score_signed <- as.numeric(score[match(out$Barcode, names(score))])
  out
}

# Build sign map from bootstrap results: sign = +1 if mostly positive
sign_map <- stab_age_sex$summary %>%
  filter(protein %in% panel_prots) %>%
  mutate(sign = ifelse(sign_pos_frac >= 0.5, 1, -1)) %>%
  dplyr::select(protein, sign) %>%
  tibble::deframe()

# Fit refit-logistic in Poland (SSc-PAH train only)
refit <- fit_logistic_refit(
  expr_mat = expr_p,
  clin_df  = train_p,
  proteins = panel_prots,
  cutoff = 0.32,
  include_age = TRUE,
  include_sex = TRUE
)

# Predict in Poland SSc-PAH (train) and Zurich SSc-PAH (test)
pred_p <- predict_logistic_refit(refit, expr_p, train_p)
pred_z <- predict_logistic_refit(refit, expr_z, test_z)

# Signed score uses the SAME Poland scaler as refit (so fully portable)
signed_p <- signed_mean_score(expr_p, train_p, panel_prots, sign_map, refit$scaler)
signed_z <- signed_mean_score(expr_z, test_z, panel_prots, sign_map, refit$scaler)

head(pred_z)
head(signed_z)
```


---

## Chunk 7 — evaluate in Zurich: ROC + your 3-group boxplot

### ROC for SSc-PAH-only test set (Zurich)
```{r panel_scores_eval_roc, fig.height=3.8, fig.width=5.2, message=FALSE, warning=FALSE}
# Outcomes
y_tr <- ifelse(pred_p$TAPSE_sPAP < 0.32, 1, 0)
y_te <- ifelse(pred_z$TAPSE_sPAP < 0.32, 1, 0)

# AUCs (refit logistic)
roc_tr <- pROC::roc(y_tr, pred_p$score_p, quiet = TRUE)
roc_te <- pROC::roc(y_te, pred_z$score_p, quiet = TRUE)

auc_tr <- as.numeric(pROC::auc(roc_tr))
auc_te <- as.numeric(pROC::auc(roc_te))

plot(roc_tr, col="red",
     main=paste0("Stable panel refit-logistic\nTrain Poland AUC=", round(auc_tr,2),
                 " | Test Zurich AUC=", round(auc_te,2)))
lines(roc_te, col="black")
abline(a=0, b=1, lty=2)
legend("bottomright", legend=c("Poland (train)", "Zurich (test)"),
       col=c("red","black"), lwd=2, bty="n")

ggsave(here::here(output_dir_figs, paste0("stablePanel_refitLogistic_ROC.png")),
       width = 5.2, height = 3.8, dpi = 300)
```
       

### 3-group boxplot (SSc-noPAH, SSc-PAH≥0.32, SSc-PAH<0.32)
This mimics your figure logic but uses the **new score**. We apply the score to **all patients** in each cohort (not only SSc-PAH), so you can show separation from SSc-noPAH as well.

```{r panel_scores_boxplot_3groups, fig.height=4.8, fig.width=9, message=FALSE, warning=FALSE}
library(ggpubr)

# Build 3-group labels for plotting
make_3groups <- function(clin_df){
  clin_df %>%
    mutate(
      group3 = case_when(
        Group == "SSc-noPAH" ~ "SSc-noPAH",
        Group == "SSc-PAH" & TAPSE_sPAP >= 0.32 ~ "SSc-PAH≥0.32",
        Group == "SSc-PAH" & TAPSE_sPAP < 0.32  ~ "SSc-PAH<0.32",
        TRUE ~ NA_character_
      ),
      group3 = factor(group3, levels = c("SSc-noPAH", "SSc-PAH≥0.32", "SSc-PAH<0.32"))
    )
}

# Score ALL subjects in each cohort (use refit logistic)
clin_p_all <- clin_p %>% filter(Barcode %in% colnames(expr_p), !is.na(Age), !is.na(Sex))
clin_z_all <- clin_z %>% filter(Barcode %in% colnames(expr_z), !is.na(Age), !is.na(Sex))

pred_p_all <- predict_logistic_refit(refit, expr_p, clin_p_all)
pred_z_all <- predict_logistic_refit(refit, expr_z, clin_z_all)

pred_p_all <- make_3groups(pred_p_all)
pred_z_all <- make_3groups(pred_z_all)

df_plot <- bind_rows(
  pred_p_all %>% mutate(Cohort = "Poland"),
  pred_z_all %>% mutate(Cohort = "Zurich")
) %>%
  filter(!is.na(group3))

# Boxplot of probability score (predicted P[Low TAPSE/sPAP])
p <- ggplot(df_plot, aes(x = group3, y = score_p)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.3, alpha = 0.7) +
  facet_wrap(~Cohort, scales = "free_x") +
  labs(x = NULL, y = "Score (predicted P[Low TAPSE/sPAP])",
       title = paste0("Stable panel refit-logistic score (", length(panel_prots), " proteins)")) +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

p
ggsave(here::here(output_dir_figs, paste0("stablePanel_refitLogistic_3groups_boxplot.png")),
       plot = p, width = 9, height = 4.8, dpi = 300)
```
       

---

## Optional (but very useful) — same evaluation for **signed mean z-score**
If you want the “super clinic-friendly index” plot/ROC too:

```{r panel_scores_signed_eval, fig.height=3.8, fig.width=5.2, message=FALSE, warning=FALSE}
# ROC on SSc-PAH-only (Zurich test)
y_te <- ifelse(signed_z$TAPSE_sPAP < 0.32, 1, 0)
roc_signed <- pROC::roc(y_te, signed_z$score_signed, quiet = TRUE)
auc_signed <- as.numeric(pROC::auc(roc_signed))

plot(roc_signed,
     main=paste0("Signed mean z-score (", length(panel_prots), " proteins)\nZurich AUC=", round(auc_signed,2)))
abline(a=0, b=1, lty=2)
ggsave(here::here(output_dir_figs, paste0("stablePanel_signedMeanZ_ROC.png")),
       width = 5.2, height = 3.8, dpi = 300)
```


## Final lasso

```{r preselect_proteins_DE_Poland, message=FALSE, warning=FALSE}
library(limma)
library(dplyr)

# Poland samples for DE (PAH vs noPAH)
de_meta_p <- clin_p %>%
  filter(Group %in% c("SSc-PAH", "SSc-noPAH")) %>%
  filter(!is.na(Age), !is.na(Sex)) %>%
  filter(Barcode %in% colnames(data_p))

X_de <- as.matrix(data_p[, de_meta_p$Barcode, drop = FALSE])

# Design: Group + Age + Sex
de_meta_p$Group <- factor(de_meta_p$Group, levels = c("SSc-noPAH", "SSc-PAH"))
de_meta_p$Sex   <- factor(de_meta_p$Sex)

design <- model.matrix(~ Group + Age + Sex, data = de_meta_p)

fit <- lmFit(X_de, design)
fit <- eBayes(fit)

DE_pah_vs_nopah <- topTable(fit, coef = "GroupSSc-PAH", number = Inf, adjust.method = "BH")
DE_pah_vs_nopah$protein <- rownames(DE_pah_vs_nopah)

# Choose your filter (match your original logic)
# Example: adj.P.Val < 0.05 (optionally add |logFC| cutoff)
preselect_proteins <- DE_pah_vs_nopah %>%
  dplyr::filter(adj.P.Val < 0.05) %>%
 # dplyr::filter(protein %in% dplyr::filter(DE_p_bin, P.Value < 0.05)$Gene) %>%
  pull(protein)

length(preselect_proteins)
head(preselect_proteins)

qs2::qs_save(DE_pah_vs_nopah, here::here(output_dir_data, "DE_Poland_SScPAH_vs_SScnoPAH_AgeSex.qs2"))
qs2::qs_save(preselect_proteins, here::here(output_dir_data, "preselect_proteins_DEadjP005.qs2"))

```


> If you want to mimic your “Top50 preselect”, replace the filter with `slice_min(adj.P.Val, n = 50)` (or rank by your score).

---

# Chunk B — re-run the stability-selection LASSO using only the preselected proteins

This is the **only change** you need: restrict `expr_p_tr` and `expr_z_te` to the preselected proteins (and keep the intersection with Zurich).


```{r stability_lasso_with_preselection, message=FALSE, warning=FALSE}
# Intersect with proteins available in BOTH cohorts
common_preselect <- intersect(preselect_proteins, intersect(rownames(data_p), rownames(data_z)))
length(common_preselect)

# Build training/test expression matrices for LASSO (proteins x samples)
expr_p_tr_pre <- as.matrix(data_p[common_preselect, train_p$Barcode, drop = FALSE])
expr_z_te_pre <- as.matrix(data_z[common_preselect, test_z$Barcode, drop = FALSE])

# Run bootstrap stability selection on the preselected feature space
B <- 500
stab_pre <- bootstrap_stability_lasso(
  expr_mat = expr_p_tr_pre,
  clin_df  = train_p,
  y        = y_p,
  B = B,
  include_age = TRUE,
  include_sex = TRUE,    # or FALSE if you want Age-only adjustment
  alpha = 1,
  nfolds = 5,
  min_prot = 3,
  max_prot = 100,
  seed = 1
)

qs2::qs_save(stab_pre, here::here(output_dir_data, paste0("stability_lasso_preselected_B", B, "_AgeSex.qs2")))
head(stab_pre$summary, 20)
```


---



```{r stability_panel_size_distribution, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

# number of proteins selected per bootstrap
panel_sizes <- sapply(stab_pre$sel_list, length)

df_sizes <- data.frame(size = panel_sizes)

p_size <- ggplot(df_sizes, aes(x = size)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = sort(unique(panel_sizes))) +
  labs(
    x = "Number of proteins selected (per bootstrap)",
    y = "Count",
    title = "Stability selection: distribution of panel sizes"
  ) +
  theme_classic(base_size = 12)

p_size
ggsave(here::here(output_dir_figs, paste0("stability_panel_size_distribution_B", stab_pre$settings$B, ".png")),
       p_size, width = 6, height = 4.2, dpi = 300)

summary(panel_sizes)
table(panel_sizes)
```


---

## 2) “Most frequent combination” (top recurring panels)
This is useful, but usually you’ll see many singletons unless you restrict to a manageable feature universe. So we compute frequency of **exact sets**.

```{r stability_most_frequent_panels, message=FALSE, warning=FALSE}
library(tibble)

panel_keys <- sapply(stab_pre$sel_list, function(v){
  if (length(v) == 0) return(NA_character_)
  paste(sort(v), collapse = " | ")
})

panel_tab <- sort(table(panel_keys), decreasing = TRUE)
panel_df <- data.frame(
  panel = names(panel_tab),
  n = as.integer(panel_tab),
  freq = as.integer(panel_tab) / length(stab_pre$sel_list),
  stringsAsFactors = FALSE
) %>% 
  filter(!is.na(panel))

head(panel_df, 20)
```


**Optional plot (top 10 panels):**
```{r stability_most_frequent_panels_plot, fig.height=4.5, fig.width=12, message=FALSE, warning=FALSE}
topM <- 10
p_panels <- panel_df %>%
  slice_head(n = topM) %>%
  mutate(panel = factor(panel, levels = rev(panel))) %>%
  ggplot(aes(x = panel, y = freq)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Frequency", title = paste0("Top ", topM, " most frequent exact panels")) +
  theme_classic(base_size = 12)

p_panels
ggsave(here::here(output_dir_figs, paste0("stability_top", topM, "_panels_B", stab_pre$settings$B, ".png")),
       p_panels, width = 9, height = 4.5, dpi = 300)
```


---

## 3) Co-selection heatmap (which proteins tend to appear together)
This is *often more informative* than exact combinations.

```{r stability_co_selection_heatmap, message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
library(Matrix)

# pick a manageable set (e.g. top 30 by selection freq)
topK <- 30
top_prots <- stab_pre$summary %>% slice_head(n = topK) %>% pull(protein)

# build co-selection matrix
M <- matrix(0, nrow = topK, ncol = topK, dimnames = list(top_prots, top_prots))

for (sel in stab_pre$sel_list) {
  sel <- intersect(sel, top_prots)
  if (length(sel) >= 2) {
    idx <- match(sel, top_prots)
    M[idx, idx] <- M[idx, idx] + 1
  } else if (length(sel) == 1) {
    i <- match(sel, top_prots)
    M[i, i] <- M[i, i] + 1
  }
}

# convert to frequency
M_freq <- M / length(stab_pre$sel_list)

# quick base heatmap
heatmap(M_freq, symm = TRUE, margins = c(8, 8),
        main = paste0("Co-selection frequency (top ", topK, " proteins)"))
```
        

       
# Chunk C — build final panel + refit-logistic + signed score (using preselection results)

```{r build_scores_from_stability_preselected, message=FALSE, warning=FALSE}
stab_df <- stab_pre$summary

# pick your panel
panel_size <- 5
panel_prots <- stab_df %>% slice_head(n = panel_size) %>% pull(protein)
panel_prots

# sign map from bootstrap
sign_map <- stab_df %>%
  filter(protein %in% panel_prots) %>%
  mutate(sign = ifelse(sign_pos_frac >= 0.5, 1, -1)) %>%
  dplyr::select(protein, sign) %>%
  tibble::deframe()

# IMPORTANT: refit/scores must use the FULL cohort matrices (data_p/data_z)
# but only these proteins
refit_pre <- fit_logistic_refit(
  expr_mat = as.matrix(data_p),  # proteins x samples
  clin_df  = train_p,
  proteins = panel_prots,
  cutoff = 0.32,
  include_age = TRUE,
  include_sex = TRUE
)

pred_p <- predict_logistic_refit(refit_pre, as.matrix(data_p), train_p)
pred_z <- predict_logistic_refit(refit_pre, as.matrix(data_z), test_z)

signed_p <- signed_mean_score(as.matrix(data_p), train_p, panel_prots, sign_map, refit_pre$scaler)
signed_z <- signed_mean_score(as.matrix(data_z), test_z, panel_prots, sign_map, refit_pre$scaler)
```


---

# Chunk D — FIXED plotting: ROC + boxplots 

## D1) ROC (same as before)

```{r roc_refit_preselected, fig.height=3.8, fig.width=5.2, message=FALSE, warning=FALSE}
y_tr <- ifelse(pred_p$TAPSE_sPAP < 0.32, 1, 0)
y_te <- ifelse(pred_z$TAPSE_sPAP < 0.32, 1, 0)

roc_tr <- pROC::roc(y_tr, pred_p$score_p, quiet = TRUE)
roc_te <- pROC::roc(y_te, pred_z$score_p, quiet = TRUE)

auc_tr <- as.numeric(pROC::auc(roc_tr))
auc_te <- as.numeric(pROC::auc(roc_te))

plot(roc_tr, col="red",
     main=paste0("Stable panel refit-logistic\nTrain Poland AUC=", round(auc_tr,2),
                 " | Test Zurich AUC=", round(auc_te,2)))
lines(roc_te, col="black")
abline(a=0, b=1, lty=2)
legend("bottomright", legend=c("Poland (train)", "Zurich (test)"),
       col=c("red","black"), lwd=2, bty="n")
```


## D2) Better “3-group” plot: use **linear predictor** (`score_eta`) not probability
Also: it’s more defensible to show this plot **within PAH** (your prediction task), and keep “noPAH” only as exploratory.

```{r boxplot_eta_preselected, fig.height=4.8, fig.width=9, message=FALSE, warning=FALSE}
library(ggpubr)

make_3groups <- function(clin_df){
  clin_df %>%
    mutate(
      group3 = case_when(
        Group == "SSc-noPAH" ~ "SSc-noPAH",
        Group == "SSc-PAH" & TAPSE_sPAP >= 0.32 ~ "SSc-PAH≥0.32",
        Group == "SSc-PAH" & TAPSE_sPAP < 0.32  ~ "SSc-PAH<0.32",
        TRUE ~ NA_character_
      ),
      group3 = factor(group3, levels = c("SSc-noPAH", "SSc-PAH≥0.32", "SSc-PAH<0.32"))
    )
}

# score ALL subjects to visualize, but plot ETA
clin_p_all <- clin_p %>% filter(Barcode %in% colnames(data_p), !is.na(Age), !is.na(Sex))
clin_z_all <- clin_z %>% filter(Barcode %in% colnames(data_z), !is.na(Age), !is.na(Sex))

pred_p_all <- predict_logistic_refit(refit_pre, as.matrix(data_p), clin_p_all) %>% make_3groups()
pred_z_all <- predict_logistic_refit(refit_pre, as.matrix(data_z), clin_z_all) %>% make_3groups()

df_plot <- bind_rows(
  pred_p_all %>% mutate(Cohort = "Poland"),
  pred_z_all %>% mutate(Cohort = "Zurich")
) %>% filter(!is.na(group3))

p <- ggplot(df_plot, aes(x = group3, y = score_eta)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.3, alpha = 0.7) +
  facet_wrap(~Cohort, scales = "free_x") +
  labs(x = NULL, y = "Score (log-odds; linear predictor)",
       title = paste0("Stable panel refit-logistic score (", length(panel_prots), " proteins)")) +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

p
```

## Correlations

```{r corr_prepare_pah_only, message=FALSE, warning=FALSE}

panel_prots <- c("CLSTN2","SIGLEC8","LRRN1","CXCL17","IL15")

# helper to build a long dataframe for correlations
make_corr_df <- function(clin_df, data_mat, cohort_label, score_df = NULL,
                         score_var = c("score_eta","score_p","score_signed")) {
  score_var <- match.arg(score_var)

  # subset to SSc-PAH with TAPSE/sPAP
  cc <- clin_df %>%
    filter(Group == "SSc-PAH") %>%
    filter(!is.na(TAPSE_sPAP)) %>%
    filter(Barcode %in% colnames(data_mat))

  # protein values (wide)
  prot_wide <- t(as.matrix(data_mat[panel_prots, cc$Barcode, drop = FALSE])) %>%
    as.data.frame()
  prot_wide$Barcode <- cc$Barcode

  out <- cc %>%
    dplyr::select(Barcode, TAPSE_sPAP) %>%
    left_join(prot_wide, by = "Barcode") %>%
    mutate(Cohort = cohort_label)

  # attach score if provided
  if (!is.null(score_df)) {
    out <- out %>%
      left_join(score_df %>% dplyr::select(Barcode, all_of(score_var)), by = "Barcode") %>%
      rename(Score = all_of(score_var))
  } else {
    out$Score <- NA_real_
  }

  out
}

# Use your previously created score data frames:
# pred_p, pred_z contain score_eta and score_p for SSc-PAH train/test
# signed_p, signed_z contain score_signed if you want that too

df_corr_p <- make_corr_df(train_p, data_p, "Poland", score_df = pred_p, score_var = "score_eta")
df_corr_z <- make_corr_df(test_z,  data_z, "Zurich", score_df = pred_z, score_var = "score_eta")

# Combine
df_corr <- bind_rows(df_corr_p, df_corr_z)

# quick sanity check
df_corr %>% group_by(Cohort) %>% summarise(n = n())
```

```{r corr_table_proteins_and_score, message=FALSE, warning=FALSE}
cor_spearman <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 5) return(c(rho = NA_real_, p = NA_real_, n = sum(ok)))
  ct <- suppressWarnings(cor.test(x[ok], y[ok], method = "spearman", exact = FALSE))
  c(rho = unname(ct$estimate), p = ct$p.value, n = sum(ok))
}

features <- c(panel_prots, "Score")

corr_tbl <- df_corr %>%
  pivot_longer(cols = all_of(panel_prots), names_to = "Feature", values_to = "Value") %>%
  # add Score as another feature
  bind_rows(
    df_corr %>% transmute(Cohort, Barcode, TAPSE_sPAP, Feature = "Score", Value = Score)
  ) %>%
  group_by(Cohort, Feature) %>%
  summarise(
    rho = cor_spearman(Value, TAPSE_sPAP)["rho"],
    p   = cor_spearman(Value, TAPSE_sPAP)["p"],
    n   = cor_spearman(Value, TAPSE_sPAP)["n"],
    .groups = "drop"
  ) %>%
  group_by(Cohort) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  arrange(Cohort, p_adj)

corr_tbl
qs2::qs_save(corr_tbl, here::here(output_dir_data, "corr_TAPSEsPAP_panelProteins_and_score_Spearman.qs2"))
write.csv(corr_tbl, here::here(output_dir_data, "corr_TAPSEsPAP_panelProteins_and_score_Spearman.csv"),
          row.names = FALSE)
```

```{r corr_plot_score_vs_tapse, fig.height=3.8, fig.width=7.5, message=FALSE, warning=FALSE}
# pull rho/p for annotation
ann_score <- corr_tbl %>% filter(Feature == "Score") %>% dplyr::select(Cohort, rho, p, p_adj)

df_score <- df_corr %>%
  filter(!is.na(Score)) %>%
  left_join(ann_score, by = "Cohort")

p_score <- ggplot(df_score, aes(x = TAPSE_sPAP, y = Score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Cohort, scales = "free_y") +
  labs(
    x = "TAPSE/sPAP",
    y = "Score (linear predictor, log-odds)",
    title = "Score vs TAPSE/sPAP in SSc-PAH"
  ) +
  theme_classic(base_size = 12)

# Add annotation per facet (simple approach)
p_score <- p_score +
  geom_text(
    data = df_score %>% distinct(Cohort, rho, p, p_adj),
    aes(x = -Inf, y = Inf,
        label = paste0("Spearman ρ=", round(rho, 2),
                       "\np=", signif(p, 2))),
    hjust = -1, vjust = 1.1,
    inherit.aes = FALSE
  )

p_score
ggsave(here::here(output_dir_figs, "corr_score_vs_TAPSEsPAP_SScPAH.png"),
       p_score, width = 7.5, height = 3.8, dpi = 300)
```

```{r corr_plot_proteins_vs_tapse_faceted, fig.height=5, fig.width=10.5, message=FALSE, warning=FALSE}
# prepare long data for plotting
df_long <- df_corr %>%
  pivot_longer(cols = all_of(panel_prots), names_to = "Protein", values_to = "NPX") %>%
  left_join(
    corr_tbl %>% filter(Feature %in% panel_prots) %>%
      rename(Protein = Feature) %>%
      dplyr::select(Cohort, Protein, rho, p, p_adj),
    by = c("Cohort", "Protein")
  )

p_prot <- ggplot(df_long, aes(x = TAPSE_sPAP, y = NPX)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(Cohort ~ Protein, scales = "free_y") +
  labs(
    x = "TAPSE/sPAP",
    y = "Protein NPX",
    title = "Individual proteins vs TAPSE/sPAP in SSc-PAH"
  ) +
  theme_classic(base_size = 12) +
  theme(
    strip.text.x = element_text(size = 10),
    strip.text.y = element_text(size = 10)
  )

# annotate each facet with rho and FDR
ann_df <- df_long %>%
  distinct(Cohort, Protein, rho, p_adj, p)

p_prot <- p_prot +
  geom_text(
    data = ann_df,
    aes(x = -Inf, y = Inf,
        label = paste0("ρ=", round(rho, 2), "\np=", signif(p, 2))),
    hjust = -1, vjust = 1.1,
    inherit.aes = FALSE,
    size = 3
  )

p_prot
ggsave(here::here(output_dir_figs, "corr_proteins_vs_TAPSEsPAP_SScPAH_faceted.png"),
       p_prot, width = 10.5, height = 7.5, dpi = 300)
```


```{r zurich_withincohort_scaling_auc, fig.height=3.8, fig.width=5.2, message=FALSE, warning=FALSE}
library(dplyr)
library(pROC)

panel_prots <- c("CLSTN2","SIGLEC8","LRRN1","CXCL17","IL15")

# If you already have sign_map from your stability selection, keep it.
# Otherwise, derive it from stab_pre$summary:
# sign_map <- stab_pre$summary %>%
#   filter(protein %in% panel_prots) %>%
#   mutate(sign = ifelse(sign_pos_frac >= 0.5, 1, -1)) %>%
#   select(protein, sign) %>%
#   tibble::deframe()

# --- Zurich SSc-PAH only ---
z_pah <- clin_z %>%
  filter(Group == "SSc-PAH") %>%
  filter(!is.na(TAPSE_sPAP)) %>%
  filter(Barcode %in% colnames(data_z))

# outcome
y_z <- ifelse(z_pah$TAPSE_sPAP < 0.32, 1, 0)

# protein matrix: proteins x samples
Zmat <- as.matrix(data_z[panel_prots, z_pah$Barcode, drop = FALSE])

# Zurich-within-cohort z-score per protein (center/scale across Zurich SSc-PAH samples)
mu_z <- rowMeans(Zmat, na.rm = TRUE)
sd_z <- apply(Zmat, 1, sd, na.rm = TRUE)
sd_z[sd_z == 0 | is.na(sd_z)] <- 1

Zz <- sweep(Zmat, 1, mu_z, "-")
Zz <- sweep(Zz,   1, sd_z, "/")
Zz[is.na(Zz)] <- 0

# signed mean z-score (higher should correspond to "worse" depending on sign_map)
s <- sign_map[panel_prots]
Zsigned <- sweep(Zz, 1, s, "*")
score_withinZ <- colMeans(Zsigned, na.rm = TRUE)

# ROC / AUC
roc_withinZ <- roc(y_z, as.numeric(score_withinZ), quiet = TRUE)
auc_withinZ <- as.numeric(auc(roc_withinZ))

plot(roc_withinZ,
     main = paste0("Zurich within-cohort scaled signed score\nAUC = ", round(auc_withinZ, 2)))
abline(a=0, b=1, lty=2)

ggsave(here::here(output_dir_figs, "AUC_Zurich_withinCohortScaling_signedScore.png"),
       width = 5.2, height = 3.8, dpi = 300)
```

## COX regression

```{r cox_prepare_surv_zurich, message=FALSE, warning=FALSE}
library(dplyr)
library(survival)

# -------------------------------
# 0) Choose which score to use
# -------------------------------
score_z <- pred_z %>%
  dplyr::select(Barcode, eta = score_eta)   # use linear predictor (recommended)

# -------------------------------
# 1) Robust survival column picker
# -------------------------------
pick_first_existing <- function(df, candidates) {
  hit <- candidates[candidates %in% colnames(df)]
  if (length(hit) == 0) return(NA_character_)
  hit[1]
}

# Try to guess your column names (adjust here if needed)
time_col  <- pick_first_existing(clin_z, c("DeathDAYS","death_days","FU_DAYS","followup_days","time_days","TimeDAYS","Time_days"))
event_col <- pick_first_existing(clin_z, c("Death","death","Death_event","event","Event","status","Status"))

# COMPERA numeric column (you mentioned COMPERA_number)
compera_col <- pick_first_existing(clin_z, c("COMPERA_number","COMPERA","COMPERA_num","COMPERA_score"))

cat("Using time_col =", time_col, "\n")
cat("Using event_col =", event_col, "\n")
cat("Using compera_col =", compera_col, "\n")

# If any are NA, set them manually like:
# time_col <- "DeathDAYS"
# event_col <- "Death"
# compera_col <- "COMPERA_number"

stopifnot(!is.na(time_col), !is.na(event_col), !is.na(compera_col))

# -------------------------------
# 2) Build Zurich SSc-PAH survival dataframe
# -------------------------------
df_surv_z <- clin_z %>%
  filter(Group == "SSc-PAH") %>%
  dplyr::select(Barcode, all_of(compera_col), all_of(time_col), all_of(event_col)) %>%
  left_join(score_z, by = "Barcode") %>%
  mutate(
    # numeric COMPERA
    COMPERA_number = as.numeric(.data[[compera_col]]),

    # time in days
    time_days = as.numeric(.data[[time_col]]),

    # event coding: try to handle yes/no, TRUE/FALSE, 1/0
    event_raw = .data[[event_col]],
    event = case_when(
      is.numeric(event_raw) ~ ifelse(event_raw > 0, 1, 0),
      is.logical(event_raw) ~ ifelse(event_raw, 1, 0),
      tolower(as.character(event_raw)) %in% c("yes","y","true","1","dead","death") ~ 1,
      tolower(as.character(event_raw)) %in% c("no","n","false","0","alive") ~ 0,
      TRUE ~ NA_real_
    )
  ) %>%
  filter(!is.na(time_days), !is.na(event), !is.na(COMPERA_number), !is.na(eta))

# quick check
df_surv_z %>% summarise(n=n(), events=sum(event), median_time=median(time_days))
table(df_surv_z$event)
```


---

## Chunk 2 — Cox models + c-index table (+ LRT)

```{r cox_zurich_cindex_table, message=FALSE, warning=FALSE}
library(dplyr)
library(survival)

m0 <- coxph(Surv(time_days, event) ~ COMPERA_number, data = df_surv_z, x = TRUE)
m1 <- coxph(Surv(time_days, event) ~ eta,            data = df_surv_z, x = TRUE)
m2 <- coxph(Surv(time_days, event) ~ COMPERA_number + eta, data = df_surv_z, x = TRUE)

c0 <- unname(summary(m0)$concordance[1])
c1 <- unname(summary(m1)$concordance[1])
c2 <- unname(summary(m2)$concordance[1])

cindex_tbl <- data.frame(
  model   = c("COMPERA", "Score", "COMPERA + score"),
  c_index = c(c0, c1, c2)
)

cindex_tbl

# Likelihood ratio test: does adding score improve COMPERA?
lrt <- anova(m0, m2, test = "LRT")
lrt
```


This gives you exactly the table format you showed (just with “Score” instead of “LASSO score” — rename if you want).

---

## Optional (recommended) — bootstrap CI for c-index difference (COMPERA vs COMPERA+score)

With small cohorts/events, c-index can be noisy. This gives you a quick sense of stability.

```{r cox_zurich_cindex_bootstrap, message=FALSE, warning=FALSE}
set.seed(1)
B <- 500

boot_delta <- replicate(B, {
  idx <- sample(seq_len(nrow(df_surv_z)), replace = TRUE)
  d <- df_surv_z[idx, , drop = FALSE]

  # handle rare bootstrap samples with 0 events
  if (length(unique(d$event)) < 2) return(NA_real_)

  m0b <- coxph(Surv(time_days, event) ~ COMPERA_number, data = d)
  m2b <- coxph(Surv(time_days, event) ~ COMPERA_number + eta, data = d)

  c0b <- unname(summary(m0b)$concordance[1])
  c2b <- unname(summary(m2b)$concordance[1])
  c2b - c0b
})

boot_delta <- boot_delta[is.finite(boot_delta)]
quantile(boot_delta, c(0.025, 0.5, 0.975))
mean(boot_delta)
```


---

## Optional — model summary table (HRs)
If you want a quick HR table for the combined model:

```{r cox_zurich_hr_table, message=FALSE, warning=FALSE}
library(broom)

tidy(m2, exponentiate = TRUE, conf.int = TRUE) %>%
  dplyr::select(term, estimate, conf.low, conf.high, p.value)
```

```{r plot_boot_delta, fig.height=3.6, fig.width=5.5}
library(ggplot2)
df_bd <- data.frame(delta_c = boot_delta)

ggplot(df_bd, aes(x = delta_c)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(x = "Δ C-index (COMPERA+score − COMPERA)",
       y = "Count",
       title = "Bootstrap distribution of c-index improvement") +
  theme_classic(base_size = 12)

```

```{r}
cox.zph(m2)
```

