---
title: "Olink_2_cohorts_2"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Olink_2_cohorts_2"

# Load libraries
library(EnhancedVolcano)
library(here)
library(enrichplot)
library(readr)
library(readxl)
library(Matrix)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(pheatmap)
library(png)
library(gridExtra)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(qs2)
library(statmod)
library(rtracklayer)
library(png)
library(reactable)
library(UpSetR)
library(limma)
library(igraph)
library(gridExtra) 
library(ggpubr)
library(DESeq2)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(UpSetR)
library(ComplexHeatmap)
library(survival)
library(broom)
library(circlize)
library(variancePartition)

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

## Import data

Olink data Zurich and Poland

```{r read data}
data_z <- qs_read(here::here("data", "o37198_2025-02-27_olink_SummarizedExperiment.qs2"))

clin_z <- read_excel(here::here("data", "Metadata_Olink_analysis.xlsx"))

meta_z <- colData(data_z) %>% data.frame()
isOutlierColumn <- meta_z$Plate == "P0194" & meta_z$Col %in% c("3", "4")
table(isOutlierColumn)
use <- meta_z$SampleType %in% c("Serum") & meta_z$Group %in% c("SSc-PAH", "SSc-noPAH") & !isOutlierColumn
table(use)
data_z <- data_z[ , use]
meta_z <- meta_z[use, ]
rownames(meta_z) <- gsub("_P0194", "", rownames(meta_z))
meta_z$Barcode <- rownames(meta_z)
rownames(data_z) <- data_z@elementMetadata@listData[["Gene"]]
rownames(data_z@assays@data@listData[["npx"]]) <- data_z@elementMetadata@listData[["Gene"]]
data_z <- assay(data_z, "npx")
data_z <- as.data.frame(data_z[-c(1:3), ])
colnames(data_z) <- rownames(meta_z)

clin_z <- clin_z %>%
  dplyr::filter(Barcode %in% colnames(data_z))
dim(data_z)
table(clin_z$Group)


data_p <- read.delim(here::here("data", "OLINK_PL", "quant.xls"), sep = "\t", check.names = FALSE)

clin_p <- read_excel(here::here("data", "OLINK_PL", "Metadata_Olink_PL.xlsx"))

rownames(data_p) <- sub("_.*", "", data_p$SampleID)
data_p <- data_p[, -1]
keep <- intersect(colnames(data_p), clin_p$Barcode)
data_p <- data_p[, keep, drop = FALSE]
dim(data_p)
table(clin_p$Group)
```

```{r, fig.height=8, fig.width=10.8}
barcodes <- colnames(data_z)

grp <- clin_z$Group[match(barcodes, clin_z$Barcode)]
cmp <- clin_z$COMPERA_4strata[match(barcodes, clin_z$Barcode)]
cmp <- factor(cmp, levels = c("low","intermediate-low","intermediate-high","high"))
sex <- clin_z$Sex[match(barcodes, clin_z$Barcode)]

# --- colors ---
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "low"               = "#9CA3AF",   # grey
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)
cols_sex <- c("F"="green4", "M"="yellow2")

# --- z-score per protein (row) ---
matz <- t(scale(t(data_z)))
matz[is.nan(matz)] <- 0

# heatmap colors
hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

# column annotations (COMPERA NAs shown as light grey)
ha <- HeatmapAnnotation(
  Group   = grp,
  COMPERA = cmp,
  Sex = sex,
  col = list(Group = cols_group, COMPERA = cols_compera, Sex = cols_sex),
  na_col = "#E5E7EB",
  annotation_legend_param = list(
    Group   = list(title = "Group"),
    COMPERA = list(title = "Risk (COMPERA)"),
    Sex   = list(title = "Sex")
  )
)

# --- draw (all proteins, no row names, default clustering, no group split) ---
Heatmap(
  matz,
  name = "Z-score",
  col = hm_cols,
  top_annotation = ha,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 9),
  row_title = NULL,
  column_title = NULL
) |>
  draw(heatmap_legend_side = "right", annotation_legend_side = "right")
```

```{r, fig.height=8, fig.width=10.8}
barcodes <- colnames(data_p)

grp <- clin_p$Group[match(barcodes, clin_p$Barcode)]
cmp <- clin_p$Compera_4strata[match(barcodes, clin_p$Barcode)]
cmp <- factor(cmp, levels = c("Low","Intermediate-low","Intermediate-high","High"))
sex <- clin_p$Sex[match(barcodes, clin_p$Barcode)]

# --- colors ---
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "Low"               = "#9CA3AF",   # grey
  "Intermediate-low"  = "#DDB7B2",
  "Intermediate-high" = "#C7685F",
  "High"              = "#A63A2A"
)
cols_sex <- c("F"="green4", "M"="yellow2")

# --- z-score per protein (row) ---
matz <- t(scale(t(data_p)))
matz[is.nan(matz)] <- 0

# heatmap colors
hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

# column annotations (COMPERA NAs shown as light grey)
ha <- HeatmapAnnotation(
  Group   = grp,
  COMPERA = cmp,
  Sex = sex,
  col = list(Group = cols_group, COMPERA = cols_compera, Sex = cols_sex),
  na_col = "#E5E7EB",
  annotation_legend_param = list(
    Group   = list(title = "Group"),
    COMPERA = list(title = "Risk (COMPERA)"),
    Sex   = list(title = "Sex")
  )
)

# --- draw (all proteins, no row names, default clustering, no group split) ---
Heatmap(
  matz,
  name = "Z-score",
  col = hm_cols,
  top_annotation = ha,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 9),
  row_title = NULL,
  column_title = NULL
) |>
  draw(heatmap_legend_side = "right", annotation_legend_side = "right")
```

```{r heatmap_z_top100var, fig.height=10.8, fig.width=10.8}
barcodes <- colnames(data_z)

grp <- clin_z$Group[match(barcodes, clin_z$Barcode)]
cmp <- clin_z$COMPERA_4strata[match(barcodes, clin_z$Barcode)]
cmp <- factor(cmp, levels = c("low","intermediate-low","intermediate-high","high"))
sex <- clin_z$Sex[match(barcodes, clin_z$Barcode)]

# --- colors ---
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "low"               = "#9CA3AF",
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)
cols_sex <- c("F"="green4", "M"="yellow2")

# --- pick top 100 variable proteins (rows) ---
vars <- apply(data_z, 1, var, na.rm = TRUE)
top100 <- names(sort(vars, decreasing = TRUE))[1:100]
mat_top <- as.matrix(data_z[top100, , drop = FALSE])

# --- z-score per protein (row) ---
matz <- t(scale(t(mat_top)))
matz[is.nan(matz)] <- 0

hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

ha <- HeatmapAnnotation(
  Group   = grp,
  COMPERA = cmp,
  Sex     = sex,
  col = list(Group = cols_group, COMPERA = cols_compera, Sex = cols_sex),
  na_col = "#E5E7EB",
  annotation_legend_param = list(
    Group   = list(title = "Group"),
    COMPERA = list(title = "Risk (COMPERA)"),
    Sex     = list(title = "Sex")
  )
)

Heatmap(
  matz,
  name = "Z-score",
  col = hm_cols,
  top_annotation = ha,
  show_row_names = TRUE,              
  row_names_gp = gpar(fontsize = 7),
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 8.5),
  row_title = "Top 100 variable proteins",
  column_title = NULL
) |>
  draw(heatmap_legend_side = "right", annotation_legend_side = "right")
```

```{r heatmap_p_top100var, fig.height=10.88, fig.width=10.8}
stopifnot(exists("data_p"), exists("clin_p"))

barcodes <- colnames(data_p)

grp <- clin_p$Group[match(barcodes, clin_p$Barcode)]
cmp <- clin_p$Compera_4strata[match(barcodes, clin_p$Barcode)]
cmp <- factor(cmp, levels = c("Low","Intermediate-low","Intermediate-high","High"))
sex <- clin_p$Sex[match(barcodes, clin_p$Barcode)]

# --- colors ---
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "Low"               = "#9CA3AF",
  "Intermediate-low"  = "#DDB7B2",
  "Intermediate-high" = "#C7685F",
  "High"              = "#A63A2A"
)
cols_sex <- c("F"="green4", "M"="yellow2")

# --- pick top 100 variable proteins (rows) ---
vars <- apply(data_p, 1, var, na.rm = TRUE)
top100 <- names(sort(vars, decreasing = TRUE))[1:100]
mat_top <- as.matrix(data_p[top100, , drop = FALSE])

# --- z-score per protein (row) ---
matz <- t(scale(t(mat_top)))
matz[is.nan(matz)] <- 0

hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

ha <- HeatmapAnnotation(
  Group   = grp,
  COMPERA = cmp,
  Sex     = sex,
  col = list(Group = cols_group, COMPERA = cols_compera, Sex = cols_sex),
  na_col = "#E5E7EB",
  annotation_legend_param = list(
    Group   = list(title = "Group"),
    COMPERA = list(title = "Risk (COMPERA)"),
    Sex     = list(title = "Sex")
  )
)

Heatmap(
  matz,
  name = "Z-score",
  col = hm_cols,
  top_annotation = ha,
  show_row_names = TRUE,              # set FALSE if you want to hide protein names
  row_names_gp = gpar(fontsize = 7),
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 8.5),
  row_title = "Top 100 variable proteins",
  column_title = NULL
) |>
  draw(heatmap_legend_side = "right", annotation_legend_side = "right")
```
```{r pca_z_group, fig.height=3.5, fig.width=5.4}
# PCA (samples as rows)
pca <- prcomp(t(as.matrix(data_z)), scale. = TRUE)
pvar <- (pca$sdev^2) / sum(pca$sdev^2) * 100

df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  Group = clin_z$Group[match(rownames(pca$x), clin_z$Barcode)]
)

cols_group <- c("SSc-noPAH" = "#004C8C", "SSc-PAH" = "#A63A2A")

ggplot(df, aes(PC1, PC2, color = Group)) +
  stat_ellipse(aes(fill = Group), type = "norm", level = 0.95,
               geom = "polygon", alpha = 0.10, color = NA) +
  stat_ellipse(type = "norm", level = 0.95, linewidth = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = cols_group) +
  scale_fill_manual(values = cols_group) +
  labs(
    x = paste0("PC1 (", round(pvar[1], 1), "%)"),
    y = paste0("PC2 (", round(pvar[2], 1), "%)")
  ) +
  theme_minimal()
```

```{r pca_z_compera, fig.height=3.5, fig.width=5.4}
# --- subset to SSc-PAH only ---
keep <- clin_z$Group == "SSc-PAH"
barcodes_keep <- clin_z$Barcode[keep]

data_sub <- data_z[, colnames(data_z) %in% barcodes_keep, drop = FALSE]
clin_sub <- clin_z[match(colnames(data_sub), clin_z$Barcode), ]

# PCA
pca <- prcomp(t(as.matrix(data_sub)), scale. = TRUE)
pvar <- (pca$sdev^2) / sum(pca$sdev^2) * 100

df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  COMPERA = factor(
    clin_sub$COMPERA_4strata,
    levels = c("low","intermediate-low","intermediate-high","high")
  )
)

cols_compera_z <- c(
  "low"               = "#9CA3AF",
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)

ggplot(df, aes(PC1, PC2, color = COMPERA)) +
  stat_ellipse(aes(fill = COMPERA), type = "norm", level = 0.90,
               geom = "polygon", alpha = 0.10, color = NA) +
  stat_ellipse(type = "norm", level = 0.90, linewidth = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = cols_compera_z, name = "Risk (COMPERA)") +
  scale_fill_manual(values = cols_compera_z,  name = "Risk (COMPERA)") +
  labs(
    x = paste0("PC1 (", round(pvar[1], 1), "%)"),
    y = paste0("PC2 (", round(pvar[2], 1), "%)")
  ) +
  theme_minimal() +
  guides(fill = "none")
```

```{r pca_p_group, fig.height=3.5, fig.width=5.4}
pca <- prcomp(t(as.matrix(data_p)), scale. = TRUE)
pvar <- (pca$sdev^2) / sum(pca$sdev^2) * 100

df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  Group = clin_p$Group[match(rownames(pca$x), clin_p$Barcode)]
)

cols_group <- c("SSc-noPAH" = "#004C8C", "SSc-PAH" = "#A63A2A")

ggplot(df, aes(PC1, PC2, color = Group)) +
  stat_ellipse(aes(fill = Group), type = "norm", level = 0.95,
               geom = "polygon", alpha = 0.10, color = NA) +
  stat_ellipse(type = "norm", level = 0.95, linewidth = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = cols_group) +
  scale_fill_manual(values = cols_group) +
  labs(
    x = paste0("PC1 (", round(pvar[1], 1), "%)"),
    y = paste0("PC2 (", round(pvar[2], 1), "%)")
  ) +
  theme_minimal()
```

```{r pca_p_compera, fig.height=3.5, fig.width=5.4}
# --- subset to SSc-PAH only ---
keep <- clin_p$Group == "SSc-PAH"
barcodes_keep <- clin_p$Barcode[keep]

data_sub <- data_p[, colnames(data_p) %in% barcodes_keep, drop = FALSE]
clin_sub <- clin_p[match(colnames(data_sub), clin_p$Barcode), ]

# PCA
pca <- prcomp(t(as.matrix(data_sub)), scale. = TRUE)
pvar <- (pca$sdev^2) / sum(pca$sdev^2) * 100

df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  COMPERA = factor(
    clin_sub$Compera_4strata,
    levels = c("Low","Intermediate-low","Intermediate-high","High")
  )
)

cols_compera_p <- c(
  "Low"               = "#9CA3AF",
  "Intermediate-low"  = "#DDB7B2",
  "Intermediate-high" = "#C7685F",
  "High"              = "#A63A2A"
)

ggplot(df, aes(PC1, PC2, color = COMPERA)) +
  stat_ellipse(aes(fill = COMPERA), type = "norm", level = 0.90,
               geom = "polygon", alpha = 0.10, color = NA) +
  stat_ellipse(type = "norm", level = 0.90, linewidth = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = cols_compera_p, name = "Risk (COMPERA)") +
  scale_fill_manual(values = cols_compera_p,  name = "Risk (COMPERA)") +
  labs(
    x = paste0("PC1 (", round(pvar[1], 1), "%)"),
    y = paste0("PC2 (", round(pvar[2], 1), "%)")
  ) +
  theme_minimal() +
  guides(fill = "none")
```

## Variance partitioning

```{r vp_z_all, fig.height=4.5, fig.width=6.5}
# expression: proteins x samples
expr <- as.matrix(data_z)

# metadata aligned to columns of expr
meta <- clin_z |>
  dplyr::distinct(Barcode, .keep_all = TRUE) |>
  dplyr::filter(Barcode %in% colnames(expr)) |>
  tibble::column_to_rownames("Barcode")

expr <- expr[, rownames(meta), drop = FALSE]

# variables for this model
vars <- c("Group", "Age", "Sex", "Sample_age_days")
meta <- meta |>
  dplyr::mutate(
    Group = factor(Group),
    Sex   = factor(Sex)
  )

# keep complete cases for model variables
n_before <- nrow(meta)

meta <- meta[complete.cases(meta[, vars]), , drop = FALSE]
expr <- expr[, rownames(meta), drop = FALSE]

n_after <- nrow(meta)
message("Zurich (all): ", n_after, " / ", n_before, " samples used")

form <- ~ Group + Age + Sex + Sample_age_days

vp <- fitExtractVarPartModel(expr, form, meta)
colnames(vp) <- gsub("`", "", colnames(vp))

# quick summaries
vp_mean <- sort(colMeans(vp), decreasing = TRUE)
df <- data.frame(variable = names(vp_mean), value = vp_mean)

ggplot(df, aes(x = reorder(variable, -value), y = value)) +
  geom_col() +
  labs(x = NULL, y = "Mean fraction of variance explained",
       title = "Zurich: all samples") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotVarPart(vp) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vp_z_pah, fig.height=4.5, fig.width=6.5}
expr <- as.matrix(data_z)

meta <- clin_z |>
  dplyr::distinct(Barcode, .keep_all = TRUE) |>
  dplyr::filter(Barcode %in% colnames(expr)) |>
  dplyr::filter(Group == "SSc-PAH") |>
  tibble::column_to_rownames("Barcode")

expr <- expr[, rownames(meta), drop = FALSE]

# requested covariates
vars <- c("TAPSE_sPAP", "Age", "Sex")
meta <- meta |>
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    Age = as.numeric(Age),
    Sex = factor(Sex)
  )

n_before <- nrow(meta)

meta <- meta[complete.cases(meta[, vars]), , drop = FALSE]
expr <- expr[, rownames(meta), drop = FALSE]

n_after <- nrow(meta)
message("SSc-PAH only: ", n_after, " / ", n_before, " samples used")

form <- ~ TAPSE_sPAP + Age + Sex 

vp <- fitExtractVarPartModel(expr, form, meta)
colnames(vp) <- gsub("`", "", colnames(vp))

vp_mean <- sort(colMeans(vp), decreasing = TRUE)
df <- data.frame(variable = names(vp_mean), value = vp_mean)

ggplot(df, aes(x = reorder(variable, -value), y = value)) +
  geom_col() +
  labs(x = NULL, y = "Mean fraction of variance explained",
       title = "Zurich: SSc-PAH only") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotVarPart(vp) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vp_p_all, fig.height=4.5, fig.width=6.5}
library(variancePartition)

expr <- as.matrix(data_p)

meta <- clin_p |>
  dplyr::distinct(Barcode, .keep_all = TRUE) |>
  dplyr::filter(Barcode %in% colnames(expr)) |>
  tibble::column_to_rownames("Barcode")

expr <- expr[, rownames(meta), drop = FALSE]

# requested covariates
vars <- c("Group", "Age", "Sex", "Sample_age")
meta <- meta |>
  dplyr::mutate(
    Group  = factor(Group),
    Sex    = factor(Sex),
  )

n_before <- nrow(meta)

meta <- meta[complete.cases(meta[, vars]), , drop = FALSE]
expr <- expr[, rownames(meta), drop = FALSE]

n_after <- nrow(meta)
message("Poland (all): ", n_after, " / ", n_before, " samples used")


form <- ~ Group + Age + Sex + Sample_age

vp <- fitExtractVarPartModel(expr, form, meta)
colnames(vp) <- gsub("`", "", colnames(vp))

vp_mean <- sort(colMeans(vp), decreasing = TRUE)
df <- data.frame(variable = names(vp_mean), value = vp_mean)

ggplot(df, aes(x = reorder(variable, -value), y = value)) +
  geom_col() +
  labs(x = NULL, y = "Mean fraction of variance explained",
       title = "Poland: all samples") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotVarPart(vp) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vp_p_pah, fig.height=4.5, fig.width=6.5}
library(variancePartition)

expr <- as.matrix(data_p)

meta <- clin_p |>
  dplyr::distinct(Barcode, .keep_all = TRUE) |>
  dplyr::filter(Barcode %in% colnames(expr)) |>
  dplyr::filter(Group == "SSc-PAH") |>
  tibble::column_to_rownames("Barcode")

expr <- expr[, rownames(meta), drop = FALSE]

# requested covariates
vars <- c("TAPSE_sPAP", "Age", "Sex")
meta <- meta |>
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    Age = as.numeric(Age),
    Sex = factor(Sex)
  )

n_before <- nrow(meta)

meta <- meta[complete.cases(meta[, vars]), , drop = FALSE]
expr <- expr[, rownames(meta), drop = FALSE]

n_after <- nrow(meta)
message("SSc-PAH only: ", n_after, " / ", n_before, " samples used")

form <- ~ TAPSE_sPAP + Age + Sex

vp <- fitExtractVarPartModel(expr, form, meta)
colnames(vp) <- gsub("`", "", colnames(vp))

vp_mean <- sort(colMeans(vp), decreasing = TRUE)
df <- data.frame(variable = names(vp_mean), value = vp_mean)

ggplot(df, aes(x = reorder(variable, -value), y = value)) +
  geom_col() +
  labs(x = NULL, y = "Mean fraction of variance explained",
       title = "Poland: SSc-PAH only") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plotVarPart(vp) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



## Differential expression analysis

```{r de_group_zurich}
# --- prepare metadata aligned to expression ---
meta_z_de <- clin_z %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::mutate(
    Group = factor(Group, levels = c("SSc-noPAH", "SSc-PAH")),
    Sex   = factor(Sex),
    Age   = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

# keep only complete cases for model variables
meta_z_de <- meta_z_de[complete.cases(meta_z_de[, c("Group", "Age", "Sex")]), , drop = FALSE]

expr_z_de <- as.matrix(data_z[, rownames(meta_z_de), drop = FALSE])

# --- design: Group + Age + Sex ---
design_z <- model.matrix(~ Group + Age + Sex, data = meta_z_de)

# rescale numeric covariate(s) like in your code
if ("Age" %in% colnames(design_z)) design_z[, "Age"] <- rescale(design_z[, "Age"])

# --- fit limma ---
fit_z <- lmFit(expr_z_de, design_z)
fit_z <- eBayes(fit_z)

# coefficient name for GroupSSc-PAH (vs reference SSc-noPAH)
coef_group_z <- grep("^Group", colnames(design_z), value = TRUE)[1]

DE_z <- topTable(fit_z, coef = coef_group_z, number = Inf, adjust.method = "BH")
DE_z$Gene <- rownames(DE_z)

DE_z_signif <- DE_z %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

# write outputs (same style as your loop, but single comparison)
write.csv(DE_z,
          file = here::here(output_dir_data, "DE_Results_Group_SScPAH_vs_noPAH_Zurich.csv"),
          quote = FALSE, row.names = FALSE)
write.csv(DE_z_signif,
          file = here::here(output_dir_data, "DE_Significant_Group_SScPAH_vs_noPAH_Zurich.csv"),
          quote = FALSE, row.names = FALSE)

# store for volcano chunk
n_z <- ncol(expr_z_de)
```

```{r de_group_poland}
meta_p_de <- clin_p %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::mutate(
    Group = factor(Group, levels = c("SSc-noPAH", "SSc-PAH")),
    Sex   = factor(Sex),
    Age   = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_p_de <- meta_p_de[complete.cases(meta_p_de[, c("Group", "Age", "Sex")]), , drop = FALSE]

expr_p_de <- as.matrix(data_p[, rownames(meta_p_de), drop = FALSE])

design_p <- model.matrix(~ Group + Age + Sex, data = meta_p_de)
if ("Age" %in% colnames(design_p)) design_p[, "Age"] <- rescale(design_p[, "Age"])

fit_p <- lmFit(expr_p_de, design_p)
fit_p <- eBayes(fit_p)

coef_group_p <- grep("^Group", colnames(design_p), value = TRUE)[1]

DE_p <- topTable(fit_p, coef = coef_group_p, number = Inf, adjust.method = "BH")
DE_p$Gene <- rownames(DE_p)

DE_p_signif <- DE_p %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

write.csv(DE_p,
          file = here::here(output_dir_data, "DE_Results_Group_SScPAH_vs_noPAH_Poland.csv"),
          quote = FALSE, row.names = FALSE)
write.csv(DE_p_signif,
          file = here::here(output_dir_data, "DE_Significant_Group_SScPAH_vs_noPAH_Poland.csv"),
          quote = FALSE, row.names = FALSE)

n_p <- ncol(expr_p_de)
```

```{r, fig.height=4, fig.width=3.5, warning=FALSE}
keyvals <- rep("grey", nrow(DE_z))
names(keyvals) <- rep("Not significant", nrow(DE_z))

keyvals[DE_z$logFC > 0.5 & DE_z$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Increased"

keyvals[DE_z$logFC < -0.5 & DE_z$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

volcano_z <- EnhancedVolcano(
  DE_z,
  lab = rownames(DE_z),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Zurich cohort\nSSc-PAH vs SSc-noPAH\nn = ", n_z)
) +
  theme(legend.position = "none")

volcano_z
```

```{r, fig.height=4, fig.width=3.5, warning=FALSE}
keyvals <- rep("grey", nrow(DE_p))
names(keyvals) <- rep("Not significant", nrow(DE_p))

keyvals[DE_p$logFC > 0.5 & DE_p$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Increased"

keyvals[DE_p$logFC < -0.5 & DE_p$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

volcano_p <- EnhancedVolcano(
  DE_p,
  lab = rownames(DE_p),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Poland cohort\nSSc-PAH vs SSc-noPAH\nn = ", n_p)
) +
  theme(legend.position = "none")

volcano_p
```

```{r de_tapse_zurich_pah}
meta_z_de <- clin_z %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    Sex = factor(Sex),
    Age = as.numeric(Age),
    TAPSE_sPAP = as.numeric(TAPSE_sPAP)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_z_de <- meta_z_de[
  complete.cases(meta_z_de[, c("TAPSE_sPAP", "Age", "Sex")]),
  , drop = FALSE
]

expr_z_de <- as.matrix(data_z[, rownames(meta_z_de), drop = FALSE])

design_z <- model.matrix(~ TAPSE_sPAP + Age + Sex, data = meta_z_de)

design_z[, "TAPSE_sPAP"] <- rescale(design_z[, "TAPSE_sPAP"])
design_z[, "Age"]        <- rescale(design_z[, "Age"])

fit_z <- lmFit(expr_z_de, design_z)
fit_z <- eBayes(fit_z)

DE_z_TAPSE <- topTable(
  fit_z,
  coef = "TAPSE_sPAP",
  number = Inf,
  adjust.method = "BH"
)

DE_z_TAPSE$Gene <- rownames(DE_z_TAPSE)

DE_z_TAPSE_signif <- DE_z_TAPSE %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

write.csv(
  DE_z_TAPSE,
  file = here::here(output_dir_data, "DE_TAPSE_sPAP_SScPAH_Zurich.csv"),
  quote = FALSE, row.names = FALSE
)

write.csv(
  DE_z_TAPSE_signif,
  file = here::here(output_dir_data, "DE_TAPSE_sPAP_SScPAH_Zurich_significant.csv"),
  quote = FALSE, row.names = FALSE
)

n_z_TAPSE <- ncol(expr_z_de)
```

```{r de_tapse_poland_pah}
meta_p_de <- clin_p %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    Sex = factor(Sex),
    Age = as.numeric(Age),
    TAPSE_sPAP = as.numeric(TAPSE_sPAP)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_p_de <- meta_p_de[
  complete.cases(meta_p_de[, c("TAPSE_sPAP", "Age", "Sex")]),
  , drop = FALSE
]

expr_p_de <- as.matrix(data_p[, rownames(meta_p_de), drop = FALSE])

design_p <- model.matrix(~ TAPSE_sPAP + Age + Sex, data = meta_p_de)

design_p[, "TAPSE_sPAP"] <- rescale(design_p[, "TAPSE_sPAP"])
design_p[, "Age"]        <- rescale(design_p[, "Age"])

fit_p <- lmFit(expr_p_de, design_p)
fit_p <- eBayes(fit_p)

DE_p_TAPSE <- topTable(
  fit_p,
  coef = "TAPSE_sPAP",
  number = Inf,
  adjust.method = "BH"
)

DE_p_TAPSE$Gene <- rownames(DE_p_TAPSE)

DE_p_TAPSE_signif <- DE_p_TAPSE %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

write.csv(
  DE_p_TAPSE,
  file = here::here(output_dir_data, "DE_TAPSE_sPAP_SScPAH_Poland.csv"),
  quote = FALSE, row.names = FALSE
)

write.csv(
  DE_p_TAPSE_signif,
  file = here::here(output_dir_data, "DE_TAPSE_sPAP_SScPAH_Poland_significant.csv"),
  quote = FALSE, row.names = FALSE
)

n_p_TAPSE <- ncol(expr_p_de)
```

```{r, fig.height=4, fig.width=3.5, warning=FALSE}
keyvals <- rep("grey", nrow(DE_z_TAPSE))
names(keyvals) <- rep("Not significant", nrow(DE_z_TAPSE))

keyvals[DE_z_TAPSE$logFC > 0.5 & DE_z_TAPSE$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Increased"

keyvals[DE_z_TAPSE$logFC < -0.5 & DE_z_TAPSE$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

volcano_z_TAPSE <- EnhancedVolcano(
  DE_z_TAPSE,
  lab = rownames(DE_z_TAPSE),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Zurich cohort SSc-PAH\nTAPSE_sPAP\nn = ", n_z_TAPSE)
) +
  theme(legend.position = "none")

volcano_z_TAPSE
```

```{r, fig.height=4, fig.width=3.5, warning=FALSE}
keyvals <- rep("grey", nrow(DE_p_TAPSE))
names(keyvals) <- rep("Not significant", nrow(DE_p_TAPSE))

keyvals[DE_p_TAPSE$logFC > 0.5 & DE_p_TAPSE$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Increased"

keyvals[DE_p_TAPSE$logFC < -0.5 & DE_p_TAPSE$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

volcano_p_TAPSE <- EnhancedVolcano(
  DE_p_TAPSE,
  lab = rownames(DE_p_TAPSE),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Poland cohort SSc-PAH\nTAPSE_sPAP\nn = ", n_p_TAPSE)
) +
  theme(legend.position = "none")

volcano_p_TAPSE
```

```{r}
meta_z_bin <- clin_z %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = ifelse(TAPSE_sPAP < 0.32, "Low", "High"),
    TAPSE_group = factor(TAPSE_group, levels = c("High", "Low")),
    Sex = factor(Sex),
    Age = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_z_bin <- meta_z_bin[
  complete.cases(meta_z_bin[, c("TAPSE_group", "Age", "Sex")]),
  , drop = FALSE
]

expr_z_bin <- as.matrix(data_z[, rownames(meta_z_bin), drop = FALSE])

design_z_bin <- model.matrix(~ TAPSE_group + Age + Sex, data = meta_z_bin)
design_z_bin[, "Age"] <- rescale(design_z_bin[, "Age"])

fit_z_bin <- lmFit(expr_z_bin, design_z_bin)
fit_z_bin <- eBayes(fit_z_bin)

coef_bin_z <- grep("^TAPSE_group", colnames(design_z_bin), value = TRUE)[1]

DE_z_bin <- topTable(fit_z_bin, coef = coef_bin_z, number = Inf, adjust.method = "BH")
DE_z_bin$Gene <- rownames(DE_z_bin)

DE_z_bin_signif <- DE_z_bin %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

n_z_bin <- ncol(expr_z_bin)
```

```{r, fig.height=4, fig.width=3.5, warning=FALSE}
keyvals <- rep("grey", nrow(DE_z_bin))
names(keyvals) <- rep("Not significant", nrow(DE_z_bin))

keyvals[DE_z_bin$logFC > 0.5 & DE_z_bin$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Higher in Low TAPSE_sPAP"

keyvals[DE_z_bin$logFC < -0.5 & DE_z_bin$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Higher in High TAPSE_sPAP"

volcano_z_bin <- EnhancedVolcano(
  DE_z_bin,
  lab = rownames(DE_z_bin),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Zurich cohort SSc-PAH\nTAPSE/sPAP<0.32 vs ≥0.32\nn = ", n_z_bin)
) +
  theme(legend.position = "none")

volcano_z_bin
```

```{r}
meta_p_bin <- clin_p %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = ifelse(TAPSE_sPAP < 0.32, "Low", "High"),
    TAPSE_group = factor(TAPSE_group, levels = c("High", "Low")),
    Sex = factor(Sex),
    Age = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_p_bin <- meta_p_bin[
  complete.cases(meta_p_bin[, c("TAPSE_group", "Age", "Sex")]),
  , drop = FALSE
]

expr_p_bin <- as.matrix(data_p[, rownames(meta_p_bin), drop = FALSE])

design_p_bin <- model.matrix(~ TAPSE_group + Age + Sex, data = meta_p_bin)
design_p_bin[, "Age"] <- rescale(design_p_bin[, "Age"])

fit_p_bin <- lmFit(expr_p_bin, design_p_bin)
fit_p_bin <- eBayes(fit_p_bin)

coef_bin_p <- grep("^TAPSE_group", colnames(design_p_bin), value = TRUE)[1]

DE_p_bin <- topTable(fit_p_bin, coef = coef_bin_p, number = Inf, adjust.method = "BH")
DE_p_bin$Gene <- rownames(DE_p_bin)

DE_p_bin_signif <- DE_p_bin %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

n_p_bin <- ncol(expr_p_bin)
```

```{r, fig.height=4, fig.width=3.5, warning=FALSE}
keyvals <- rep("grey", nrow(DE_p_bin))
names(keyvals) <- rep("Not significant", nrow(DE_p_bin))

keyvals[DE_p_bin$logFC > 0.5 & DE_p_bin$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Higher in Low TAPSE_sPAP"

keyvals[DE_p_bin$logFC < -0.5 & DE_p_bin$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Higher in High TAPSE_sPAP"

volcano_p_bin <- EnhancedVolcano(
  DE_p_bin,
  lab = rownames(DE_p_bin),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Poland cohort SSc-PAH\nTAPSE/sPAP<0.32 vs ≥0.32\nn = ", n_p_bin)
) +
  theme(legend.position = "none")

volcano_p_bin
```

```{r}
meta_z_comp <- clin_z %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    COMPERA_number = as.numeric(COMPERA_number),
    Sex = factor(Sex),
    Age = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_z_comp <- meta_z_comp[
  complete.cases(meta_z_comp[, c("COMPERA_number", "Age", "Sex")]),
  , drop = FALSE
]

expr_z_comp <- as.matrix(data_z[, rownames(meta_z_comp), drop = FALSE])

design_z_comp <- model.matrix(~ COMPERA_number + Age + Sex, data = meta_z_comp)

design_z_comp[, "COMPERA_number"] <- rescale(design_z_comp[, "COMPERA_number"])
design_z_comp[, "Age"]            <- rescale(design_z_comp[, "Age"])

fit_z_comp <- lmFit(expr_z_comp, design_z_comp)
fit_z_comp <- eBayes(fit_z_comp)

DE_z_comp <- topTable(
  fit_z_comp,
  coef = "COMPERA_number",
  number = Inf,
  adjust.method = "BH"
)

DE_z_comp$Gene <- rownames(DE_z_comp)

DE_z_comp_signif <- DE_z_comp %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

n_z_comp <- ncol(expr_z_comp)
```

```{r, fig.height=4, fig.width=3.5, warning=FALSE}
keyvals <- rep("grey", nrow(DE_z_comp))
names(keyvals) <- rep("Not significant", nrow(DE_z_comp))

keyvals[DE_z_comp$logFC > 0.5 & DE_z_comp$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Positive association"

keyvals[DE_z_comp$logFC < -0.5 & DE_z_comp$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Negative association"

volcano_z_comp <- EnhancedVolcano(
  DE_z_comp,
  lab = rownames(DE_z_comp),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Zurich cohort SSc-PAH\nCOMPERA_number\nn = ", n_z_comp)
) +
  theme(legend.position = "none")

volcano_z_comp
```

```{r}
meta_p_comp <- clin_p %>%
  dplyr::distinct(Barcode, .keep_all = TRUE) %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::mutate(
    Compera_number = as.numeric(Compera_number),
    Sex = factor(Sex),
    Age = as.numeric(Age)
  ) %>%
  tibble::column_to_rownames("Barcode")

meta_p_comp <- meta_p_comp[
  complete.cases(meta_p_comp[, c("Compera_number", "Age", "Sex")]),
  , drop = FALSE
]

expr_p_comp <- as.matrix(data_p[, rownames(meta_p_comp), drop = FALSE])

design_p_comp <- model.matrix(~ Compera_number + Age + Sex, data = meta_p_comp)

design_p_comp[, "Compera_number"] <- rescale(design_p_comp[, "Compera_number"])
design_p_comp[, "Age"]            <- rescale(design_p_comp[, "Age"])

fit_p_comp <- lmFit(expr_p_comp, design_p_comp)
fit_p_comp <- eBayes(fit_p_comp)

DE_p_comp <- topTable(
  fit_p_comp,
  coef = "Compera_number",
  number = Inf,
  adjust.method = "BH"
)

DE_p_comp$Gene <- rownames(DE_p_comp)

DE_p_comp_signif <- DE_p_comp %>%
  dplyr::filter(adj.P.Val < 0.1, abs(logFC) > 0.5) %>%
  dplyr::arrange(P.Value)

n_p_comp <- ncol(expr_p_comp)
```

```{r, fig.height=4, fig.width=3.5, warning=FALSE}
keyvals <- rep("grey", nrow(DE_p_comp))
names(keyvals) <- rep("Not significant", nrow(DE_p_comp))

keyvals[DE_p_comp$logFC > 0.5 & DE_p_comp$adj.P.Val < 0.1] <- "#A63A2A"
names(keyvals)[keyvals == "#A63A2A"] <- "Positive association"

keyvals[DE_p_comp$logFC < -0.5 & DE_p_comp$adj.P.Val < 0.1] <- "#004C8C"
names(keyvals)[keyvals == "#004C8C"] <- "Negative association"

volcano_p_comp <- EnhancedVolcano(
  DE_p_comp,
  lab = rownames(DE_p_comp),
  x = "logFC",
  y = "adj.P.Val",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 1,
  FCcutoff = 0.5,
  pCutoff  = 0.1,
  ylim = c(0, 4),
  colCustom = keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  subtitle = NULL,
  title = paste0("Poland cohort\nSSc-PAH only\nCOMPERA_number\nn = ", n_p_comp)
) +
  theme(legend.position = "none")

volcano_p_comp
```

```{r cor_group, fig.height=4, fig.width=4}
df_group <- dplyr::inner_join(
  DE_z[, c("Gene", "logFC")],
  DE_p[, c("Gene", "logFC")],
  by = "Gene",
  suffix = c("_z", "_p")
)

cor_group <- cor(df_group$logFC_z, df_group$logFC_p)

ggplot(df_group, aes(logFC_z, logFC_p)) +
  geom_point(size = 1, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  labs(
    x = "Zurich logFC",
    y = "Poland logFC",
    title = paste0("SSc-PAH vs SSc-noPAH\nr = ", round(cor_group, 2))
  ) +
  theme_classic()
```

```{r cor_tapse_cont, fig.height=4, fig.width=4}
df_tapse <- dplyr::inner_join(
  DE_z_TAPSE[, c("Gene", "logFC")],
  DE_p_TAPSE[, c("Gene", "logFC")],
  by = "Gene",
  suffix = c("_z", "_p")
)

cor_tapse <- cor(df_tapse$logFC_z, df_tapse$logFC_p)

ggplot(df_tapse, aes(logFC_z, logFC_p)) +
  geom_point(size = 1, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  labs(
    x = "Zurich logFC",
    y = "Poland logFC",
    title = paste0("TAPSE_sPAP (continuous)\nr = ", round(cor_tapse, 2))
  ) +
  theme_classic()
```

```{r cor_tapse_binary, fig.height=4, fig.width=4}
df_tapse_bin <- dplyr::inner_join(
  DE_z_bin[, c("Gene", "logFC")],
  DE_p_bin[, c("Gene", "logFC")],
  by = "Gene",
  suffix = c("_z", "_p")
)

cor_tapse_bin <- cor(df_tapse_bin$logFC_z, df_tapse_bin$logFC_p)

ggplot(df_tapse_bin, aes(logFC_z, logFC_p)) +
  geom_point(size = 1, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  labs(
    x = "Zurich logFC",
    y = "Poland logFC",
    title = paste0("TAPSE/sPAP groups (<0.32 vs ≥0.32)\nr = ", round(cor_tapse_bin, 2))
  ) +
  theme_classic()
```

```{r cor_compera, fig.height=4, fig.width=4}
df_comp <- dplyr::inner_join(
  DE_z_comp[, c("Gene", "logFC")],
  DE_p_comp[, c("Gene", "logFC")],
  by = "Gene",
  suffix = c("_z", "_p")
)

cor_comp <- cor(df_comp$logFC_z, df_comp$logFC_p)

ggplot(df_comp, aes(logFC_z, logFC_p)) +
  geom_point(size = 1, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  labs(
    x = "Zurich logFC",
    y = "Poland logFC",
    title = paste0("COMPERA_number\nr = ", round(cor_comp, 2))
  ) +
  theme_classic()
```

```{r score_heatmap_helpers, message=FALSE, warning=FALSE}
# score for ranking (for GSEA later)
add_score <- function(de) {
  de$score <- de$logFC * (-log10(de$P.Value))
  de
}

# heatmap: top20 up/down by score
plot_top_score_heatmap <- function(expr, clin, de, samples, ha, title_text) {
  de <- add_score(de)

  de_sub <- de[match(rownames(expr), de$Gene), ]
  rownames(de_sub) <- de_sub$Gene

  up   <- de_sub[order(de_sub$score, decreasing = TRUE), , drop = FALSE]
  down <- de_sub[order(de_sub$score, decreasing = FALSE), , drop = FALSE]

  top_up   <- rownames(up)[1:20]
  top_down <- rownames(down)[1:20]

  genes_use <- c(top_up, top_down)

  mat <- as.matrix(expr[genes_use, samples, drop = FALSE])

  # z-score per protein (row), same as your heatmaps
  matz <- t(scale(t(mat)))
  matz[is.nan(matz)] <- 0

  hm_cols <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

  Heatmap(
    matz,
    name = "Z-score",
    col = hm_cols,
    top_annotation = ha,
    cluster_rows = FALSE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    row_names_gp = gpar(fontsize = 13),
    show_column_names = TRUE,
    column_names_gp = gpar(fontsize = 8),
    row_title = title_text,
    column_title = NULL,
    height = unit(nrow(matz) * 5, "mm")
  ) |>
    draw(heatmap_legend_side = "right", annotation_legend_side = "right")
}
```

```{r add_tapse_group}
clin_z <- clin_z %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = dplyr::case_when(
      is.na(TAPSE_sPAP)      ~ NA_character_,
      TAPSE_sPAP < 0.32      ~ "<0.32",
      TAPSE_sPAP >= 0.32     ~ "≥0.32"
    ),
    TAPSE_group = factor(TAPSE_group, levels = c("≥0.32", "<0.32"))
  )

clin_p <- clin_p %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    TAPSE_group = dplyr::case_when(
      is.na(TAPSE_sPAP)      ~ NA_character_,
      TAPSE_sPAP < 0.32      ~ "<0.32",
      TAPSE_sPAP >= 0.32     ~ "≥0.32"
    ),
    TAPSE_group = factor(TAPSE_group, levels = c("≥0.32", "<0.32"))
  )
```

```{r heatmaps_scores_zurich_tapseGroup, fig.height=10, fig.width=10.8}
# ---- add TAPSE/sPAP group to clin_z once ----
clin_z <- clin_z %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    `TAPSE/sPAP` = dplyr::case_when(
      is.na(TAPSE_sPAP)  ~ NA_character_,
      TAPSE_sPAP < 0.32  ~ "<0.32",
      TRUE               ~ "≥0.32"
    ),
    `TAPSE/sPAP` = factor(`TAPSE/sPAP`, levels = c("≥0.32", "<0.32"))
  )

# ---- colors (same palette logic) ----
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "low"               = "#9CA3AF",
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)
cols_tapse_spap <- c("≥0.32" = "#9CA3AF", "<0.32" = "#A63A2A")

# 1) SSc-PAH vs SSc-noPAH
samples <- intersect(colnames(data_z), clin_z$Barcode)
meta_hm <- clin_z[match(samples, clin_z$Barcode), ]

ha <- HeatmapAnnotation(
  Group        = meta_hm$Group,
  COMPERA      = factor(meta_hm$COMPERA_4strata, levels = c("low","intermediate-low","intermediate-high","high")),
  `TAPSE/sPAP` = meta_hm$`TAPSE/sPAP`,
  col = list(Group = cols_group, COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_z,
  clin = clin_z,
  de = DE_z,
  samples = samples,
  ha = ha,
  title_text = "Zurich: top20 up/down by score\nSSc-PAH vs SSc-noPAH"
)

# 2) TAPSE_sPAP (continuous), SSc-PAH only
meta2 <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

samples2 <- meta2$Barcode

ha2 <- HeatmapAnnotation(
  COMPERA      = factor(meta2$COMPERA_4strata, levels = c("low","intermediate-low","intermediate-high","high")),
  `TAPSE/sPAP` = meta2$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_z,
  clin = clin_z,
  de = DE_z_TAPSE,
  samples = samples2,
  ha = ha2,
  title_text = "Zurich: top20 up/down by score\nTAPSE_sPAP (continuous), SSc-PAH"
)

# 3) TAPSE_sPAP groups (<0.32 vs ≥0.32), SSc-PAH only
meta3 <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

samples3 <- meta3$Barcode

ha3 <- HeatmapAnnotation(
  COMPERA      = factor(meta3$COMPERA_4strata, levels = c("low","intermediate-low","intermediate-high","high")),
  `TAPSE/sPAP` = meta3$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_z,
  clin = clin_z,
  de = DE_z_bin,
  samples = samples3,
  ha = ha3,
  title_text = "Zurich: top20 up/down by score\nTAPSE_sPAP <0.32 vs ≥0.32, SSc-PAH"
)

# 4) COMPERA_number (continuous), SSc-PAH only
meta4 <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(COMPERA_number), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

samples4 <- meta4$Barcode

ha4 <- HeatmapAnnotation(
  COMPERA      = factor(meta4$COMPERA_4strata, levels = c("low","intermediate-low","intermediate-high","high")),
  `TAPSE/sPAP` = meta4$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_z,
  clin = clin_z,
  de = DE_z_comp,
  samples = samples4,
  ha = ha4,
  title_text = "Zurich: top20 up/down by score\nCOMPERA_number (continuous), SSc-PAH"
)
```

```{r heatmaps_scores_poland_tapseGroup, fig.height=10, fig.width=10.8}
# ---- add TAPSE/sPAP group to clin_p once ----
clin_p <- clin_p %>%
  dplyr::mutate(
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    `TAPSE/sPAP` = dplyr::case_when(
      is.na(TAPSE_sPAP)  ~ NA_character_,
      TAPSE_sPAP < 0.32  ~ "<0.32",
      TRUE               ~ "≥0.32"
    ),
    `TAPSE/sPAP` = factor(`TAPSE/sPAP`, levels = c("≥0.32", "<0.32"))
  )

# ---- colors ----
cols_group <- c("SSc-noPAH"="#004C8C", "SSc-PAH"="#A63A2A")
cols_compera <- c(
  "Low"               = "#9CA3AF",
  "Intermediate-low"  = "#DDB7B2",
  "Intermediate-high" = "#C7685F",
  "High"              = "#A63A2A"
)
cols_tapse_spap <- c("≥0.32" = "#9CA3AF", "<0.32" = "#A63A2A")

# 1) SSc-PAH vs SSc-noPAH
samples <- intersect(colnames(data_p), clin_p$Barcode)
meta_hm <- clin_p[match(samples, clin_p$Barcode), ]

ha <- HeatmapAnnotation(
  Group        = meta_hm$Group,
  COMPERA      = factor(meta_hm$Compera_4strata, levels = c("Low","Intermediate-low","Intermediate-high","High")),
  `TAPSE/sPAP` = meta_hm$`TAPSE/sPAP`,
  col = list(Group = cols_group, COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_p,
  clin = clin_p,
  de = DE_p,
  samples = samples,
  ha = ha,
  title_text = "Poland: top20 up/down by score\nSSc-PAH vs SSc-noPAH"
)

# 2) TAPSE_sPAP (continuous), SSc-PAH only
meta2 <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

samples2 <- meta2$Barcode

ha2 <- HeatmapAnnotation(
  COMPERA      = factor(meta2$Compera_4strata, levels = c("Low","Intermediate-low","Intermediate-high","High")),
  `TAPSE/sPAP` = meta2$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_p,
  clin = clin_p,
  de = DE_p_TAPSE,
  samples = samples2,
  ha = ha2,
  title_text = "Poland: top20 up/down by score\nTAPSE_sPAP (continuous), SSc-PAH"
)

# 3) TAPSE_sPAP groups (<0.32 vs ≥0.32), SSc-PAH only
meta3 <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

samples3 <- meta3$Barcode

ha3 <- HeatmapAnnotation(
  COMPERA      = factor(meta3$Compera_4strata, levels = c("Low","Intermediate-low","Intermediate-high","High")),
  `TAPSE/sPAP` = meta3$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_p,
  clin = clin_p,
  de = DE_p_bin,
  samples = samples3,
  ha = ha3,
  title_text = "Poland: top20 up/down by score\nTAPSE_sPAP <0.32 vs ≥0.32, SSc-PAH"
)

# 4) Compera_number (continuous), SSc-PAH only
meta4 <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(Compera_number), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

samples4 <- meta4$Barcode

ha4 <- HeatmapAnnotation(
  COMPERA      = factor(meta4$Compera_4strata, levels = c("Low","Intermediate-low","Intermediate-high","High")),
  `TAPSE/sPAP` = meta4$`TAPSE/sPAP`,
  col = list(COMPERA = cols_compera, `TAPSE/sPAP` = cols_tapse_spap),
  na_col = "#E5E7EB"
)

plot_top_score_heatmap(
  expr = data_p,
  clin = clin_p,
  de = DE_p_comp,
  samples = samples4,
  ha = ha4,
  title_text = "Poland: top20 up/down by score\nCOMPERA_number (continuous), SSc-PAH"
)
```

## Check age and sex by TAPSE/sPAP group

```{r age_sex_tapse_helpers, message=FALSE, warning=FALSE}
library(ggpubr)

prepare_tapse_group_df <- function(clin, cohort_label) {
  clin %>%
    dplyr::mutate(
      TAPSE_sPAP = as.numeric(TAPSE_sPAP),
      `TAPSE/sPAP` = dplyr::case_when(
        is.na(TAPSE_sPAP) ~ NA_character_,
        TAPSE_sPAP < 0.32 ~ "<0.32",
        TRUE ~ "≥0.32"
      ),
      `TAPSE/sPAP` = factor(`TAPSE/sPAP`, levels = c("≥0.32", "<0.32")),
      Age = as.numeric(Age),
      Sex = factor(Sex),
      Cohort = cohort_label
    ) %>%
    dplyr::filter(!is.na(`TAPSE/sPAP`))
}

age_test_tbl <- function(df) {
  df %>%
    dplyr::filter(!is.na(Age)) %>%
    dplyr::summarise(
      n = dplyr::n(),
      p_value = wilcox.test(Age ~ `TAPSE/sPAP`, data = dplyr::cur_data())$p.value,
      .groups = "drop"
    )
}

sex_test_tbl <- function(df) {
  tab <- table(df$`TAPSE/sPAP`, df$Sex)
  list(table = tab, fisher = fisher.test(tab))
}

plot_age_by_tapse <- function(df, cohort_label) {
  p_age <- wilcox.test(Age ~ `TAPSE/sPAP`, data = df %>% dplyr::filter(!is.na(Age)))$p.value

  ggplot(df, aes(x = `TAPSE/sPAP`, y = Age, fill = `TAPSE/sPAP`)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA) +
    geom_jitter(width = 0.1, size = 2, alpha = 0.7) +
    annotate(
      "text", x = 1.5, y = Inf,
      label = paste0("Wilcoxon p = ", signif(p_age, 3)),
      vjust = 1.4, size = 4
    ) +
    theme_classic() +
    labs(
      title = paste0(cohort_label, " cohort: Age by TAPSE/sPAP group"),
      y = "Age (years)",
      x = "TAPSE/sPAP group"
    ) +
    theme(legend.position = "none")
}

plot_sex_by_tapse <- function(df, cohort_label) {
  df %>%
    dplyr::filter(!is.na(Sex)) %>%
    ggplot(aes(x = `TAPSE/sPAP`, fill = Sex)) +
    geom_bar(position = "fill") +
    theme_classic() +
    labs(
      title = paste0(cohort_label, " cohort: Sex distribution by TAPSE/sPAP group"),
      y = "Proportion",
      x = "TAPSE/sPAP group"
    )
}

clin_z_tapse <- prepare_tapse_group_df(clin_z, "Zurich")
clin_p_tapse <- prepare_tapse_group_df(clin_p, "Poland")
```

```{r age_sex_poland, fig.height=4.5, fig.width=6.5, message=FALSE, warning=FALSE}
age_test_tbl(clin_p_tapse)
sex_test_p <- sex_test_tbl(clin_p_tapse)
sex_test_p$table
sex_test_p$fisher

plot_age_by_tapse(clin_p_tapse, "Poland")
plot_sex_by_tapse(clin_p_tapse, "Poland")
```

```{r age_sex_zurich, fig.height=4.5, fig.width=6.5, message=FALSE, warning=FALSE}
age_test_tbl(clin_z_tapse)
sex_test_z <- sex_test_tbl(clin_z_tapse)
sex_test_z$table
sex_test_z$fisher

plot_age_by_tapse(clin_z_tapse, "Zurich")
plot_sex_by_tapse(clin_z_tapse, "Zurich")
```

# Final lasso

```{r lasso_final_prepare_and_helpers, message=FALSE, warning=FALSE}
library(glmnet)
library(pROC)
library(Matrix)

# SSc-PAH only for the training/testing task.
# Outcome: 1 = low TAPSE/sPAP (<0.32), 0 = preserved TAPSE/sPAP (≥0.32).
train_p <- clin_p %>%
  dplyr::mutate(TAPSE_sPAP = as.numeric(TAPSE_sPAP), Age = as.numeric(Age), Sex = factor(Sex)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

test_z <- clin_z %>%
  dplyr::mutate(TAPSE_sPAP = as.numeric(TAPSE_sPAP), Age = as.numeric(Age), Sex = factor(Sex)) %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

y_p <- ifelse(train_p$TAPSE_sPAP < 0.32, 1, 0)
y_z <- ifelse(test_z$TAPSE_sPAP < 0.32, 1, 0)

make_design <- function(expr_mat, clin_df, include_age = TRUE, include_sex = TRUE){
  X_prot <- t(expr_mat[, clin_df$Barcode, drop = FALSE])

  cov_parts <- list()
  if (include_age) cov_parts$Age <- as.numeric(clin_df$Age)
  if (include_sex) cov_parts$Sex <- factor(clin_df$Sex)

  if (length(cov_parts) == 0) {
    X <- X_prot
    pf <- rep(1, ncol(X))
    cov_names <- character(0)
  } else {
    X_cov <- model.matrix(~ ., data = as.data.frame(cov_parts))[, -1, drop = FALSE]
    X <- cbind(X_cov, X_prot)
    pf <- c(rep(0, ncol(X_cov)), rep(1, ncol(X_prot)))
    cov_names <- colnames(X_cov)
  }

  list(X = X, pf = pf, cov_names = cov_names, prot_names = colnames(X_prot))
}

choose_lambda_by_protein_count <- function(cvfit, prot_names, min_prot = 3, max_prot = 15){
  lambdas <- cvfit$glmnet.fit$lambda
  beta_mat <- as.matrix(coef(cvfit$glmnet.fit, s = lambdas))
  prot_idx <- which(rownames(beta_mat) %in% prot_names)
  nprot <- apply(beta_mat[prot_idx, , drop = FALSE] != 0, 2, sum)

  ok <- which(nprot >= min_prot & nprot <= max_prot)
  if (length(ok) > 0) {
    return(lambdas[ok[which.min(cvfit$cvm[match(lambdas[ok], cvfit$lambda)])]])
  }

  target <- round((min_prot + max_prot) / 2)
  lambdas[which.min(abs(nprot - target))]
}

bootstrap_stability_lasso <- function(expr_mat, clin_df, y,
                                      B = 500,
                                      include_age = TRUE,
                                      include_sex = TRUE,
                                      alpha = 1,
                                      nfolds = 5,
                                      min_prot = 3,
                                      max_prot = 15,
                                      seed = 1){
  set.seed(seed)
  stopifnot(all(clin_df$Barcode %in% colnames(expr_mat)))

  expr_mat <- expr_mat[, clin_df$Barcode, drop = FALSE]
  y <- as.numeric(y)

  des <- make_design(expr_mat, clin_df, include_age, include_sex)
  X_full <- des$X
  pf <- des$pf
  all_prots <- des$prot_names
  n <- nrow(X_full)

  sel_list <- vector("list", B)
  coef_list <- vector("list", B)
  auc_oob <- rep(NA_real_, B)
  lambda_used <- rep(NA_real_, B)

  for (b in seq_len(B)) {
    idx_boot <- sample(seq_len(n), size = n, replace = TRUE)
    idx_oob <- setdiff(seq_len(n), unique(idx_boot))

    Xb <- X_full[idx_boot, , drop = FALSE]
    yb <- y[idx_boot]

    cvfit <- glmnet::cv.glmnet(
      x = Xb, y = yb, family = "binomial", alpha = alpha,
      penalty.factor = pf, standardize = TRUE, nfolds = nfolds,
      type.measure = "auc"
    )

    lam <- choose_lambda_by_protein_count(cvfit, prot_names = all_prots,
                                          min_prot = min_prot, max_prot = max_prot)
    lambda_used[b] <- lam

    beta <- as.matrix(coef(cvfit$glmnet.fit, s = lam))[, 1]
    names(beta) <- rownames(as.matrix(coef(cvfit$glmnet.fit, s = lam)))

    sel_prot <- intersect(names(beta)[beta != 0], all_prots)
    sel_list[[b]] <- sel_prot
    coef_list[[b]] <- beta

    if (length(idx_oob) >= 8 && length(unique(y[idx_oob])) == 2) {
      fitb <- glmnet::glmnet(
        x = Xb, y = yb, family = "binomial", alpha = alpha,
        penalty.factor = pf, standardize = TRUE, lambda = lam
      )
      p_oob <- as.numeric(predict(fitb, newx = X_full[idx_oob, , drop = FALSE], type = "response"))
      auc_oob[b] <- as.numeric(pROC::auc(pROC::roc(y[idx_oob], p_oob, quiet = TRUE)))
    }
  }

  sel_counts <- table(unlist(sel_list))
  freq <- as.numeric(sel_counts[all_prots])
  freq[is.na(freq)] <- 0
  freq <- freq / B

  sign_pos <- rep(NA_real_, length(all_prots))
  names(sign_pos) <- all_prots

  for (p in all_prots) {
    signs <- c()
    for (b in seq_len(B)) {
      beta <- coef_list[[b]]
      if (!is.null(beta) && p %in% names(beta) && beta[p] != 0) {
        signs <- c(signs, sign(beta[p]))
      }
    }
    if (length(signs) > 0) sign_pos[p] <- mean(signs > 0)
  }

  summary <- data.frame(
    protein = all_prots,
    select_freq = freq,
    sign_pos_frac = sign_pos,
    stringsAsFactors = FALSE
  ) %>%
    dplyr::arrange(dplyr::desc(select_freq))

  list(
    summary = summary,
    sel_list = sel_list,
    coef_list = coef_list,
    auc_oob = auc_oob,
    lambda_used = lambda_used,
    settings = list(B = B, include_age = include_age, include_sex = include_sex,
                    alpha = alpha, nfolds = nfolds, min_prot = min_prot, max_prot = max_prot)
  )
}

get_scaler <- function(expr_mat, proteins){
  m <- rowMeans(expr_mat[proteins, , drop = FALSE], na.rm = TRUE)
  s <- apply(expr_mat[proteins, , drop = FALSE], 1, sd, na.rm = TRUE)
  s[s == 0 | is.na(s)] <- 1
  list(mu = m, sd = s)
}

apply_scaler <- function(expr_mat, proteins, scaler){
  Z <- expr_mat[proteins, , drop = FALSE]
  Z <- sweep(Z, 1, scaler$mu, "-")
  Z <- sweep(Z, 1, scaler$sd, "/")
  Z[is.na(Z)] <- 0
  Z
}

fit_logistic_refit <- function(expr_mat, clin_df, proteins, cutoff = 0.32,
                               include_age = TRUE, include_sex = TRUE){
  df <- clin_df %>% dplyr::mutate(y = ifelse(TAPSE_sPAP < cutoff, 1, 0))
  stopifnot(all(df$Barcode %in% colnames(expr_mat)))

  scaler <- get_scaler(expr_mat[, df$Barcode, drop = FALSE], proteins)
  Z <- t(apply_scaler(expr_mat[, df$Barcode, drop = FALSE], proteins, scaler)) %>% as.data.frame()
  Z$Barcode <- df$Barcode

  dd <- df %>% dplyr::select(Barcode, y, Age, Sex) %>% dplyr::left_join(Z, by = "Barcode")

  rhs <- c(if (include_age) "Age", if (include_sex) "Sex", proteins)
  fml <- as.formula(paste("y ~", paste(rhs, collapse = " + ")))

  m <- glm(fml, data = dd, family = binomial())
  list(model = m, scaler = scaler, proteins = proteins, include_age = include_age, include_sex = include_sex)
}

predict_logistic_refit <- function(fit, expr_mat, clin_df){
  df <- clin_df
  stopifnot(all(df$Barcode %in% colnames(expr_mat)))

  Z <- t(apply_scaler(expr_mat[, df$Barcode, drop = FALSE], fit$proteins, fit$scaler)) %>% as.data.frame()
  Z$Barcode <- df$Barcode

  keep_cols <- intersect(c("Barcode", "Age", "Sex", "TAPSE_sPAP", "TAPSE_group", "Group"), colnames(df))
  dd <- df %>% dplyr::select(dplyr::all_of(keep_cols)) %>% dplyr::left_join(Z, by = "Barcode")

  dd$score_eta <- as.numeric(predict(fit$model, newdata = dd, type = "link"))
  dd$score_p <- as.numeric(predict(fit$model, newdata = dd, type = "response"))
  dd
}

signed_mean_score <- function(expr_mat, clin_df, proteins, sign_map, scaler){
  stopifnot(all(names(sign_map) %in% proteins))
  Z <- apply_scaler(expr_mat[, clin_df$Barcode, drop = FALSE], proteins, scaler)
  Zs <- sweep(Z, 1, sign_map[proteins], "*")
  score <- colMeans(Zs, na.rm = TRUE)

  out <- clin_df %>% dplyr::select(dplyr::any_of(c("Barcode", "Age", "Sex", "TAPSE_sPAP", "TAPSE_group", "Group")))
  out$score_signed <- as.numeric(score[match(out$Barcode, names(score))])
  out
}
```

```{r preselect_proteins_DE_Poland, message=FALSE, warning=FALSE}
library(limma)
library(dplyr)

# Poland samples for DE (PAH vs noPAH)
de_meta_p <- clin_p %>%
  filter(Group %in% c("SSc-PAH", "SSc-noPAH")) %>%
  filter(!is.na(Age), !is.na(Sex)) %>%
  filter(Barcode %in% colnames(data_p))

X_de <- as.matrix(data_p[, de_meta_p$Barcode, drop = FALSE])

# Design: Group + Age + Sex
de_meta_p$Group <- factor(de_meta_p$Group, levels = c("SSc-noPAH", "SSc-PAH"))
de_meta_p$Sex   <- factor(de_meta_p$Sex)

design <- model.matrix(~ Group + Age + Sex, data = de_meta_p)

fit <- lmFit(X_de, design)
fit <- eBayes(fit)

DE_pah_vs_nopah <- topTable(fit, coef = "GroupSSc-PAH", number = Inf, adjust.method = "BH")
DE_pah_vs_nopah$protein <- rownames(DE_pah_vs_nopah)

# Choose your filter (match your original logic)
# Example: adj.P.Val < 0.05 (optionally add |logFC| cutoff)
preselect_proteins <- DE_pah_vs_nopah %>%
  dplyr::filter(adj.P.Val < 0.05) %>%
 # dplyr::filter(protein %in% dplyr::filter(DE_p_bin, P.Value < 0.05)$Gene) %>%
  pull(protein)

length(preselect_proteins)
head(preselect_proteins)

qs2::qs_save(DE_pah_vs_nopah, here::here(output_dir_data, "DE_Poland_SScPAH_vs_SScnoPAH_AgeSex.qs2"))
qs2::qs_save(preselect_proteins, here::here(output_dir_data, "preselect_proteins_DEadjP005.qs2"))

```


## Stability-selection lasso after protein preselection

```{r stability_lasso_with_preselection, message=FALSE, warning=FALSE}
# Intersect with proteins available in BOTH cohorts
common_preselect <- intersect(preselect_proteins, intersect(rownames(data_p), rownames(data_z)))
length(common_preselect)

# Build training/test expression matrices for LASSO (proteins x samples)
expr_p_tr_pre <- as.matrix(data_p[common_preselect, train_p$Barcode, drop = FALSE])
expr_z_te_pre <- as.matrix(data_z[common_preselect, test_z$Barcode, drop = FALSE])

# Run bootstrap stability selection on the preselected feature space
B <- 500
stab_pre <- bootstrap_stability_lasso(
  expr_mat = expr_p_tr_pre,
  clin_df  = train_p,
  y        = y_p,
  B = B,
  include_age = TRUE,
  include_sex = TRUE,    # or FALSE if you want Age-only adjustment
  alpha = 1,
  nfolds = 5,
  min_prot = 3,
  max_prot = 100,
  seed = 1
)

qs2::qs_save(stab_pre, here::here(output_dir_data, paste0("stability_lasso_preselected_B", B, "_AgeSex.qs2")))
head(stab_pre$summary, 20)
```


---



```{r stability_panel_size_distribution, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

# number of proteins selected per bootstrap
panel_sizes <- sapply(stab_pre$sel_list, length)

df_sizes <- data.frame(size = panel_sizes)

p_size <- ggplot(df_sizes, aes(x = size)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = sort(unique(panel_sizes))) +
  labs(
    x = "Number of proteins selected (per bootstrap)",
    y = "Count",
    title = "Stability selection: distribution of panel sizes"
  ) +
  theme_classic(base_size = 12)

p_size
ggsave(here::here(output_dir_figs, paste0("stability_panel_size_distribution_B", stab_pre$settings$B, ".png")),
       p_size, width = 6, height = 4.2, dpi = 300)

summary(panel_sizes)
table(panel_sizes)
```


---

## 2) “Most frequent combination” (top recurring panels)
This is useful, but usually you’ll see many singletons unless you restrict to a manageable feature universe. So we compute frequency of **exact sets**.

```{r stability_most_frequent_panels, message=FALSE, warning=FALSE}
library(tibble)

panel_keys <- sapply(stab_pre$sel_list, function(v){
  if (length(v) == 0) return(NA_character_)
  paste(sort(v), collapse = " | ")
})

panel_tab <- sort(table(panel_keys), decreasing = TRUE)
panel_df <- data.frame(
  panel = names(panel_tab),
  n = as.integer(panel_tab),
  freq = as.integer(panel_tab) / length(stab_pre$sel_list),
  stringsAsFactors = FALSE
) %>% 
  filter(!is.na(panel))

head(panel_df, 20)
```


**Optional plot (top 10 panels):**
```{r stability_most_frequent_panels_plot, fig.height=4.5, fig.width=12, message=FALSE, warning=FALSE}
topM <- 10
p_panels <- panel_df %>%
  slice_head(n = topM) %>%
  mutate(panel = factor(panel, levels = rev(panel))) %>%
  ggplot(aes(x = panel, y = freq)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Frequency", title = paste0("Top ", topM, " most frequent exact panels")) +
  theme_classic(base_size = 12)

p_panels
ggsave(here::here(output_dir_figs, paste0("stability_top", topM, "_panels_B", stab_pre$settings$B, ".png")),
       p_panels, width = 9, height = 4.5, dpi = 300)
```


---

## 3) Co-selection heatmap (which proteins tend to appear together)
This is *often more informative* than exact combinations.

```{r stability_co_selection_heatmap, message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
library(Matrix)

# pick a manageable set (e.g. top 30 by selection freq)
topK <- 30
top_prots <- stab_pre$summary %>% slice_head(n = topK) %>% pull(protein)

# build co-selection matrix
M <- matrix(0, nrow = topK, ncol = topK, dimnames = list(top_prots, top_prots))

for (sel in stab_pre$sel_list) {
  sel <- intersect(sel, top_prots)
  if (length(sel) >= 2) {
    idx <- match(sel, top_prots)
    M[idx, idx] <- M[idx, idx] + 1
  } else if (length(sel) == 1) {
    i <- match(sel, top_prots)
    M[i, i] <- M[i, i] + 1
  }
}

# convert to frequency
M_freq <- M / length(stab_pre$sel_list)

# quick base heatmap
heatmap(M_freq, symm = TRUE, margins = c(8, 8),
        main = paste0("Co-selection frequency (top ", topK, " proteins)"))
```
        
## Build final panel and scores

```{r build_scores_from_stability_preselected, message=FALSE, warning=FALSE}
stab_df <- stab_pre$summary

# pick your panel
panel_size <- 5
panel_prots <- stab_df %>% slice_head(n = panel_size) %>% pull(protein)
panel_prots

# sign map from bootstrap
sign_map <- stab_df %>%
  filter(protein %in% panel_prots) %>%
  mutate(sign = ifelse(sign_pos_frac >= 0.5, 1, -1)) %>%
  dplyr::select(protein, sign) %>%
  tibble::deframe()

# IMPORTANT: refit/scores must use the FULL cohort matrices (data_p/data_z)
# but only these proteins
refit_pre <- fit_logistic_refit(
  expr_mat = as.matrix(data_p),  # proteins x samples
  clin_df  = train_p,
  proteins = panel_prots,
  cutoff = 0.32,
  include_age = TRUE,
  include_sex = TRUE
)

pred_p <- predict_logistic_refit(refit_pre, as.matrix(data_p), train_p)
pred_z <- predict_logistic_refit(refit_pre, as.matrix(data_z), test_z)

signed_p <- signed_mean_score(as.matrix(data_p), train_p, panel_prots, sign_map, refit_pre$scaler)
signed_z <- signed_mean_score(as.matrix(data_z), test_z, panel_prots, sign_map, refit_pre$scaler)
```


---

## Final lasso performance and score plots

### ROC with AUC confidence intervals

```{r roc_refit_preselected, fig.height=4.5, fig.width=5, message=FALSE, warning=FALSE}
y_tr <- ifelse(pred_p$TAPSE_sPAP < 0.32, 1, 0)
y_te <- ifelse(pred_z$TAPSE_sPAP < 0.32, 1, 0)

roc_tr <- pROC::roc(y_tr, pred_p$score_p, quiet = TRUE, ci = TRUE, direction = "<")
roc_te <- pROC::roc(y_te, pred_z$score_p, quiet = TRUE, ci = TRUE, direction = "<")

auc_tr <- as.numeric(pROC::auc(roc_tr))
auc_te <- as.numeric(pROC::auc(roc_te))
ci_auc_tr <- as.numeric(pROC::ci.auc(roc_tr))
ci_auc_te <- as.numeric(pROC::ci.auc(roc_te))

# 95% ROC-curve confidence bands
set.seed(1)
ci_se_tr <- pROC::ci.se(
  roc_tr,
  specificities = seq(0, 1, length.out = 101),
  boot.n = 2000,
  progress = "none"
)
ci_se_te <- pROC::ci.se(
  roc_te,
  specificities = seq(0, 1, length.out = 101),
  boot.n = 2000,
  progress = "none"
)

plot(
  roc_tr,
  legacy.axes = TRUE,
  identity = FALSE,
  col = "red",
  lwd = 2,
  xlab = "1 - Specificity",
  ylab = "Sensitivity",
  main = paste0(
    "Stable panel refit-logistic
",
    "Poland AUC=", sprintf("%.2f", auc_tr),
    " | Zurich AUC=", sprintf("%.2f", auc_te)
  )
)

plot(
  ci_se_tr,
  type = "shape",
  col = adjustcolor("red", alpha.f = 0.20),
  border = NA,
  no.roc = TRUE,
  add = TRUE,
  legacy.axes = TRUE
)
plot(
  ci_se_te,
  type = "shape",
  col = adjustcolor("black", alpha.f = 0.15),
  border = NA,
  no.roc = TRUE,
  add = TRUE,
  legacy.axes = TRUE
)

plot(roc_tr, legacy.axes = TRUE, identity = FALSE, col = "red", lwd = 2, add = TRUE)
plot(roc_te, legacy.axes = TRUE, identity = FALSE, col = "black", lwd = 2, add = TRUE)

legend(
  "bottomright",
  inset = 0.02,
  legend = c(
    paste0(
      "Poland (train): AUC ", sprintf("%.2f", auc_tr),
      " [", sprintf("%.2f", ci_auc_tr[1]), ", ", sprintf("%.2f", ci_auc_tr[3]), "]"
    ),
    paste0(
      "Zurich (test): AUC ", sprintf("%.2f", auc_te),
      " [", sprintf("%.2f", ci_auc_te[1]), ", ", sprintf("%.2f", ci_auc_te[3]), "]"
    )
  ),
  col = c("red", "black"),
  lwd = 2,
  bty = "n",
  cex = 0.99,
  seg.len = 2.3
)
```


### Three-group score plot

```{r boxplot_eta_preselected, fig.height=4.8, fig.width=9, message=FALSE, warning=FALSE}
library(ggpubr)

make_3groups <- function(clin_df){
  clin_df %>%
    mutate(
      group3 = case_when(
        Group == "SSc-noPAH" ~ "SSc-noPAH",
        Group == "SSc-PAH" & TAPSE_sPAP >= 0.32 ~ "SSc-PAH≥0.32",
        Group == "SSc-PAH" & TAPSE_sPAP < 0.32  ~ "SSc-PAH<0.32",
        TRUE ~ NA_character_
      ),
      group3 = factor(group3, levels = c("SSc-noPAH", "SSc-PAH≥0.32", "SSc-PAH<0.32"))
    )
}

# score ALL subjects to visualize, but plot ETA
clin_p_all <- clin_p %>% filter(Barcode %in% colnames(data_p), !is.na(Age), !is.na(Sex))
clin_z_all <- clin_z %>% filter(Barcode %in% colnames(data_z), !is.na(Age), !is.na(Sex))

pred_p_all <- predict_logistic_refit(refit_pre, as.matrix(data_p), clin_p_all) %>% make_3groups()
pred_z_all <- predict_logistic_refit(refit_pre, as.matrix(data_z), clin_z_all) %>% make_3groups()

df_plot <- bind_rows(
  pred_p_all %>% mutate(Cohort = "Poland"),
  pred_z_all %>% mutate(Cohort = "Zurich")
) %>% filter(!is.na(group3))

p <- ggplot(df_plot, aes(x = group3, y = score_eta)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.3, alpha = 0.7) +
  facet_wrap(~Cohort, scales = "free_x") +
  labs(x = NULL, y = "Score (log-odds; linear predictor)",
       title = paste0("Stable panel refit-logistic score (", length(panel_prots), " proteins)")) +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

p
```

## Correlations

```{r corr_prepare_pah_only, message=FALSE, warning=FALSE}

panel_prots <- refit_pre$proteins

# helper to build a long dataframe for correlations
make_corr_df <- function(clin_df, data_mat, cohort_label, score_df = NULL,
                         score_var = c("score_eta","score_p","score_signed")) {
  score_var <- match.arg(score_var)

  # subset to SSc-PAH with TAPSE/sPAP
  cc <- clin_df %>%
    filter(Group == "SSc-PAH") %>%
    filter(!is.na(TAPSE_sPAP)) %>%
    filter(Barcode %in% colnames(data_mat))

  # protein values (wide)
  prot_wide <- t(as.matrix(data_mat[panel_prots, cc$Barcode, drop = FALSE])) %>%
    as.data.frame()
  prot_wide$Barcode <- cc$Barcode

  out <- cc %>%
    dplyr::select(Barcode, TAPSE_sPAP) %>%
    left_join(prot_wide, by = "Barcode") %>%
    mutate(Cohort = cohort_label)

  # attach score if provided
  if (!is.null(score_df)) {
    out <- out %>%
      left_join(score_df %>% dplyr::select(Barcode, all_of(score_var)), by = "Barcode") %>%
      rename(Score = all_of(score_var))
  } else {
    out$Score <- NA_real_
  }

  out
}

# Use your previously created score data frames:
# pred_p, pred_z contain score_eta and score_p for SSc-PAH train/test
# signed_p, signed_z contain score_signed if you want that too

df_corr_p <- make_corr_df(train_p, data_p, "Poland", score_df = pred_p, score_var = "score_eta")
df_corr_z <- make_corr_df(test_z,  data_z, "Zurich", score_df = pred_z, score_var = "score_eta")

# Combine
df_corr <- bind_rows(df_corr_p, df_corr_z)

# quick sanity check
df_corr %>% group_by(Cohort) %>% summarise(n = n())
```

```{r corr_table_proteins_and_score, message=FALSE, warning=FALSE}
cor_spearman <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 5) return(c(rho = NA_real_, p = NA_real_, n = sum(ok)))
  ct <- suppressWarnings(cor.test(x[ok], y[ok], method = "spearman", exact = FALSE))
  c(rho = unname(ct$estimate), p = ct$p.value, n = sum(ok))
}

features <- c(panel_prots, "Score")

corr_tbl <- df_corr %>%
  pivot_longer(cols = all_of(panel_prots), names_to = "Feature", values_to = "Value") %>%
  # add Score as another feature
  bind_rows(
    df_corr %>% transmute(Cohort, Barcode, TAPSE_sPAP, Feature = "Score", Value = Score)
  ) %>%
  group_by(Cohort, Feature) %>%
  summarise(
    rho = cor_spearman(Value, TAPSE_sPAP)["rho"],
    p   = cor_spearman(Value, TAPSE_sPAP)["p"],
    n   = cor_spearman(Value, TAPSE_sPAP)["n"],
    .groups = "drop"
  ) %>%
  group_by(Cohort) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  arrange(Cohort, p_adj)

corr_tbl
qs2::qs_save(corr_tbl, here::here(output_dir_data, "corr_TAPSEsPAP_panelProteins_and_score_Spearman.qs2"))
write.csv(corr_tbl, here::here(output_dir_data, "corr_TAPSEsPAP_panelProteins_and_score_Spearman.csv"),
          row.names = FALSE)
```

```{r corr_plot_score_vs_tapse, fig.height=3.8, fig.width=7.5, message=FALSE, warning=FALSE}
# pull rho/p for annotation
ann_score <- corr_tbl %>% filter(Feature == "Score") %>% dplyr::select(Cohort, rho, p, p_adj)

df_score <- df_corr %>%
  filter(!is.na(Score)) %>%
  left_join(ann_score, by = "Cohort")

p_score <- ggplot(df_score, aes(x = TAPSE_sPAP, y = Score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Cohort, scales = "free_y") +
  labs(
    x = "TAPSE/sPAP",
    y = "Score (linear predictor, log-odds)",
    title = "Score vs TAPSE/sPAP in SSc-PAH"
  ) +
  theme_classic(base_size = 12)

# Add annotation per facet (simple approach)
p_score <- p_score +
  geom_text(
    data = df_score %>% distinct(Cohort, rho, p, p_adj),
    aes(x = -Inf, y = Inf,
        label = paste0("Spearman ρ=", round(rho, 2),
                       "\np=", signif(p, 2))),
    hjust = -1, vjust = 1.1,
    inherit.aes = FALSE
  )

p_score
ggsave(here::here(output_dir_figs, "corr_score_vs_TAPSEsPAP_SScPAH.png"),
       p_score, width = 7.5, height = 3.8, dpi = 300)
```

```{r corr_plot_proteins_vs_tapse_faceted, fig.height=5, fig.width=10.5, message=FALSE, warning=FALSE}
# prepare long data for plotting
df_long <- df_corr %>%
  pivot_longer(cols = all_of(panel_prots), names_to = "Protein", values_to = "NPX") %>%
  left_join(
    corr_tbl %>% filter(Feature %in% panel_prots) %>%
      rename(Protein = Feature) %>%
      dplyr::select(Cohort, Protein, rho, p, p_adj),
    by = c("Cohort", "Protein")
  )

p_prot <- ggplot(df_long, aes(x = TAPSE_sPAP, y = NPX)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(Cohort ~ Protein, scales = "free_y") +
  labs(
    x = "TAPSE/sPAP",
    y = "Protein NPX",
    title = "Individual proteins vs TAPSE/sPAP in SSc-PAH"
  ) +
  theme_classic(base_size = 12) +
  theme(
    strip.text.x = element_text(size = 10),
    strip.text.y = element_text(size = 10)
  )

# annotate each facet with rho and FDR
ann_df <- df_long %>%
  distinct(Cohort, Protein, rho, p_adj, p)

p_prot <- p_prot +
  geom_text(
    data = ann_df,
    aes(x = -Inf, y = Inf,
        label = paste0("ρ=", round(rho, 2), "\np=", signif(p, 2))),
    hjust = -1, vjust = 1.1,
    inherit.aes = FALSE,
    size = 3
  )

p_prot
ggsave(here::here(output_dir_figs, "corr_proteins_vs_TAPSEsPAP_SScPAH_faceted.png"),
       p_prot, width = 10.5, height = 7.5, dpi = 300)
```


```{r zurich_withincohort_scaling_auc, fig.height=4.5, fig.width=5, message=FALSE, warning=FALSE}
library(dplyr)
library(pROC)

panel_prots <- refit_pre$proteins

# Zurich SSc-PAH only
z_pah <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

# Outcome: 1 = low TAPSE/sPAP, 0 = preserved TAPSE/sPAP
y_z <- ifelse(z_pah$TAPSE_sPAP < 0.32, 1, 0)

# Protein matrix: proteins x samples
Zmat <- as.matrix(data_z[panel_prots, z_pah$Barcode, drop = FALSE])

# Zurich-within-cohort z-score per protein
mu_z <- rowMeans(Zmat, na.rm = TRUE)
sd_z <- apply(Zmat, 1, sd, na.rm = TRUE)
sd_z[sd_z == 0 | is.na(sd_z)] <- 1

Zz <- sweep(Zmat, 1, mu_z, "-")
Zz <- sweep(Zz, 1, sd_z, "/")
Zz[is.na(Zz)] <- 0

# Signed mean z-score
s <- sign_map[panel_prots]
Zsigned <- sweep(Zz, 1, s, "*")
score_withinZ <- colMeans(Zsigned, na.rm = TRUE)

# ROC / AUC with CI
roc_withinZ <- pROC::roc(
  y_z,
  as.numeric(score_withinZ),
  quiet = TRUE,
  ci = TRUE,
  direction = "<"
)

auc_withinZ <- as.numeric(pROC::auc(roc_withinZ))
ci_auc_withinZ <- as.numeric(pROC::ci.auc(roc_withinZ))

# 95% ROC-curve confidence band
set.seed(1)
ci_se_withinZ <- pROC::ci.se(
  roc_withinZ,
  specificities = seq(0, 1, length.out = 101),
  boot.n = 2000,
  progress = "none"
)

plot(
  roc_withinZ,
  legacy.axes = TRUE,
  identity = FALSE,
  col = "black",
  lwd = 2,
  xlab = "1 - Specificity",
  ylab = "Sensitivity",
  main = paste0(
    "Zurich within-cohort scaled signed score\n",
    "Zurich AUC=", sprintf("%.2f", auc_withinZ)
  )
)

plot(
  ci_se_withinZ,
  type = "shape",
  col = adjustcolor("black", alpha.f = 0.15),
  border = NA,
  no.roc = TRUE,
  add = TRUE,
  legacy.axes = TRUE
)

plot(
  roc_withinZ,
  legacy.axes = TRUE,
  identity = FALSE,
  col = "black",
  lwd = 2,
  add = TRUE
)

legend(
  "bottomright",
  inset = 0.02,
  legend = paste0(
    "Zurich: AUC ", sprintf("%.2f", auc_withinZ),
    " [", sprintf("%.2f", ci_auc_withinZ[1]), ", ",
    sprintf("%.2f", ci_auc_withinZ[3]), "]"
  ),
  col = "black",
  lwd = 2,
  bty = "n",
  cex = 0.99,
  seg.len = 2.3
)

# Save base R plot
png(
  filename = here::here(output_dir_figs, "AUC_Zurich_withinCohortScaling_signedScore.png"),
  width = 5,
  height = 4.5,
  units = "in",
  res = 300
)

plot(
  roc_withinZ,
  legacy.axes = TRUE,
  identity = FALSE,
  col = "black",
  lwd = 2,
  xlab = "1 - Specificity",
  ylab = "Sensitivity",
  main = paste0(
    "Zurich within-cohort scaled signed score\n",
    "Zurich AUC=", sprintf("%.2f", auc_withinZ)
  )
)

plot(
  ci_se_withinZ,
  type = "shape",
  col = adjustcolor("black", alpha.f = 0.15),
  border = NA,
  no.roc = TRUE,
  add = TRUE,
  legacy.axes = TRUE
)

plot(
  roc_withinZ,
  legacy.axes = TRUE,
  identity = FALSE,
  col = "black",
  lwd = 2,
  add = TRUE
)

legend(
  "bottomright",
  inset = 0.02,
  legend = paste0(
    "Zurich: AUC ", sprintf("%.2f", auc_withinZ),
    " [", sprintf("%.2f", ci_auc_withinZ[1]), ", ",
    sprintf("%.2f", ci_auc_withinZ[3]), "]"
  ),
  col = "black",
  lwd = 2,
  bty = "n",
  cex = 0.99,
  seg.len = 2.3
)

dev.off()
```

## COX regression

```{r cox_prepare_surv_zurich, message=FALSE, warning=FALSE}
library(dplyr)
library(survival)

# -------------------------------
# 0) Choose which score to use
# -------------------------------
score_z <- pred_z %>%
  dplyr::select(Barcode, eta = score_eta)   # use linear predictor (recommended)

# -------------------------------
# 1) Robust survival column picker
# -------------------------------
pick_first_existing <- function(df, candidates) {
  hit <- candidates[candidates %in% colnames(df)]
  if (length(hit) == 0) return(NA_character_)
  hit[1]
}

# Try to guess your column names (adjust here if needed)
time_col  <- pick_first_existing(clin_z, c("DeathDAYS","death_days","FU_DAYS","followup_days","time_days","TimeDAYS","Time_days"))
event_col <- pick_first_existing(clin_z, c("Death","death","Death_event","event","Event","status","Status"))

# COMPERA numeric column (you mentioned COMPERA_number)
compera_col <- pick_first_existing(clin_z, c("COMPERA_number","COMPERA","COMPERA_num","COMPERA_score"))

cat("Using time_col =", time_col, "\n")
cat("Using event_col =", event_col, "\n")
cat("Using compera_col =", compera_col, "\n")

# If any are NA, set them manually like:
# time_col <- "DeathDAYS"
# event_col <- "Death"
# compera_col <- "COMPERA_number"

stopifnot(!is.na(time_col), !is.na(event_col), !is.na(compera_col))

# -------------------------------
# 2) Build Zurich SSc-PAH survival dataframe
# -------------------------------
df_surv_z <- clin_z %>%
  filter(Group == "SSc-PAH") %>%
  dplyr::select(Barcode, all_of(compera_col), all_of(time_col), all_of(event_col)) %>%
  left_join(score_z, by = "Barcode") %>%
  mutate(
    # numeric COMPERA
    COMPERA_number = as.numeric(.data[[compera_col]]),

    # time in days
    time_days = as.numeric(.data[[time_col]]),

    # event coding: try to handle yes/no, TRUE/FALSE, 1/0
    event_raw = .data[[event_col]],
    event = case_when(
      is.numeric(event_raw) ~ ifelse(event_raw > 0, 1, 0),
      is.logical(event_raw) ~ ifelse(event_raw, 1, 0),
      tolower(as.character(event_raw)) %in% c("yes","y","true","1","dead","death") ~ 1,
      tolower(as.character(event_raw)) %in% c("no","n","false","0","alive") ~ 0,
      TRUE ~ NA_real_
    )
  ) %>%
  filter(!is.na(time_days), !is.na(event), !is.na(COMPERA_number), !is.na(eta))

# quick check
df_surv_z %>% summarise(n=n(), events=sum(event), median_time=median(time_days))
table(df_surv_z$event)
```


---

## Chunk 2 — Cox models + c-index table (+ LRT)

```{r cox_zurich_cindex_table, message=FALSE, warning=FALSE}
library(dplyr)
library(survival)

m0 <- coxph(Surv(time_days, event) ~ COMPERA_number, data = df_surv_z, x = TRUE)
m1 <- coxph(Surv(time_days, event) ~ eta,            data = df_surv_z, x = TRUE)
m2 <- coxph(Surv(time_days, event) ~ COMPERA_number + eta, data = df_surv_z, x = TRUE)

c0 <- unname(summary(m0)$concordance[1])
c1 <- unname(summary(m1)$concordance[1])
c2 <- unname(summary(m2)$concordance[1])

cindex_tbl <- data.frame(
  model   = c("COMPERA", "Score", "COMPERA + score"),
  c_index = c(c0, c1, c2)
)

cindex_tbl

# Likelihood ratio test: does adding score improve COMPERA?
lrt <- anova(m0, m2, test = "LRT")
lrt
```


This gives you exactly the table format you showed (just with “Score” instead of “LASSO score” — rename if you want).

---

## Optional (recommended) — bootstrap CI for c-index difference (COMPERA vs COMPERA+score)

With small cohorts/events, c-index can be noisy. This gives you a quick sense of stability.

```{r cox_zurich_cindex_bootstrap, message=FALSE, warning=FALSE}
set.seed(1)
B <- 500

boot_delta <- replicate(B, {
  idx <- sample(seq_len(nrow(df_surv_z)), replace = TRUE)
  d <- df_surv_z[idx, , drop = FALSE]

  # handle rare bootstrap samples with 0 events
  if (length(unique(d$event)) < 2) return(NA_real_)

  m0b <- coxph(Surv(time_days, event) ~ COMPERA_number, data = d)
  m2b <- coxph(Surv(time_days, event) ~ COMPERA_number + eta, data = d)

  c0b <- unname(summary(m0b)$concordance[1])
  c2b <- unname(summary(m2b)$concordance[1])
  c2b - c0b
})

boot_delta <- boot_delta[is.finite(boot_delta)]
quantile(boot_delta, c(0.025, 0.5, 0.975))
mean(boot_delta)
```


---

## Optional — model summary table (HRs)
If you want a quick HR table for the combined model:

```{r cox_zurich_hr_table, message=FALSE, warning=FALSE}
library(broom)

tidy(m2, exponentiate = TRUE, conf.int = TRUE) %>%
  dplyr::select(term, estimate, conf.low, conf.high, p.value)
```

```{r plot_boot_delta, fig.height=3.6, fig.width=5.5}
library(ggplot2)
df_bd <- data.frame(delta_c = boot_delta)

ggplot(df_bd, aes(x = delta_c)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(x = "Δ C-index (COMPERA+score − COMPERA)",
       y = "Count",
       title = "Bootstrap distribution of c-index improvement") +
  theme_classic(base_size = 12)

```

```{r}
cox.zph(m2)
```

## Suggested figures for COMPERA and mortality results

These chunks assume the objects from your final-part analysis already exist, especially:
- `clin_z`, `pred_z`, `df_surv_z`
- `m0`, `m1`, `m2`, `cindex_tbl`
- `output_dir_figs`

### 1) Final score across COMPERA risk strata

This is probably the clearest figure for the COMPERA-related result.
It shows whether the final score increases across clinical risk strata.

```{r score_by_compera_boxplot, fig.height=4.3, fig.width=5.8, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

score_by_compera <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::select(Barcode, COMPERA_4strata, COMPERA_number) %>%
  dplyr::left_join(pred_z %>% dplyr::select(Barcode, score_eta), by = "Barcode") %>%
  dplyr::mutate(
    COMPERA_4strata = factor(
      COMPERA_4strata,
      levels = c("low", "intermediate-low", "intermediate-high", "high")
    ),
    COMPERA_number = as.numeric(COMPERA_number)
  ) %>%
  dplyr::filter(!is.na(COMPERA_4strata), !is.na(COMPERA_number), !is.na(score_eta))

p_kw <- kruskal.test(score_eta ~ COMPERA_4strata, data = score_by_compera)$p.value
rho   <- suppressWarnings(cor.test(score_by_compera$COMPERA_number, score_by_compera$score_eta, method = "spearman"))

cols_compera <- c(
  "low"               = "#9CA3AF",
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)

ggplot(score_by_compera, aes(x = COMPERA_4strata, y = score_eta, fill = COMPERA_4strata)) +
  geom_boxplot(alpha = 0.75, outlier.shape = NA) +
  geom_jitter(width = 0.12, size = 2, alpha = 0.75) +
  scale_fill_manual(values = cols_compera) +
  annotate(
    "text",
    x = 1,
    y = max(score_by_compera$score_eta, na.rm = TRUE),
    hjust = 0,
    vjust = 1.2,
    size = 4,
    label = paste0(
      "Kruskal-Wallis p = ", signif(p_kw, 3),
      "\nSpearman rho = ", round(unname(rho$estimate), 2),
      ", p = ", signif(rho$p.value, 3)
    )
  ) +
  labs(
    title = "Final score across COMPERA risk strata",
    x = "COMPERA risk stratum",
    y = "Final score (linear predictor)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```


### 2) Trend plot across COMPERA strata

This is a nice companion to the boxplot if you want to emphasize monotonic increase.

```{r score_by_compera_trend, fig.height=3.8, fig.width=4.6, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

trend_df <- score_by_compera %>%
  dplyr::group_by(COMPERA_number, COMPERA_4strata) %>%
  dplyr::summarise(
    mean_score = mean(score_eta, na.rm = TRUE),
    se_score   = sd(score_eta, na.rm = TRUE) / sqrt(dplyr::n()),
    n = dplyr::n(),
    .groups = "drop"
  )

ggplot(trend_df, aes(x = COMPERA_number, y = mean_score)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.6) +
  geom_errorbar(aes(ymin = mean_score - se_score, ymax = mean_score + se_score), width = 0.08) +
  scale_x_continuous(breaks = trend_df$COMPERA_number, labels = trend_df$COMPERA_4strata) +
  labs(
    title = "Final score increases with COMPERA risk",
    x = "COMPERA risk stratum",
    y = "Mean final score ± SE"
  ) +
  theme_classic(base_size = 12)
```


### 3) Forest plot for Cox regression results

Yes: for the mortality results, a forest plot is one of the best ways to show the Cox model output.
This chunk shows HRs for:
- COMPERA alone
- score alone
- combined model

```{r cox_forest_plot, fig.height=4.6, fig.width=6.6, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(broom)

forest_df <- dplyr::bind_rows(
  broom::tidy(m0, exponentiate = TRUE, conf.int = TRUE) %>% dplyr::mutate(model = "COMPERA only"),
  broom::tidy(m1, exponentiate = TRUE, conf.int = TRUE) %>% dplyr::mutate(model = "Score only"),
  broom::tidy(m2, exponentiate = TRUE, conf.int = TRUE) %>% dplyr::mutate(model = "COMPERA + score")
) %>%
  dplyr::filter(term %in% c("COMPERA_number", "eta")) %>%
  dplyr::mutate(
    term_label = dplyr::case_when(
      term == "COMPERA_number" & model == "COMPERA only"     ~ "COMPERA",
      term == "eta"            & model == "Score only"       ~ "Score",
      term == "COMPERA_number" & model == "COMPERA + score" ~ "COMPERA (adjusted for score)",
      term == "eta"            & model == "COMPERA + score" ~ "Score (adjusted for COMPERA)",
      TRUE ~ term
    ),
    term_label = factor(
      term_label,
      levels = rev(c(
        "Score (adjusted for COMPERA)",
        "COMPERA (adjusted for score)",
        "Score",
        "COMPERA"
      ))
    )
  )

ggplot(forest_df, aes(x = estimate, y = term_label)) +
  geom_vline(xintercept = 1, linetype = 2, color = "grey60") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.18, linewidth = 0.6) +
  geom_point(size = 2.5) +
  scale_x_log10() +
  labs(
    title = "Cox regression for mortality",
    x = "Hazard ratio (log scale)",
    y = NULL
  ) +
  theme_classic(base_size = 12)
```


### 4) Kaplan-Meier plot by final score group

This is the other figure I would definitely include for mortality.
Use a median split to avoid overfitting and to keep the figure simple.

```{r km_by_score_group, fig.height=4.5, fig.width=5.6, message=FALSE, warning=FALSE}
library(dplyr)
library(survival)

km_df <- df_surv_z %>%
  dplyr::mutate(
    score_group = ifelse(eta >= median(eta, na.rm = TRUE), "High score", "Low score"),
    score_group = factor(score_group, levels = c("Low score", "High score"))
  )

fit_km <- survival::survfit(Surv(time_days, event) ~ score_group, data = km_df)
logrank_p <- 1 - pchisq(survival::survdiff(Surv(time_days, event) ~ score_group, data = km_df)$chisq, df = 1)

plot(
  fit_km,
  col = c("#004C8C", "#A63A2A"),
  lwd = 2,
  xlab = "Follow-up (days)",
  ylab = "Survival probability",
  main = "Kaplan-Meier survival by final score"
)
legend(
  "bottomleft",
  legend = c(
    paste0("Low score (n=", sum(km_df$score_group == "Low score"), ")"),
    paste0("High score (n=", sum(km_df$score_group == "High score"), ")")
  ),
  col = c("#004C8C", "#A63A2A"),
  lwd = 2,
  bty = "n"
)
mtext(paste0("Log-rank p = ", signif(logrank_p, 3)), side = 3, adj = 1, line = 0.2, cex = 0.9)
```


### 5) C-index comparison plot

This is better than leaving the c-index only as a printed table.

```{r cindex_dotplot, fig.height=3.8, fig.width=5.2, message=FALSE, warning=FALSE}
library(ggplot2)

cindex_tbl$model <- factor(cindex_tbl$model, levels = c("COMPERA", "Score", "COMPERA + score"))

ggplot(cindex_tbl, aes(x = c_index, y = model)) +
  geom_point(size = 3) +
  geom_text(aes(label = sprintf("%.2f", c_index)), nudge_x = 0.015, size = 4) +
  xlim(0.5, 1.0) +
  labs(
    title = "Model discrimination",
    x = "C-index",
    y = NULL
  ) +
  theme_classic(base_size = 12)
```


### 6) Optional: bootstrap distribution of C-index improvement

This is useful if you want to support the claim that adding the score improves prediction over COMPERA.

```{r delta_cindex_bootstrap_plot, fig.height=3.8, fig.width=5.5, message=FALSE, warning=FALSE}
library(ggplot2)

df_bd <- data.frame(delta_c = boot_delta)
df_bd <- df_bd[is.finite(df_bd$delta_c), , drop = FALSE]

ggplot(df_bd, aes(x = delta_c)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(
    title = "Bootstrap distribution of C-index improvement",
    x = expression(Delta*" C-index (COMPERA + score minus COMPERA)"),
    y = "Count"
  ) +
  theme_classic(base_size = 12)
```


### 7) Optional: save the important figures

```{r save_compera_mortality_figures, eval=FALSE}
ggsave(here::here(output_dir_figs, "Score_by_COMPERA_boxplot.png"), width = 5.8, height = 4.3, dpi = 300)
ggsave(here::here(output_dir_figs, "Score_by_COMPERA_trend.png"), width = 4.6, height = 3.8, dpi = 300)
ggsave(here::here(output_dir_figs, "Cox_forest_plot.png"), width = 6.6, height = 4.6, dpi = 300)
ggsave(here::here(output_dir_figs, "Cindex_dotplot.png"), width = 5.2, height = 3.8, dpi = 300)
ggsave(here::here(output_dir_figs, "Delta_Cindex_bootstrap.png"), width = 5.5, height = 3.8, dpi = 300)

png(here::here(output_dir_figs, "KM_by_score_group.png"), width = 5.6, height = 4.5, units = "in", res = 300)
plot(
  fit_km,
  col = c("#004C8C", "#A63A2A"),
  lwd = 2,
  xlab = "Follow-up (days)",
  ylab = "Survival probability",
  main = "Kaplan-Meier survival by final score"
)
legend(
  "bottomleft",
  legend = c(
    paste0("Low score (n=", sum(km_df$score_group == "Low score"), ")"),
    paste0("High score (n=", sum(km_df$score_group == "High score"), ")")
  ),
  col = c("#004C8C", "#A63A2A"),
  lwd = 2,
  bty = "n"
)
mtext(paste0("Log-rank p = ", signif(logrank_p, 3)), side = 3, adj = 1, line = 0.2, cex = 0.9)
dev.off()
```

## Additional figures: Poland COMPERA score and cohort severity comparison

These chunks assume that the main analysis has already been run and that the following objects already exist:
- `clin_z`, `clin_p`
- `pred_z`, `pred_p`
- `output_dir_figs`

They add:
1. Final score across COMPERA strata in the Poland cohort
2. Final score across COMPERA strata in both cohorts (optional combined figure)
3. TAPSE/sPAP comparison between Zurich and Poland in SSc-PAH patients
4. COMPERA group counts in the two cohorts shown as side-by-side barplots

```{r helper_compera_labels, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)

standardize_compera <- function(x) {
  x <- as.character(x)
  x <- trimws(tolower(x))
  dplyr::case_when(
    x %in% c("low") ~ "low",
    x %in% c("intermediate-low", "intermediate low", "intermediate_low") ~ "intermediate-low",
    x %in% c("intermediate-high", "intermediate high", "intermediate_high") ~ "intermediate-high",
    x %in% c("high") ~ "high",
    TRUE ~ NA_character_
  )
}

compera_levels <- c("low", "intermediate-low", "intermediate-high", "high")
compera_labels <- c("low", "intermediate-low", "intermediate-high", "high")
compera_cols <- c(
  "low"               = "#9CA3AF",
  "intermediate-low"  = "#DDB7B2",
  "intermediate-high" = "#C7685F",
  "high"              = "#A63A2A"
)
cohort_cols <- c("Zurich" = "#004C8C", "Poland" = "#A63A2A")
```

### 1) Poland cohort: final score across COMPERA strata

```{r poland_score_by_compera, fig.height=4.4, fig.width=5.8, message=FALSE, warning=FALSE}
score_poland_compera <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::select(Barcode, Compera_4strata, Compera_number) %>%
  dplyr::left_join(pred_p %>% dplyr::select(Barcode, score_eta), by = "Barcode") %>%
  dplyr::mutate(
    COMPERA = standardize_compera(Compera_4strata),
    COMPERA = factor(COMPERA, levels = compera_levels),
    Compera_number = as.numeric(Compera_number)
  ) %>%
  dplyr::filter(!is.na(COMPERA), !is.na(Compera_number), !is.na(score_eta))

p_kw_pol <- kruskal.test(score_eta ~ COMPERA, data = score_poland_compera)$p.value
rho_pol  <- suppressWarnings(cor.test(score_poland_compera$Compera_number,
                                      score_poland_compera$score_eta,
                                      method = "spearman"))

ggplot(score_poland_compera, aes(x = COMPERA, y = score_eta, fill = COMPERA)) +
  geom_boxplot(alpha = 0.75, outlier.shape = NA) +
  geom_jitter(width = 0.12, size = 2, alpha = 0.75) +
  scale_fill_manual(values = compera_cols, drop = FALSE) +
  annotate(
    "text",
    x = 1,
    y = max(score_poland_compera$score_eta, na.rm = TRUE),
    hjust = 0,
    vjust = 1.2,
    size = 4,
    label = paste0(
      "Kruskal-Wallis p = ", signif(p_kw_pol, 3),
      "\nSpearman rho = ", round(unname(rho_pol$estimate), 2),
      ", p = ", signif(rho_pol$p.value, 3)
    )
  ) +
  labs(
    title = "Poland cohort: final score across COMPERA risk strata",
    x = "COMPERA risk stratum",
    y = "Final score (linear predictor)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```

### 2) Optional combined figure: final score across COMPERA strata in both cohorts

```{r both_cohorts_score_by_compera, fig.height=4.6, fig.width=8.5, message=FALSE, warning=FALSE}
score_zurich_compera <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::select(Barcode, COMPERA_4strata, COMPERA_number) %>%
  dplyr::left_join(pred_z %>% dplyr::select(Barcode, score_eta), by = "Barcode") %>%
  dplyr::mutate(
    COMPERA = standardize_compera(COMPERA_4strata),
    COMPERA = factor(COMPERA, levels = compera_levels),
    COMPERA_number = as.numeric(COMPERA_number),
    Cohort = "Zurich"
  ) %>%
  dplyr::filter(!is.na(COMPERA), !is.na(COMPERA_number), !is.na(score_eta))

score_poland_compera2 <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::select(Barcode, Compera_4strata, Compera_number) %>%
  dplyr::left_join(pred_p %>% dplyr::select(Barcode, score_eta), by = "Barcode") %>%
  dplyr::mutate(
    COMPERA = standardize_compera(Compera_4strata),
    COMPERA = factor(COMPERA, levels = compera_levels),
    Compera_number = as.numeric(Compera_number),
    Cohort = "Poland"
  ) %>%
  dplyr::rename(COMPERA_number = Compera_number) %>%
  dplyr::filter(!is.na(COMPERA), !is.na(COMPERA_number), !is.na(score_eta))

score_both_compera <- dplyr::bind_rows(score_zurich_compera, score_poland_compera2)

ggplot(score_both_compera, aes(x = COMPERA, y = score_eta, fill = COMPERA)) +
  geom_boxplot(alpha = 0.75, outlier.shape = NA) +
  geom_jitter(width = 0.12, size = 1.8, alpha = 0.75) +
  facet_wrap(~Cohort) +
  scale_fill_manual(values = compera_cols, drop = FALSE) +
  labs(
    title = "Final score across COMPERA risk strata",
    x = "COMPERA risk stratum",
    y = "Final score (linear predictor)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```

### 3) TAPSE/sPAP comparison between cohorts in SSc-PAH patients

```{r tapse_spap_boxplot_by_cohort, fig.height=4.5, fig.width=5.8, message=FALSE, warning=FALSE}
tapse_spap_compare <- dplyr::bind_rows(
  clin_z %>%
    dplyr::filter(Group == "SSc-PAH") %>%
    dplyr::transmute(
      Barcode,
      TAPSE_sPAP = as.numeric(TAPSE_sPAP),
      Cohort = "Zurich"
    ),
  clin_p %>%
    dplyr::filter(Group == "SSc-PAH") %>%
    dplyr::transmute(
      Barcode,
      TAPSE_sPAP = as.numeric(TAPSE_sPAP),
      Cohort = "Poland"
    )
) %>%
  dplyr::filter(!is.na(TAPSE_sPAP))

p_tapse <- wilcox.test(TAPSE_sPAP ~ Cohort, data = tapse_spap_compare)$p.value

ggplot(tapse_spap_compare, aes(x = Cohort, y = TAPSE_sPAP, fill = Cohort)) +
  geom_hline(yintercept = 0.32, linetype = 2, color = "black", linewidth = 0.6) +
  geom_boxplot(alpha = 0.75, outlier.shape = NA) +
  geom_jitter(width = 0.12, size = 2, alpha = 0.75) +
  scale_fill_manual(values = cohort_cols) +
  annotate(
    "text",
    x = 1.5,
    y = max(tapse_spap_compare$TAPSE_sPAP, na.rm = TRUE),
    label = paste0("Wilcoxon p = ", signif(p_tapse, 3)),
    vjust = 1.2,
    size = 4
  ) +
  labs(
    title = "TAPSE/sPAP in SSc-PAH patients",
    x = NULL,
    y = "TAPSE/sPAP"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```

### 4) COMPERA group distribution by cohort (counts)

This gives two barplots next to each other, one for Zurich and one for Poland, with patient counts per risk group.

```{r compera_counts_by_cohort, fig.height=4.5, fig.width=8, message=FALSE, warning=FALSE}
compera_counts <- dplyr::bind_rows(
  clin_z %>%
    dplyr::filter(Group == "SSc-PAH") %>%
    dplyr::transmute(
      Cohort = "Zurich",
      COMPERA = standardize_compera(COMPERA_4strata)
    ),
  clin_p %>%
    dplyr::filter(Group == "SSc-PAH") %>%
    dplyr::transmute(
      Cohort = "Poland",
      COMPERA = standardize_compera(Compera_4strata)
    )
) %>%
  dplyr::filter(!is.na(COMPERA)) %>%
  dplyr::mutate(COMPERA = factor(COMPERA, levels = compera_levels)) %>%
  dplyr::count(Cohort, COMPERA, .drop = FALSE)

ggplot(compera_counts, aes(x = COMPERA, y = n, fill = COMPERA)) +
  geom_col(width = 0.8) +
  geom_text(aes(label = n), vjust = -0.25, size = 4) +
  facet_wrap(~Cohort, nrow = 1) +
  scale_fill_manual(values = compera_cols, drop = FALSE) +
  labs(
    title = "COMPERA risk-group distribution in SSc-PAH patients",
    x = "COMPERA risk stratum",
    y = "Number of patients"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
  )
```

### 5) Optional: percentages instead of counts

If you also want a cleaner comparison independent of cohort size, this percentage version can be useful in supplement.

```{r compera_percent_by_cohort, fig.height=4.5, fig.width=8, message=FALSE, warning=FALSE}
compera_props <- compera_counts %>%
  dplyr::group_by(Cohort) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(compera_props, aes(x = COMPERA, y = prop, fill = COMPERA)) +
  geom_col(width = 0.8) +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), vjust = -0.25, size = 4) +
  facet_wrap(~Cohort, nrow = 1) +
  scale_fill_manual(values = compera_cols, drop = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1.05)) +
  labs(
    title = "COMPERA risk-group proportions in SSc-PAH patients",
    x = "COMPERA risk stratum",
    y = "Proportion of patients"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```

### 6) Optional save commands

```{r save_poland_and_cohort_comparison_figures, eval=FALSE}
ggsave(here::here(output_dir_figs, "Poland_score_by_COMPERA.png"), width = 5.8, height = 4.4, dpi = 300)
ggsave(here::here(output_dir_figs, "Both_cohorts_score_by_COMPERA.png"), width = 8.5, height = 4.6, dpi = 300)
ggsave(here::here(output_dir_figs, "TAPSE_sPAP_by_cohort_PAH.png"), width = 5.8, height = 4.5, dpi = 300)
ggsave(here::here(output_dir_figs, "COMPERA_counts_by_cohort.png"), width = 8, height = 4.5, dpi = 300)
ggsave(here::here(output_dir_figs, "COMPERA_percent_by_cohort.png"), width = 8, height = 4.5, dpi = 300)
```

```{r poland_score_by_death_status, fig.height=4.2, fig.width=4.8, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

death_col_p <- "Death (no/yes; 0/1)"

score_poland_death <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::select(Barcode, all_of(death_col_p)) %>%
  dplyr::left_join(pred_p %>% dplyr::select(Barcode, score_eta), by = "Barcode") %>%
  dplyr::mutate(
    death = as.numeric(.data[[death_col_p]]),
    death_status = dplyr::case_when(
      death == 0 ~ "No",
      death == 1 ~ "Yes",
      TRUE ~ NA_character_
    ),
    death_status = factor(death_status, levels = c("No", "Yes"))
  ) %>%
  dplyr::filter(!is.na(death_status), !is.na(score_eta))

score_poland_death %>%
  dplyr::count(death_status)

p_death_pol <- wilcox.test(score_eta ~ death_status, data = score_poland_death)$p.value

ggplot(score_poland_death, aes(x = death_status, y = score_eta, fill = death_status)) +
  geom_boxplot(alpha = 0.75, outlier.shape = NA) +
  geom_jitter(width = 0.12, size = 2, alpha = 0.75) +
  scale_fill_manual(values = c("No" = "grey75", "Yes" = "#A63A2A")) +
  annotate(
    "text",
    x = 1.5,
    y = max(score_poland_death$score_eta, na.rm = TRUE),
    label = paste0("Wilcoxon p = ", signif(p_death_pol, 3)),
    vjust = 1.2,
    size = 4
  ) +
  labs(
    title = "Poland cohort: final score by death status",
    subtitle = "SSc-PAH patients only",
    x = "Death",
    y = "Final score (linear predictor)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```
