---
title: "SSc-PAH serum proteomics"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
current_file <- "SSc_PAH_serum_proteomics"

library(here)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(scales)
library(patchwork)
library(qs2)
library(SummarizedExperiment)
library(limma)
library(EnhancedVolcano)
library(glmnet)
library(pROC)
library(pheatmap)
library(survival)
library(broom)
library(grid)

output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"), recursive = TRUE)

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs, recursive = TRUE)

# Main group colors
cols_group <- c("SSc-noPAH" = "#2C6E9B", "SSc-PAH" = "#B23A2B")
cols_cohort <- c("Zurich" = "#2C6E9B", "Krakow" = "#B23A2B")
cols_tapse <- c("Preserved" = "#7FA6BB", "Impaired" = "#B23A2B")
cols_three <- c(
  "SSc-noPAH" = "#BDBDBD",
  "SSc-PAH\nPreserved coupling" = "#7FA6BB",
  "SSc-PAH\nImpaired coupling" = "#B23A2B"
)

# Sequential red scale for COMPERA
cols_compera <- c(
  "low" = "#E6B2AB",
  "intermediate-low" = "#D8877C",
  "intermediate-high" = "#C65A4C",
  "high" = "#942C22"
)

coupling_labels <- c(
  "Preserved" = "Preserved coupling",
  "Impaired" = "Impaired coupling"
)

coupling_axis_labels <- c(
  "Preserved" = "Preserved\ncoupling",
  "Impaired" = "Impaired\ncoupling"
)

standardize_compera <- function(x){
  x <- tolower(trimws(as.character(x)))
  dplyr::case_when(
    x == "low" ~ "low",
    x %in% c("intermediate-low", "intermediate low", "intermediate_low") ~ "intermediate-low",
    x %in% c("intermediate-high", "intermediate high", "intermediate_high") ~ "intermediate-high",
    x == "high" ~ "high",
    TRUE ~ NA_character_
  )
}

# ROC helpers --------------------------------------------------------
roc_curve_df <- function(y, score){
  ok <- is.finite(y) & is.finite(score)
  y <- y[ok]
  score <- score[ok]

  rr <- pROC::roc(
    y, score,
    levels = c(0, 1),
    direction = "<",
    quiet = TRUE
  )

  data.frame(
    FPR = 1 - rr$specificities,
    Sensitivity = rr$sensitivities
  ) %>%
    arrange(FPR, Sensitivity)
}

roc_bootstrap_band <- function(y, score, B = 2000, seed = 1,
                               grid = seq(0, 1, length.out = 101)){
  ok <- is.finite(y) & is.finite(score)
  y <- y[ok]
  score <- score[ok]

  set.seed(seed)
  sens_mat <- matrix(NA_real_, nrow = B, ncol = length(grid))

  for (b in seq_len(B)){
    idx <- sample(seq_along(y), replace = TRUE)
    yb <- y[idx]
    sb <- score[idx]

    if (length(unique(yb)) < 2) next

    dd <- roc_curve_df(yb, sb) %>%
      group_by(FPR) %>%
      summarise(Sensitivity = max(Sensitivity), .groups = "drop")

    dd <- bind_rows(
      data.frame(FPR = 0, Sensitivity = 0),
      dd,
      data.frame(FPR = 1, Sensitivity = 1)
    ) %>%
      group_by(FPR) %>%
      summarise(Sensitivity = max(Sensitivity), .groups = "drop") %>%
      arrange(FPR)

    sens_mat[b, ] <- approx(
      x = dd$FPR,
      y = dd$Sensitivity,
      xout = grid,
      ties = max,
      rule = 2
    )$y
  }

  data.frame(
    FPR = grid,
    lower = apply(sens_mat, 2, quantile, 0.025, na.rm = TRUE),
    median = apply(sens_mat, 2, quantile, 0.50, na.rm = TRUE),
    upper = apply(sens_mat, 2, quantile, 0.975, na.rm = TRUE)
  )
}
```

## Import data

```{r read_data}
data_z_se <- qs_read(here::here("data", "o37198_2025-02-27_olink_SummarizedExperiment.qs2"))
clin_z <- read_excel(here::here("data", "Metadata_Olink_analysis.xlsx"), sheet = "Merged")

meta_z_olink <- colData(data_z_se) %>% data.frame()
isOutlierColumn <- meta_z_olink$Plate == "P0194" & meta_z_olink$Col %in% c("3", "4")
use <- meta_z_olink$SampleType %in% "Serum" & meta_z_olink$Group %in% c("SSc-PAH", "SSc-noPAH") & !isOutlierColumn

data_z_se <- data_z_se[, use]
meta_z_olink <- meta_z_olink[use, ]
rownames(meta_z_olink) <- gsub("_P0194", "", rownames(meta_z_olink))
meta_z_olink$Barcode <- rownames(meta_z_olink)

rownames(data_z_se) <- data_z_se@elementMetadata@listData[["Gene"]]
rownames(data_z_se@assays@data@listData[["npx"]]) <- data_z_se@elementMetadata@listData[["Gene"]]
data_z <- assay(data_z_se, "npx")
data_z <- as.data.frame(data_z[-c(1:3), ])
colnames(data_z) <- rownames(meta_z_olink)

clin_z <- clin_z %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  dplyr::distinct(Barcode, .keep_all = TRUE)

data_p <- read.delim(here::here("data", "OLINK_PL", "quant.xls"), sep = "\t", check.names = FALSE)
clin_p <- read_excel(here::here("data", "OLINK_PL", "Metadata_Olink_PL.xlsx"), sheet = "Metadata")

rownames(data_p) <- sub("_.*", "", data_p$SampleID)
data_p <- data_p[, -1]
keep <- intersect(colnames(data_p), clin_p$Barcode)
data_p <- data_p[, keep, drop = FALSE]
clin_p <- clin_p %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  dplyr::distinct(Barcode, .keep_all = TRUE)

cat("Zurich Olink matrix:", dim(data_z), "\n")
print(table(clin_z$Group))
cat("Krakow Olink matrix:", dim(data_p), "\n")
print(table(clin_p$Group))
cat("Missing NPX values Zurich:", sum(is.na(data_z)), "\n")
cat("Missing NPX values Krakow:", sum(is.na(data_p)), "\n")
cat("Duplicated protein names Zurich:", anyDuplicated(rownames(data_z)), "\n")
cat("Duplicated protein names Krakow:", anyDuplicated(rownames(data_p)), "\n")
```

## Harmonize metadata

```{r harmonize_metadata}
if ("DETECT_step2 " %in% colnames(clin_z)) names(clin_z)[names(clin_z) == "DETECT_step2 "] <- "DETECT_step2"

clin_z <- clin_z %>%
  mutate(
    Cohort = "Zurich",
    Age = as.numeric(Age), Sex = factor(Sex), TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    COMPERA = standardize_compera(COMPERA_4strata), COMPERA_number_common = as.numeric(COMPERA_number),
    SixMWD = as.numeric(`6_Minute_walk_test_distance.in.m`), NTproBNP_common = as.numeric(NTproBNP_pg.ml),
    WHO_FC = as.numeric(`NYHA class`), TAPSE = as.numeric(TAPSE_mm), sPAP = as.numeric(sPAP_ECHO),
    mPAP = as.numeric(mPAP_mmHg), PCWP = as.numeric(Pulmonary_wedge_pressure_mmHg), PVR = as.numeric(PVR_WU),
    CI = as.numeric(Cardiac_Index_L.min.m2), Creatinine_mg_dl = as.numeric(Creatinine),
    ILD_common = tolower(trimws(as.character(ILDonHRCT))), FVC_pred_common = as.numeric(FVC),
    DLCO_pred_common = as.numeric(Dlco), mRSS_common = as.numeric(mRSS), DETECT_score = as.numeric(DETECT_step2)
  )

clin_p <- clin_p %>%
  mutate(
    Cohort = "Krakow",
    Age = as.numeric(Age), Sex = factor(Sex), TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    COMPERA = standardize_compera(Compera_4strata), COMPERA_number_common = as.numeric(Compera_number),
    SixMWD = as.numeric(`6MWT`), NTproBNP_common = as.numeric(NTproBNP), WHO_FC = as.numeric(NYHA),
    TAPSE = as.numeric(TAPSE),
    # Krakow TAPSE/sPAP is calculated using RVSP + RAP.
    sPAP = as.numeric(RVSP) + as.numeric(RAP),
    mPAP = as.numeric(mPAP), PCWP = as.numeric(PCWP), PVR = as.numeric(PVR_WU), CI = as.numeric(CI),
    Creatinine_mg_dl = as.numeric(Creatinine_mg_dL), ILD_common = tolower(trimws(as.character(ILD))),
    FVC_pred_common = as.numeric(FVC_pred), DLCO_pred_common = as.numeric(DLCO_pred), mRSS_common = as.numeric(mRSS)
  )

clin_z$COMPERA <- factor(clin_z$COMPERA, levels = c("low", "intermediate-low", "intermediate-high", "high"))
clin_p$COMPERA <- factor(clin_p$COMPERA, levels = c("low", "intermediate-low", "intermediate-high", "high"))

# S319 TAPSE was entered as 181 instead of 18.1 mm
clin_p$TAPSE[clin_p$Barcode == "S319"] <- 18.1
clin_p$TAPSE_sPAP[clin_p$Barcode == "S319"] <- 18.1 / clin_p$sPAP[clin_p$Barcode == "S319"]

common_proteins <- intersect(rownames(data_p), rownames(data_z))
cat("Proteins measured in both cohorts:", length(common_proteins), "\n")
```

## Define analysis populations

```{r analysis_populations}
z_all <- clin_z %>% filter(Barcode %in% colnames(data_z))
p_all <- clin_p %>% filter(Barcode %in% colnames(data_p))

z_pah <- z_all %>% filter(Group == "SSc-PAH")
z_nopah <- z_all %>% filter(Group == "SSc-noPAH")
p_pah <- p_all %>% filter(Group == "SSc-PAH")
p_nopah <- p_all %>% filter(Group == "SSc-noPAH")

z_pah_tapse <- z_pah %>% filter(!is.na(TAPSE_sPAP)) %>% mutate(TAPSE_group = ifelse(TAPSE_sPAP <= 0.32, "Impaired", "Preserved"))
p_pah_tapse <- p_pah %>% filter(!is.na(TAPSE_sPAP)) %>% mutate(TAPSE_group = ifelse(TAPSE_sPAP <= 0.32, "Impaired", "Preserved"))

z_pah_tapse$TAPSE_group <- factor(z_pah_tapse$TAPSE_group, levels = c("Preserved", "Impaired"))
p_pah_tapse$TAPSE_group <- factor(p_pah_tapse$TAPSE_group, levels = c("Preserved", "Impaired"))

cat("Zurich SSc-PAH with TAPSE/sPAP:\n")
print(table(z_pah_tapse$TAPSE_group))
cat("Krakow SSc-PAH with TAPSE/sPAP:\n")
print(table(p_pah_tapse$TAPSE_group))
```

# Cohort characteristics and QC

## Table 1

```{r table1}
fmt_cont <- function(x, digits = 1){
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (length(x) == 0) return("-")
  paste0(formatC(median(x), digits = digits, format = "f"), " [",
         formatC(quantile(x, 0.25), digits = digits, format = "f"), "–",
         formatC(quantile(x, 0.75), digits = digits, format = "f"), "], n=", length(x))
}

fmt_binary <- function(x, positive = "yes"){
  x <- tolower(trimws(as.character(x)))
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) return("-")
  npos <- sum(x %in% positive)
  paste0(npos, " (", round(100 * npos / length(x), 1), "%), n=", length(x))
}

fmt_female <- function(x){
  x <- toupper(trimws(as.character(x)))
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) return("-")
  npos <- sum(x == "F")
  paste0(npos, " (", round(100 * npos / length(x), 1), "%), n=", length(x))
}

fmt_compera <- function(x, level){
  x <- as.character(x)
  x <- x[!is.na(x)]
  if (length(x) == 0) return("-")
  nlevel <- sum(x == level)
  paste0(nlevel, " (", round(100 * nlevel / length(x), 1), "%), n=", length(x))
}

make_table1_column <- function(df){
  c(
    fmt_cont(df$Age, 0),
    fmt_female(df$Sex),
    fmt_cont(df$BMI, 1),
    fmt_binary(df$ILD_common, "yes"),
    fmt_cont(df$FVC_pred_common, 0),
    fmt_cont(df$DLCO_pred_common, 0),
    fmt_cont(df$Creatinine_mg_dl, 2),
    fmt_cont(df$WHO_FC, 0),
    fmt_cont(df$SixMWD, 0),
    fmt_cont(df$NTproBNP_common, 0),
    fmt_cont(df$TAPSE, 1),
    fmt_cont(df$sPAP, 1),
    fmt_cont(df$TAPSE_sPAP, 2),
    fmt_cont(df$mPAP, 1),
    fmt_cont(df$PCWP, 1),
    fmt_cont(df$PVR, 1),
    fmt_cont(df$CI, 2),
    fmt_compera(df$COMPERA, "low"),
    fmt_compera(df$COMPERA, "intermediate-low"),
    fmt_compera(df$COMPERA, "intermediate-high"),
    fmt_compera(df$COMPERA, "high")
  )
}

table1 <- data.frame(
  Characteristic = c(
    "Age, years", "Female, n (%)", "BMI, kg/m2", "ILD, n (%)", "FVC, % predicted", "DLCO, % predicted",
    "Creatinine, mg/dL", "WHO functional class", "6-minute walk distance, m", "NT-proBNP, pg/mL",
    "TAPSE, mm", "sPAP, mmHg", "TAPSE/sPAP", "mPAP, mmHg", "PCWP, mmHg", "PVR, WU", "Cardiac index, L/min/m2",
    "COMPERA low", "COMPERA intermediate-low", "COMPERA intermediate-high", "COMPERA high"
  ),
  Zurich_noPAH = make_table1_column(z_nopah),
  Zurich_PAH = make_table1_column(z_pah),
  Krakow_noPAH = make_table1_column(p_nopah),
  Krakow_PAH = make_table1_column(p_pah),
  check.names = FALSE
)

colnames(table1)[2:5] <- c(
  paste0("Zurich SSc-noPAH (n=", nrow(z_nopah), ")"),
  paste0("Zurich SSc-PAH (n=", nrow(z_pah), ")"),
  paste0("Krakow SSc-noPAH (n=", nrow(p_nopah), ")"),
  paste0("Krakow SSc-PAH (n=", nrow(p_pah), ")")
)

write.csv(table1, here::here(output_dir_data, "Table1_clinical_characteristics.csv"), row.names = FALSE)
knitr::kable(table1, align = c("l", "c", "c", "c", "c"))
```

## NPX distributions

```{r npx_distribution_zurich, fig.height=4, fig.width=10, dpi=300}
order_z <- z_all %>% arrange(Group, Barcode) %>% pull(Barcode)

npx_long_z <- data_z %>%
  as.data.frame() %>%
  rownames_to_column("Protein") %>%
  pivot_longer(-Protein, names_to = "Barcode", values_to = "NPX") %>%
  left_join(z_all %>% select(Barcode, Group), by = "Barcode") %>%
  mutate(Barcode = factor(Barcode, levels = order_z))

p_npx_z <- ggplot(npx_long_z, aes(Barcode, NPX, fill = Group)) +
  geom_boxplot(outlier.shape = NA, linewidth = 0.25) +
  scale_fill_manual(values = cols_group) +
  labs(title = "Zurich cohort", x = NULL, y = "NPX") +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5), legend.position = "top")

p_npx_z
ggsave(here::here(output_dir_figs, "SupplementaryFigure1A_NPX_Zurich.png"), p_npx_z, width = 11, height = 5.2, dpi = 300)
```

```{r npx_distribution_poland, fig.height=4, fig.width=10, dpi=300}
order_p <- p_all %>% arrange(Group, Barcode) %>% pull(Barcode)

npx_long_p <- data_p %>%
  as.data.frame() %>%
  rownames_to_column("Protein") %>%
  pivot_longer(-Protein, names_to = "Barcode", values_to = "NPX") %>%
  left_join(p_all %>% select(Barcode, Group), by = "Barcode") %>%
  mutate(Barcode = factor(Barcode, levels = order_p))

p_npx_p <- ggplot(npx_long_p, aes(Barcode, NPX, fill = Group)) +
  geom_boxplot(outlier.shape = NA, linewidth = 0.25) +
  scale_fill_manual(values = cols_group) +
  labs(title = "Krakow cohort", x = NULL, y = "NPX") +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5), legend.position = "top")

p_npx_p
ggsave(here::here(output_dir_figs, "SupplementaryFigure1B_NPX_Krakow.png"), p_npx_p, width = 11, height = 5.2, dpi = 300)
```

## PCA

```{r pca_zurich, fig.height=4.2, fig.width=5.2, dpi=300}
sd_pca_z <- apply(data_z, 1, sd, na.rm = TRUE)
data_z_pca <- data_z[is.finite(sd_pca_z) & sd_pca_z > 0, , drop = FALSE]
pca_z <- prcomp(t(as.matrix(data_z_pca)), center = TRUE, scale. = TRUE)
var_z <- 100 * pca_z$sdev^2 / sum(pca_z$sdev^2)
pca_z_df <- data.frame(Barcode = rownames(pca_z$x), PC1 = pca_z$x[, 1], PC2 = pca_z$x[, 2]) %>%
  left_join(z_all %>% select(Barcode, Group, COMPERA), by = "Barcode")

p_pca_z <- ggplot(pca_z_df, aes(PC1, PC2, color = Group)) +
  geom_point(size = 2.2, alpha = 0.85) +
  scale_color_manual(values = cols_group) +
  labs(title = "Zurich cohort", x = paste0("PC1 (", round(var_z[1], 1), "%)"), y = paste0("PC2 (", round(var_z[2], 1), "%)")) +
  theme_classic(base_size = 14)

p_pca_z
ggsave(here::here(output_dir_figs, "SupplementaryFigure1C_PCA_Zurich.png"), p_pca_z, width = 5.2, height = 4.2, dpi = 300)
```

```{r pca_poland, fig.height=4.2, fig.width=5.2, dpi=300}
sd_pca_p <- apply(data_p, 1, sd, na.rm = TRUE)
data_p_pca <- data_p[is.finite(sd_pca_p) & sd_pca_p > 0, , drop = FALSE]
pca_p <- prcomp(t(as.matrix(data_p_pca)), center = TRUE, scale. = TRUE)
var_p <- 100 * pca_p$sdev^2 / sum(pca_p$sdev^2)
pca_p_df <- data.frame(Barcode = rownames(pca_p$x), PC1 = pca_p$x[, 1], PC2 = pca_p$x[, 2]) %>%
  left_join(p_all %>% select(Barcode, Group, COMPERA), by = "Barcode")

p_pca_p <- ggplot(pca_p_df, aes(PC1, PC2, color = Group)) +
  geom_point(size = 2.2, alpha = 0.85) +
  scale_color_manual(values = cols_group) +
  labs(title = "Krakow cohort", x = paste0("PC1 (", round(var_p[1], 1), "%)"), y = paste0("PC2 (", round(var_p[2], 1), "%)")) +
  theme_classic(base_size = 14)

p_pca_p
ggsave(here::here(output_dir_figs, "SupplementaryFigure1D_PCA_Krakow.png"), p_pca_p, width = 5.2, height = 4.2, dpi = 300)
```

```{r supplementary_figure1_panel, fig.height=9, fig.width=12, dpi=300}
SupplementaryFigure1 <- (p_npx_z / p_npx_p) / (p_pca_z | p_pca_p) +
  plot_annotation(tag_levels = "A")

SupplementaryFigure1
ggsave(here::here(output_dir_figs, "SupplementaryFigure1_proteomics_QC.png"),
       SupplementaryFigure1, width = 12, height = 9, dpi = 300)
```






## Figure 1: study design and cohort overview

```{r figure1_data}
cohort_summary_fig1 <- bind_rows(
  p_all %>% mutate(Cohort = "Krakow"),
  z_all %>% mutate(Cohort = "Zurich")
) %>%
  group_by(Cohort) %>%
  summarise(
    total = n(),
    noPAH = sum(Group == "SSc-noPAH", na.rm = TRUE),
    PAH = sum(Group == "SSc-PAH", na.rm = TRUE),
    .groups = "drop"
  )

rvpa_plot_df <- bind_rows(
  p_all %>% transmute(Cohort = "Krakow", Group, TAPSE_sPAP),
  z_all %>% transmute(Cohort = "Zurich", Group, TAPSE_sPAP)
) %>%
  filter(!is.na(TAPSE_sPAP)) %>%
  mutate(
    Cohort = factor(Cohort, levels = c("Krakow", "Zurich")),
    RVPA_group = case_when(
      Group == "SSc-noPAH" ~ "SSc-noPAH",
      Group == "SSc-PAH" & TAPSE_sPAP > 0.32 ~ "SSc-PAH\nPreserved coupling",
      Group == "SSc-PAH" & TAPSE_sPAP <= 0.32 ~ "SSc-PAH\nImpaired coupling",
      TRUE ~ NA_character_
    ),
    RVPA_group = factor(
      RVPA_group,
      levels = c(
        "SSc-noPAH",
        "SSc-PAH\nPreserved coupling",
        "SSc-PAH\nImpaired coupling"
      )
    )
  ) %>%
  filter(!is.na(RVPA_group))

compera_counts <- bind_rows(
  z_pah %>% transmute(Cohort = "Zurich", COMPERA),
  p_pah %>% transmute(Cohort = "Krakow", COMPERA)
) %>%
  filter(!is.na(COMPERA)) %>%
  mutate(Cohort = factor(Cohort, levels = c("Krakow", "Zurich"))) %>%
  dplyr::count(Cohort, COMPERA, .drop = FALSE)
```

```{r figure1A_workflow, fig.height=4.5, fig.width=12, dpi=300}
n_k <- cohort_summary_fig1 %>% filter(Cohort == "Krakow")
n_z <- cohort_summary_fig1 %>% filter(Cohort == "Zurich")

workflow_boxes <- data.frame(
  x = c(0.9, 2.7, 4.8, 6.8, 8.8, 10.7, 4.8, 6.9),
  y = c(2.0, 2.0, 3.0, 3.0, 3.0, 3.0, 1.0, 1.0),
  label = c(
    paste0("Serum cohorts\nKrakow: n=", n_k$total, " (", n_k$noPAH, " noPAH; ", n_k$PAH, " PAH)\nZurich: n=", n_z$total, " (", n_z$noPAH, " noPAH; ", n_z$PAH, " PAH)"),
    "Olink REVEAL\n1,034 proteins",
    "Krakow development\nAll SSc with TAPSE/sPAP",
    "Top 50 by |t|\nLASSO -> 5 proteins",
    "Zurich SSc-PAH\nExternal validation",
    "COMPERA +\n3-year mortality",
    "Zurich DETECT-positive\nRHC: PAH vs noPAH",
    "DETECT vs proteomics\nNested-CV elastic net"
  )
)
workflow_arrows <- data.frame(
  x = c(1.6, 3.3, 5.5, 7.5, 9.5, 3.3, 5.5),
  xend = c(2.1, 4.0, 6.1, 8.1, 10.1, 4.0, 6.2),
  y = c(2.0, 2.35, 3.0, 3.0, 3.0, 1.65, 1.0),
  yend = c(2.0, 2.75, 3.0, 3.0, 3.0, 1.25, 1.0)
)
p_fig1A <- ggplot() +
  geom_label(data = workflow_boxes, aes(x, y, label = label), size = 4.0, fill = "white", label.size = 0.35, label.padding = unit(0.42, "lines")) +
  geom_segment(data = workflow_arrows, aes(x = x, y = y, xend = xend, yend = yend), arrow = grid::arrow(length = unit(0.12, "inches")), linewidth = 0.5) +
  xlim(0.0, 11.7) + ylim(0.25, 3.75) + theme_void()
p_fig1A
ggsave(here::here(output_dir_figs, "Figure1A_study_workflow_draft.png"), p_fig1A, width = 12, height = 4.5, dpi = 300)
```


```{r figure1C_rvpa_coupling, fig.height=5.0, fig.width=9.2, dpi=300}
p_fig1C <- ggplot(rvpa_plot_df, aes(RVPA_group, TAPSE_sPAP, fill = RVPA_group)) +
  geom_hline(yintercept = 0.32, linetype = 2, linewidth = 0.6) +
  geom_boxplot(outlier.shape = NA, width = 0.55, alpha = 0.78) +
  geom_jitter(width = 0.10, size = 2.1, alpha = 0.78) +
  facet_wrap(~Cohort, nrow = 1) +
  scale_fill_manual(values = cols_three) +
  labs(x = NULL, y = "TAPSE/sPAP", title = "RV-PA coupling") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none", strip.background = element_rect(fill = "white"), strip.text = element_text(size = 14), axis.text.x = element_text(size = 11.5))
p_fig1C
ggsave(here::here(output_dir_figs, "Figure1B_RVPA_coupling.png"), p_fig1C, width = 9.2, height = 5.0, dpi = 300)
```

```{r figure1D_compera, fig.height=5.0, fig.width=8.5, dpi=300}
p_fig1D <- ggplot(compera_counts, aes(COMPERA, n, fill = COMPERA)) +
  geom_col(width = 0.72) + geom_text(aes(label = n), vjust = -0.35, size = 4.3) +
  facet_wrap(~Cohort, nrow = 1) + scale_fill_manual(values = cols_compera, drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "COMPERA risk stratum", y = "Patients", title = "COMPERA 2.0 risk") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none", strip.background = element_rect(fill = "white"), strip.text = element_text(size = 14), axis.text.x = element_text(angle = 30, hjust = 1))
p_fig1D
ggsave(here::here(output_dir_figs, "Figure1C_COMPERA.png"), p_fig1D, width = 8.5, height = 5.0, dpi = 300)
```

```{r figure1_panel, fig.height=9, fig.width=12, dpi=300}
Figure1 <- p_fig1A / (p_fig1C | p_fig1D) + plot_layout(heights = c(0.9, 1.1)) + plot_annotation(tag_levels = "A")
Figure1
ggsave(here::here(output_dir_figs, "Figure1_panel.png"), Figure1, width = 12, height = 9, dpi = 300)
```


## Supplementary Figure 2: age and sex across RV-PA coupling groups

```{r supplementary2_data}
tapse_group_clin <- bind_rows(
  z_pah_tapse %>% transmute(Cohort = "Zurich", TAPSE_group, Age, Sex),
  p_pah_tapse %>% transmute(Cohort = "Krakow", TAPSE_group, Age, Sex)
) %>%
  mutate(Cohort = factor(Cohort, levels = c("Krakow", "Zurich")))

tapse_group_summary <- tapse_group_clin %>%
  group_by(Cohort, TAPSE_group) %>%
  summarise(
    n = n(),
    age_median = median(Age, na.rm = TRUE),
    age_q1 = quantile(Age, 0.25, na.rm = TRUE),
    age_q3 = quantile(Age, 0.75, na.rm = TRUE),
    female_n = sum(toupper(as.character(Sex)) == "F", na.rm = TRUE),
    sex_n = sum(!is.na(Sex)),
    .groups = "drop"
  ) %>%
  mutate(female_percent = 100 * female_n / sex_n)

tapse_group_summary
write.csv(tapse_group_summary,
          here::here(output_dir_data, "TAPSE_group_age_sex_summary.csv"),
          row.names = FALSE)
```

```{r supplementary2A_age, fig.height=4.8, fig.width=8, dpi=300}
p_supp2A <- ggplot(tapse_group_clin, aes(TAPSE_group, Age, fill = TAPSE_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.75, width = 0.55) +
  geom_jitter(width = 0.10, size = 2.0, alpha = 0.75) +
  facet_wrap(~Cohort, nrow = 1) +
  scale_fill_manual(values = cols_tapse) +
  scale_x_discrete(labels = coupling_axis_labels) +
  labs(x = NULL, y = "Age, years") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 14)
  )

p_supp2A
ggsave(here::here(output_dir_figs, "SupplementaryFigure2A_age.png"),
       p_supp2A, width = 8, height = 4.8, dpi = 300)
```

```{r supplementary2B_sex, fig.height=4.8, fig.width=8, dpi=300}
p_supp2B <- ggplot(tapse_group_summary, aes(TAPSE_group, female_percent, fill = TAPSE_group)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = paste0(female_n, "/", sex_n)), vjust = -0.35, size = 4.4) +
  facet_wrap(~Cohort, nrow = 1) +
  scale_fill_manual(values = cols_tapse) +
  scale_x_discrete(labels = coupling_axis_labels) +
  coord_cartesian(ylim = c(0, 105)) +
  labs(x = NULL, y = "Female, %") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 14)
  )

p_supp2B
ggsave(here::here(output_dir_figs, "SupplementaryFigure2B_sex.png"),
       p_supp2B, width = 8, height = 4.8, dpi = 300)
```

```{r supplementary_figure2_panel, fig.height=4.8, fig.width=12, dpi=300}
SupplementaryFigure2 <- p_supp2A | p_supp2B
SupplementaryFigure2 <- SupplementaryFigure2 + plot_annotation(tag_levels = "A")

SupplementaryFigure2
ggsave(here::here(output_dir_figs, "SupplementaryFigure2_age_sex.png"),
       SupplementaryFigure2, width = 12, height = 4.8, dpi = 300)
```

# Primary analysis: RV-PA coupling biomarker panel

## Differential abundance of RV-PA coupling - both cohorts

The same contrast is used in both cohorts: all SSc patients with available TAPSE/sPAP are classified as preserved or impaired RV-PA coupling using the 0.32 threshold. Models are adjusted for age and sex.

```{r de_coupling_zurich}
meta_z_coupling <- z_all %>%
  filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  mutate(
    TAPSE_group = ifelse(TAPSE_sPAP <= 0.32, "Impaired", "Preserved"),
    TAPSE_group = factor(TAPSE_group, levels = c("Preserved", "Impaired")),
    Age_z = as.numeric(scale(Age))
  ) %>%
  column_to_rownames("Barcode")

expr_z_coupling <- as.matrix(data_z[, rownames(meta_z_coupling), drop = FALSE])
design_z_coupling <- model.matrix(~ TAPSE_group + Age_z + Sex, data = meta_z_coupling)
fit_z_coupling <- eBayes(lmFit(expr_z_coupling, design_z_coupling))

DE_coupling_z <- topTable(
  fit_z_coupling, coef = "TAPSE_groupImpaired",
  number = Inf, adjust.method = "BH"
) %>% rownames_to_column("Protein")

write.csv(DE_coupling_z, here::here(output_dir_data, "DE_impaired_vs_preserved_RVPA_Zurich_all_SSc.csv"), row.names = FALSE)
print(table(meta_z_coupling$TAPSE_group))
```

```{r de_coupling_poland}
meta_p_coupling <- p_all %>%
  filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  mutate(
    TAPSE_group = ifelse(TAPSE_sPAP <= 0.32, "Impaired", "Preserved"),
    TAPSE_group = factor(TAPSE_group, levels = c("Preserved", "Impaired")),
    Age_z = as.numeric(scale(Age))
  ) %>%
  column_to_rownames("Barcode")

expr_p_coupling <- as.matrix(data_p[, rownames(meta_p_coupling), drop = FALSE])
design_p_coupling <- model.matrix(~ TAPSE_group + Age_z + Sex, data = meta_p_coupling)
fit_p_coupling <- eBayes(lmFit(expr_p_coupling, design_p_coupling))

DE_coupling_p <- topTable(
  fit_p_coupling, coef = "TAPSE_groupImpaired",
  number = Inf, adjust.method = "BH"
) %>% rownames_to_column("Protein")

write.csv(DE_coupling_p, here::here(output_dir_data, "DE_impaired_vs_preserved_RVPA_Krakow_all_SSc.csv"), row.names = FALSE)
print(table(meta_p_coupling$TAPSE_group))
```

## Supplementary Figure 3: RV-PA coupling differential abundance

```{r coupling_plot_helpers}
make_coupling_volcano <- function(res, title){
  keyvals <- rep("grey75", nrow(res))
  names(keyvals) <- rep("Not significant", nrow(res))
  keyvals[res$adj.P.Val < 0.1 & res$logFC > 0] <- "#B23A2B"
  keyvals[res$adj.P.Val < 0.1 & res$logFC < 0] <- "#2C6E9B"
  names(keyvals)[keyvals == "#B23A2B"] <- "Higher with impaired coupling"
  names(keyvals)[keyvals == "#2C6E9B"] <- "Lower with impaired coupling"

  EnhancedVolcano(
    res,
    lab = res$Protein,
    x = "logFC",
    y = "adj.P.Val",
    labSize = 0,
    pointSize = 1.5,
    pCutoff = 0.1,
    FCcutoff = 0,
    colCustom = keyvals,
    colAlpha = 1,
    subtitle = NULL,
    title = title,
    xlab = "NPX difference (impaired - preserved)"
  ) +
    theme(
      legend.position = "none",
      text = element_text(size = 14),
      plot.title = element_text(size = 16),
      axis.title = element_text(size = 14)
    )
}

concordance_coupling <- inner_join(
  DE_coupling_z %>% select(Protein, effect_z = logFC),
  DE_coupling_p %>% select(Protein, effect_p = logFC),
  by = "Protein"
)

pear_coupling <- cor.test(
  concordance_coupling$effect_z,
  concordance_coupling$effect_p,
  method = "pearson"
)

spear_coupling <- cor.test(
  concordance_coupling$effect_z,
  concordance_coupling$effect_p,
  method = "spearman",
  exact = FALSE
)

write.csv(concordance_coupling,
          here::here(output_dir_data, "RVPA_cross_cohort_effect_size_concordance.csv"),
          row.names = FALSE)
```

```{r supplementary3A_krakow_volcano, fig.height=5, fig.width=5.8, dpi=300}
n_imp_k <- sum(meta_p_coupling$TAPSE_group == "Impaired")
n_pres_k <- sum(meta_p_coupling$TAPSE_group == "Preserved")
title_k <- paste0("Krakow\nImpaired coupling (n=", n_imp_k, ") vs \npreserved coupling (n=", n_pres_k, ")")
p_supp3A <- make_coupling_volcano(DE_coupling_p, title_k)
p_supp3A
ggsave(here::here(output_dir_figs, "SupplementaryFigure3A_Krakow_RVPA_volcano.png"), p_supp3A, width = 5.8, height = 5, dpi = 300)
```

```{r supplementary3B_zurich_volcano, fig.height=5, fig.width=5.8, dpi=300}
n_imp_z <- sum(meta_z_coupling$TAPSE_group == "Impaired")
n_pres_z <- sum(meta_z_coupling$TAPSE_group == "Preserved")
title_z <- paste0("Zurich\nImpaired coupling (n=", n_imp_z, ") vs \npreserved coupling (n=", n_pres_z, ")")
p_supp3B <- make_coupling_volcano(DE_coupling_z, title_z)
p_supp3B
ggsave(here::here(output_dir_figs, "SupplementaryFigure3B_Zurich_RVPA_volcano.png"), p_supp3B, width = 5.8, height = 5, dpi = 300)
```

```{r supplementary3C_concordance, fig.height=5, fig.width=5.8, dpi=300}
p_supp3C <- ggplot(concordance_coupling, aes(effect_z, effect_p)) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "grey75") +
  geom_vline(xintercept = 0, linewidth = 0.4, color = "grey75") +
  geom_point(size = 1.6, alpha = 0.6) +
  geom_smooth(method = MASS::rlm, formula = y ~ x, se = FALSE, linewidth = 0.9, color = "grey30") +
  labs(x = "Zurich effect size", y = "Krakow effect size", title = paste0("Effect-size concordance\nPearson r = ", round(pear_coupling$estimate, 2), "; \nSpearman rho = ", round(spear_coupling$estimate, 2))) +
  theme_classic(base_size = 14)
p_supp3C
ggsave(here::here(output_dir_figs, "SupplementaryFigure3C_RVPA_concordance.png"), p_supp3C, width = 5.8, height = 5, dpi = 300)
```

```{r supplementary_figure3_panel, fig.height=5, fig.width=13, dpi=300}
SupplementaryFigure3 <- p_supp3A | p_supp3B | p_supp3C
SupplementaryFigure3 <- SupplementaryFigure3 + plot_annotation(tag_levels = "A")

SupplementaryFigure3
ggsave(here::here(output_dir_figs, "SupplementaryFigure3_RVPA_differential_abundance.png"),
       SupplementaryFigure3, width = 13, height = 5, dpi = 300)
```



# Protein panel development in Krakow

The development cohort includes all Krakow SSc patients with TAPSE/sPAP available. TAPSE/sPAP <= 0.32 defines impaired coupling and TAPSE/sPAP > 0.32 defines preserved coupling. External validation remains restricted to Zurich SSc-PAH patients.

## Prepare development and external validation cohorts

```{r panel_prepare}
p_dev <- p_all %>%
  filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex), Barcode %in% colnames(data_p)) %>%
  arrange(Barcode) %>%
  mutate(y_impaired = ifelse(TAPSE_sPAP <= 0.32, 1, 0), Impaired = y_impaired,
         TAPSE_group = factor(ifelse(y_impaired == 1, "Impaired", "Preserved"), levels = c("Preserved", "Impaired")))

z_val <- z_pah %>%
  filter(!is.na(TAPSE_sPAP), Barcode %in% colnames(data_z)) %>%
  arrange(Barcode) %>%
  mutate(y_impaired = ifelse(TAPSE_sPAP <= 0.32, 1, 0), Impaired = y_impaired,
         TAPSE_group = factor(ifelse(y_impaired == 1, "Impaired", "Preserved"), levels = c("Preserved", "Impaired")))

cat("Krakow development cohort n =", nrow(p_dev), "\n")
print(table(p_dev$Group, p_dev$TAPSE_group))
cat("Zurich external validation SSc-PAH n =", nrow(z_val), "\n")
print(table(z_val$TAPSE_group))
```

## Candidate ranking in Krakow

The top 50 proteins ranked by the absolute moderated limma t-statistic in the full Krakow development cohort are used as the final candidate pool. For estimation of internal predictive performance and panel size, the limma ranking is repeated inside each outer cross-validation training fold. For bootstrap stability selection, the top 50 candidates from the full Krakow development cohort are kept fixed.

```{r candidate_ranking_poland}
candidate_n <- 50
candidate_ranking_p <- DE_coupling_p %>% filter(Protein %in% common_proteins) %>% arrange(desc(abs(t)))
candidate_proteins <- candidate_ranking_p$Protein[seq_len(min(candidate_n, nrow(candidate_ranking_p)))]

candidate_ranking_p %>% select(Protein, logFC, t, P.Value, adj.P.Val) %>% head(50)
write.csv(candidate_ranking_p, here::here(output_dir_data, "Krakow_RVPA_candidate_ranking.csv"), row.names = FALSE)
write.csv(candidate_ranking_p %>% slice_head(n = candidate_n), here::here(output_dir_data, "Krakow_RVPA_top50_candidates.csv"), row.names = FALSE)
```

## Helper functions

```{r helper_panel_folds}
make_stratified_folds <- function(y, k = 5){
  fold_id <- integer(length(y))
  for (cl in sort(unique(y))){
    ids <- sample(which(y == cl))
    fold_id[ids] <- rep(seq_len(k), length.out = length(ids))
  }
  fold_id
}
```

```{r helper_panel_limma}
rank_proteins_limma <- function(expr_mat, meta, proteins){
  expr <- as.matrix(expr_mat[proteins, meta$Barcode, drop = FALSE])
  meta$Sex_M <- ifelse(as.character(meta$Sex) == "M", 1, 0)
  design <- model.matrix(~ Impaired + Age + Sex_M, data = meta)
  fit <- limma::eBayes(limma::lmFit(expr, design))
  limma::topTable(fit, coef = "Impaired", number = Inf, sort.by = "none") %>%
    rownames_to_column("Protein") %>% arrange(desc(abs(t)))
}
```

```{r helper_panel_auc}
auc_safe <- function(y, pred){
  ok <- is.finite(y) & is.finite(pred)
  if (sum(ok) < 3 || length(unique(y[ok])) < 2) return(NA_real_)
  as.numeric(pROC::auc(pROC::roc(y[ok], pred[ok], levels = c(0, 1), direction = "<", quiet = TRUE)))
}
```

## LASSO panel size: 1-10 proteins

For every outer training fold, candidate ranking is repeated from scratch, the top 50 proteins are passed to LASSO, and an inner cross-validation identifies the best solution for each panel size from 1 to 10 proteins. Performance across panel sizes is shown for transparency. For the final analysis, the panel size is fixed at five proteins.

```{r helper_lasso_lambda}
choose_lasso_lambda_for_k <- function(cvfit, k, kmax = 10){
  beta <- as.matrix(coef(cvfit$glmnet.fit))
  nzero <- colSums(beta[-1, , drop = FALSE] != 0)
  idx <- which(nzero == k & is.finite(cvfit$cvm))

  if (length(idx) == 0){
    allowed <- which(nzero >= 1 & nzero <= kmax & is.finite(cvfit$cvm))
    if (length(allowed) == 0) return(NULL)
    distance <- abs(nzero[allowed] - k)
    allowed <- allowed[distance == min(distance)]
    idx <- allowed
  }

  best <- idx[which.max(cvfit$cvm[idx])]
  list(lambda = cvfit$lambda[best], nzero = nzero[best], cv_auc = cvfit$cvm[best])
}
```

```{r helper_lasso_nested_cv}
run_lasso_nested_cv <- function(expr_mat, meta, proteins, candidate_n = 50, kmax = 10, repeats = 100, outer_folds = 5, inner_folds = 5, seed = 1){
  set.seed(seed)
  y <- meta$Impaired
  n <- nrow(meta)
  result_list <- vector("list", repeats * kmax)
  pred_sum <- matrix(0, nrow = n, ncol = kmax, dimnames = list(meta$Barcode, paste0("k", 1:kmax)))
  pred_n <- matrix(0, nrow = n, ncol = kmax, dimnames = list(meta$Barcode, paste0("k", 1:kmax)))
  counter <- 1

  for (r in seq_len(repeats)){
    outer_id <- make_stratified_folds(y, outer_folds)
    pred_repeat <- matrix(NA_real_, nrow = n, ncol = kmax)

    for (f in seq_len(outer_folds)){
      tr <- which(outer_id != f)
      te <- which(outer_id == f)
      meta_tr <- meta[tr, , drop = FALSE]
      meta_te <- meta[te, , drop = FALSE]

      ranking <- rank_proteins_limma(expr_mat, meta_tr, proteins)
      candidates <- ranking$Protein[seq_len(min(candidate_n, nrow(ranking)))]
      Xtr <- t(as.matrix(expr_mat[candidates, meta_tr$Barcode, drop = FALSE]))
      Xte <- t(as.matrix(expr_mat[candidates, meta_te$Barcode, drop = FALSE]))
      ytr <- meta_tr$Impaired

      inner_id <- make_stratified_folds(ytr, inner_folds)
      cvfit <- cv.glmnet(Xtr, ytr, family = "binomial", alpha = 1, type.measure = "auc", foldid = inner_id,
                         nlambda = 500, lambda.min.ratio = 1e-4, standardize = TRUE, dfmax = kmax)

      for (k in seq_len(kmax)){
        choice <- choose_lasso_lambda_for_k(cvfit, k, kmax)
        if (is.null(choice)) next
        pred_repeat[te, k] <- as.numeric(predict(cvfit$glmnet.fit, newx = Xte, s = choice$lambda, type = "response"))
      }
    }

    for (k in seq_len(kmax)){
      pred <- pred_repeat[, k]
      ok <- is.finite(pred)
      result_list[[counter]] <- data.frame(rep_id = r, k = k, AUC = auc_safe(y, pred))
      counter <- counter + 1
      pred_sum[ok, k] <- pred_sum[ok, k] + pred[ok]
      pred_n[ok, k] <- pred_n[ok, k] + 1
    }
  }

  results <- bind_rows(result_list)
  summary <- results %>% group_by(k) %>% summarise(
    mean_AUC = mean(AUC, na.rm = TRUE), sd_AUC = sd(AUC, na.rm = TRUE),
    se_AUC = sd_AUC / sqrt(sum(is.finite(AUC))),
    AUC_low = quantile(AUC, 0.025, na.rm = TRUE), AUC_high = quantile(AUC, 0.975, na.rm = TRUE), .groups = "drop")
  pred_avg <- pred_sum / pred_n
  list(results = results, summary = summary, pred_avg = pred_avg)
}
```

```{r lasso_nested_cv}
lasso_cv <- run_lasso_nested_cv(as.matrix(data_p), p_dev, common_proteins, candidate_n = 50, kmax = 10,
                                repeats = 100, outer_folds = 5, inner_folds = 5, seed = 100)
lasso_cv$summary
write.csv(lasso_cv$results, here::here(output_dir_data, "LASSO_nested_CV_all_repeats.csv"), row.names = FALSE)
write.csv(lasso_cv$summary, here::here(output_dir_data, "LASSO_nested_CV_summary.csv"), row.names = FALSE)
```

```{r select_lasso_size}
lasso_cv$summary <- lasso_cv$summary %>%
  arrange(k) %>%
  mutate(Delta_AUC_vs_previous = c(NA, diff(mean_AUC)))

lasso_delta_by_repeat <- lasso_cv$results %>%
  arrange(rep_id, k) %>%
  group_by(rep_id) %>%
  mutate(Delta_AUC_vs_previous = AUC - lag(AUC)) %>%
  ungroup() %>%
  filter(k > 1)

lasso_delta_summary <- lasso_delta_by_repeat %>%
  group_by(k) %>%
  summarise(
    mean_delta = mean(Delta_AUC_vs_previous, na.rm = TRUE),
    delta_low = quantile(Delta_AUC_vs_previous, 0.025, na.rm = TRUE),
    delta_high = quantile(Delta_AUC_vs_previous, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

lasso_k <- 5

lasso_cv$summary
lasso_delta_summary

cat("Panel size fixed at:", lasso_k, "proteins\n")

write.csv(lasso_cv$summary, here::here(output_dir_data, "LASSO_nested_CV_summary_with_incremental_AUC.csv"), row.names = FALSE)
write.csv(lasso_delta_summary, here::here(output_dir_data, "LASSO_incremental_AUC_by_panel_size.csv"), row.names = FALSE)
```

```{r plot_lasso_cv, fig.width=7, fig.height=5, dpi=300}
auc_ymin <- max(0.5, min(lasso_cv$summary$AUC_low, na.rm = TRUE) - 0.02)
auc_ymax <- min(1, max(lasso_cv$summary$AUC_high, na.rm = TRUE) + 0.02)

p_panel_size <- ggplot(lasso_cv$summary, aes(k, mean_AUC)) +
  geom_errorbar(aes(ymin = AUC_low, ymax = AUC_high), width = 0.12, linewidth = 0.55) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3.0) +
  geom_vline(xintercept = lasso_k, linetype = 2, linewidth = 0.6) +
  scale_x_continuous(breaks = 1:10) +
  coord_cartesian(ylim = c(auc_ymin, auc_ymax)) +
  labs(
    x = "Number of proteins",
    y = "Repeated nested-CV AUC",
    title = "Panel size"
  ) +
  theme_classic(base_size = 14)

p_panel_size
ggsave(here::here(output_dir_figs, "Figure2A_panel_size.png"),
       p_panel_size, width = 7, height = 5, dpi = 300)
```

## Select the final LASSO panel in Krakow

After the panel size is fixed, the top 50 candidates from the full Krakow development cohort are kept fixed. LASSO is fitted in the full Krakow development cohort, and the solution with the selected number of proteins and the highest internal cross-validated AUC is used as the final panel. Bootstrap selection frequencies are calculated afterwards as a stability analysis and do not determine the final proteins.

```{r helper_final_panel_selection}
select_best_panel_for_k <- function(cvfit, k){
  beta <- as.matrix(coef(cvfit$glmnet.fit))
  protein_names <- rownames(beta)[-1]
  nzero <- colSums(beta[-1, , drop = FALSE] != 0)
  valid <- which(is.finite(cvfit$cvm))
  exact <- valid[nzero[valid] == k]

  if (length(exact) > 0){
    best <- exact[which.max(cvfit$cvm[exact])]
    co <- beta[-1, best]
    names(co) <- protein_names
    selected <- names(co)[co != 0]
  } else {
    above <- valid[nzero[valid] > k]
    if (length(above) == 0) stop("No LASSO solution with at least the selected number of proteins was found.")
    closest_n <- min(nzero[above])
    closest <- above[nzero[above] == closest_n]
    best <- closest[which.max(cvfit$cvm[closest])]
    co <- beta[-1, best]
    names(co) <- protein_names
    co <- co[co != 0]
    selected <- names(sort(abs(co), decreasing = TRUE))[seq_len(k)]
  }

  list(proteins = selected, lambda = cvfit$lambda[best], cv_auc = cvfit$cvm[best], path_n = nzero[best])
}
```

```{r final_lasso_panel}
X_lasso_full <- t(as.matrix(data_p[candidate_proteins, p_dev$Barcode, drop = FALSE]))
y_lasso_full <- p_dev$Impaired

set.seed(102)
foldid_lasso_full <- make_stratified_folds(y_lasso_full, 5)
cvfit_lasso_full <- cv.glmnet(X_lasso_full, y_lasso_full, family = "binomial", alpha = 1, type.measure = "auc",
                              foldid = foldid_lasso_full, nlambda = 500, lambda.min.ratio = 1e-4,
                              standardize = TRUE, dfmax = 10)

final_panel_choice <- select_best_panel_for_k(cvfit_lasso_full, lasso_k)
final_panel <- final_panel_choice$proteins

cat("Selected panel size:", lasso_k, "proteins\n")
cat("Internal CV AUC of the selected full-Krakow LASSO solution:", round(final_panel_choice$cv_auc, 3), "\n")
cat("Final proteins:\n")
print(final_panel)

write.csv(data.frame(Protein = final_panel), here::here(output_dir_data, "Final_panel_proteins.csv"), row.names = FALSE)
```

## Bootstrap LASSO stability

The final panel is already fixed above. The same top 50 candidate proteins are kept fixed in the bootstrap analysis. Each bootstrap LASSO selects five proteins. Selection frequency is used only to assess stability and does not determine the final panel.

```{r helper_lasso_bootstrap}
bootstrap_lasso_selection <- function(expr_mat, meta, candidates, k, B = 500, inner_folds = 5, seed = 1){
  set.seed(seed)
  selected_lists <- vector("list", B)

  for (b in seq_len(B)){
    ids <- sample(seq_len(nrow(meta)), nrow(meta), replace = TRUE)
    meta_b <- meta[ids, , drop = FALSE]
    y <- meta_b$Impaired
    if (length(unique(y)) < 2) next

    X <- t(as.matrix(expr_mat[candidates, meta_b$Barcode, drop = FALSE]))
    folds_here <- min(inner_folds, min(table(y)))
    if (folds_here < 3) next
    foldid <- make_stratified_folds(y, folds_here)

    cvfit <- try(
      cv.glmnet(
        X, y, family = "binomial", alpha = 1, type.measure = "auc",
        foldid = foldid, nlambda = 500, lambda.min.ratio = 1e-4,
        standardize = TRUE, dfmax = 10
      ),
      silent = TRUE
    )
    if (inherits(cvfit, "try-error")) next

    choice <- try(select_best_panel_for_k(cvfit, k), silent = TRUE)
    if (inherits(choice, "try-error") || is.null(choice)) next
    selected_lists[[b]] <- choice$proteins
  }

  selected_lists <- selected_lists[lengths(selected_lists) == k]
  if (length(selected_lists) == 0) stop("No valid bootstrap LASSO models were obtained.")

  selection_matrix <- matrix(
    0,
    nrow = length(selected_lists),
    ncol = length(candidates),
    dimnames = list(NULL, candidates)
  )

  for (i in seq_along(selected_lists)){
    selection_matrix[i, selected_lists[[i]]] <- 1
  }

  selection_frequency <- colMeans(selection_matrix)

  summary <- data.frame(
    Protein = candidates,
    Selection_frequency = selection_frequency,
    row.names = NULL
  ) %>%
    arrange(desc(Selection_frequency))

  list(
    summary = summary,
    selected_lists = selected_lists,
    selection_matrix = selection_matrix,
    n_valid = nrow(selection_matrix)
  )
}
```
```{r lasso_bootstrap}
stability50 <- bootstrap_lasso_selection(
  as.matrix(data_p), p_dev, candidate_proteins,
  k = lasso_k, B = 500, inner_folds = 5, seed = 101
)

stability50$summary <- stability50$summary %>%
  left_join(
    candidate_ranking_p %>% select(Protein, logFC, t, P.Value, adj.P.Val),
    by = "Protein"
  ) %>%
  arrange(desc(Selection_frequency), desc(abs(t)))

cat("Valid bootstrap models:", stability50$n_valid, "of 500\n")
cat("Final panel selected from the best-performing full-Krakow five-protein LASSO solution:\n")
print(final_panel)

stability50$summary %>% head(20)

write.csv(
  stability50$summary,
  here::here(output_dir_data, "LASSO_bootstrap_selection_frequency.csv"),
  row.names = FALSE
)
```

```{r plot_lasso_stability, fig.width=7, fig.height=5.2, dpi=300}
stability_plot_proteins <- unique(c(
  stability50$summary$Protein[seq_len(min(15, nrow(stability50$summary)))],
  final_panel
))

p_stability <- stability50$summary %>%
  filter(Protein %in% stability_plot_proteins) %>%
  arrange(Selection_frequency) %>%
  mutate(
    Protein = factor(Protein, levels = Protein),
    Final_panel = ifelse(Protein %in% final_panel, "Final panel", "Other")
  ) %>%
  ggplot(aes(Protein, Selection_frequency, fill = Final_panel)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Final panel" = "#B23A2B", "Other" = "#BDBDBD")) +
  labs(
    x = NULL,
    y = "Bootstrap selection frequency",
    fill = NULL,
    title = "Bootstrap selection frequency"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "top")

p_stability
ggsave(here::here(output_dir_figs, "Figure2B_bootstrap_selection_frequency.png"),
       p_stability, width = 7, height = 5.2, dpi = 300)
```

## Fit the frozen Krakow model

A standard logistic model is fitted using the final selected proteins. Protein means and standard deviations, model coefficients, and score standardization are all estimated in Krakow and then applied unchanged to Zurich.


```{r final_panel_model}
fit_final_panel <- function(expr_mat, clin_df, proteins){
  X <- t(as.matrix(expr_mat[proteins, clin_df$Barcode, drop = FALSE]))
  mu <- colMeans(X)
  sdv <- apply(X, 2, sd)
  sdv[sdv == 0 | is.na(sdv)] <- 1

  Z <- sweep(sweep(X, 2, mu, "-"), 2, sdv, "/")
  y <- clin_df$y_impaired

  fit <- suppressWarnings(
    glm.fit(cbind(Intercept = 1, Z), y, family = binomial())
  )

  beta <- fit$coefficients
  names(beta) <- c("Intercept", colnames(Z))

  if (!fit$converged){
    warning("Final logistic model did not fully converge. Inspect the final panel before interpretation.")
  }

  if (any(!is.finite(beta))){
    stop("Non-finite coefficient in final logistic model. Inspect separation in the selected panel.")
  }

  raw_score <- as.numeric(cbind(Intercept = 1, Z) %*% beta)
  score_center <- mean(raw_score)
  score_scale <- sd(raw_score)
  if (score_scale == 0 || is.na(score_scale)) score_scale <- 1

  list(
    proteins = proteins,
    mu = mu,
    sd = sdv,
    beta = beta,
    score_center = score_center,
    score_scale = score_scale
  )
}

predict_final_panel <- function(model, expr_mat, barcodes){
  X <- t(as.matrix(expr_mat[model$proteins, barcodes, drop = FALSE]))
  Z <- sweep(sweep(X, 2, model$mu, "-"), 2, model$sd, "/")

  raw_score <- as.numeric(
    cbind(Intercept = 1, Z) %*% model$beta
  )

  data.frame(
    Barcode = barcodes,
    probability_impaired = plogis(raw_score),
    panel_score = (raw_score - model$score_center) / model$score_scale
  )
}

panel_model <- fit_final_panel(
  as.matrix(data_p),
  p_dev,
  final_panel
)

pred_p_dev <- predict_final_panel(
  panel_model,
  as.matrix(data_p),
  p_dev$Barcode
) %>%
  left_join(
    p_dev %>% select(Barcode, Group, TAPSE_sPAP, TAPSE_group, y_impaired),
    by = "Barcode"
  )

pred_z_final <- predict_final_panel(
  panel_model,
  as.matrix(data_z),
  z_val$Barcode
) %>%
  left_join(
    z_val %>% select(Barcode, TAPSE_sPAP, TAPSE_group, y_impaired),
    by = "Barcode"
  )

# PAH-only Krakow predictions for downstream clinical comparisons
p_pah_score <- p_pah %>%
  filter(
    !is.na(TAPSE_sPAP),
    Barcode %in% colnames(data_p)
  ) %>%
  arrange(Barcode) %>%
  mutate(
    y_impaired = ifelse(TAPSE_sPAP <= 0.32, 1, 0),
    TAPSE_group = factor(
      ifelse(y_impaired == 1, "Impaired", "Preserved"),
      levels = c("Preserved", "Impaired")
    )
  )

pred_p_final <- predict_final_panel(
  panel_model,
  as.matrix(data_p),
  p_pah_score$Barcode
) %>%
  left_join(
    p_pah_score %>% select(Barcode, TAPSE_sPAP, TAPSE_group, y_impaired),
    by = "Barcode"
  )

panel_coefficients <- data.frame(
  Protein = final_panel,
  Coefficient = panel_model$beta[final_panel],
  OR_per_1SD_NPX = exp(panel_model$beta[final_panel])
) %>%
  left_join(
    stability50$summary %>%
      select(Protein, Selection_frequency, logFC, t, P.Value, adj.P.Val),
    by = "Protein"
  )

panel_coefficients

write.csv(
  panel_coefficients,
  here::here(output_dir_data, "Final_panel_coefficients.csv"),
  row.names = FALSE
)

qs_save(
  panel_model,
  here::here(output_dir_data, "Final_panel_model_frozen_from_Krakow.qs2")
)
```

```{r figure2B_alternative_coefficients, fig.width=6.5, fig.height=5, dpi=300}
p_panel_coefficients <- panel_coefficients %>%
  mutate(Protein = factor(Protein, levels = rev(final_panel))) %>%
  ggplot(aes(Coefficient, Protein)) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey65") +
  geom_segment(
    aes(x = 0, xend = Coefficient, y = Protein, yend = Protein),
    linewidth = 0.9
  ) +
  geom_point(size = 3.2) +
  labs(
    x = "Final logistic-model coefficient",
    y = NULL,
    title = "Final five-protein model"
  ) +
  theme_classic(base_size = 14)

p_panel_coefficients
ggsave(
  here::here(output_dir_figs, "Figure2B_alternative_final_model_coefficients.png"),
  p_panel_coefficients,
  width = 6.5,
  height = 5,
  dpi = 300
)
```

## Development and external validation performance

```{r panel_validation_metrics}
poland_oof_prob <- lasso_cv$pred_avg[, paste0("k", lasso_k)]
roc_panel_p <- pROC::roc(p_dev$y_impaired, poland_oof_prob, levels = c(0, 1), direction = "<", quiet = TRUE)
roc_panel_z <- pROC::roc(pred_z_final$y_impaired, pred_z_final$probability_impaired, levels = c(0, 1), direction = "<", quiet = TRUE, ci = TRUE)
roc_panel_p_apparent <- pROC::roc(pred_p_dev$y_impaired, pred_p_dev$probability_impaired, levels = c(0, 1), direction = "<", quiet = TRUE)

cv_selected_summary <- lasso_cv$summary %>% filter(k == lasso_k)
panel_auc <- data.frame(
  Cohort = c("Krakow nested CV", "Krakow apparent", "Zurich external"),
  AUC = c(cv_selected_summary$mean_AUC, as.numeric(pROC::auc(roc_panel_p_apparent)), as.numeric(pROC::auc(roc_panel_z))),
  AUC_low = c(cv_selected_summary$AUC_low, NA, as.numeric(pROC::ci.auc(roc_panel_z)[1])),
  AUC_high = c(cv_selected_summary$AUC_high, NA, as.numeric(pROC::ci.auc(roc_panel_z)[3]))
)
panel_auc
write.csv(panel_auc, here::here(output_dir_data, "Final_panel_AUC_validation.csv"), row.names = FALSE)


roc_band_panel_k <- roc_bootstrap_band(
  p_dev$y_impaired,
  poland_oof_prob,
  B = 2000,
  seed = 201
) %>% mutate(Cohort = "Krakow nested CV")

roc_band_panel_z <- roc_bootstrap_band(
  pred_z_final$y_impaired,
  pred_z_final$probability_impaired,
  B = 2000,
  seed = 202
) %>% mutate(Cohort = "Zurich external")

roc_band_panel <- bind_rows(roc_band_panel_k, roc_band_panel_z)
```


## Score across SSc-noPAH, SSc-PAH with preserved coupling and SSc-PAH with impaired coupling

```{r score_three_groups, fig.height=5.2, fig.width=9.5, dpi=300}
pred_p_all <- predict_final_panel(panel_model, as.matrix(data_p), p_all$Barcode) %>%
  left_join(p_all %>% select(Barcode, Group, TAPSE_sPAP), by = "Barcode")

pred_z_all <- predict_final_panel(panel_model, as.matrix(data_z), z_all$Barcode) %>%
  left_join(z_all %>% select(Barcode, Group, TAPSE_sPAP), by = "Barcode")

score_group_df <- bind_rows(
  pred_p_all %>% mutate(Cohort = "Krakow"),
  pred_z_all %>% mutate(Cohort = "Zurich")
) %>%
  mutate(
    Cohort = factor(Cohort, levels = c("Krakow", "Zurich")),
    Group3 = case_when(
      Group == "SSc-noPAH" ~ "SSc-noPAH",
      Group == "SSc-PAH" & TAPSE_sPAP > 0.32 ~ "SSc-PAH\nPreserved\ncoupling",
      Group == "SSc-PAH" & TAPSE_sPAP <= 0.32 ~ "SSc-PAH\nImpaired\ncoupling",
      TRUE ~ NA_character_
    ),
    Group3 = factor(
      Group3,
      levels = c(
        "SSc-noPAH",
        "SSc-PAH\nPreserved\ncoupling",
        "SSc-PAH\nImpaired\ncoupling"
      )
    )
  ) %>%
  filter(!is.na(Group3))

p_score_three <- ggplot(score_group_df, aes(Group3, panel_score, fill = Group3)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.75, width = 0.55) +
  geom_jitter(width = 0.10, size = 2.0, alpha = 0.75) +
  facet_wrap(~Cohort, nrow = 1) +
  scale_fill_manual(values = cols_three) +
  labs(x = NULL, y = "RV-PA coupling protein score") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 14),
    axis.text.x = element_text(size = 11.5)
  )

p_score_three
ggsave(here::here(output_dir_figs, "Figure2C_score_across_groups.png"),
       p_score_three, width = 9.5, height = 5.2, dpi = 300)
```



```{r figure2D_validation_ROC, fig.height=5.2, fig.width=6.5, dpi=300}
roc_panel_df <- bind_rows(
  roc_curve_df(p_dev$y_impaired, poland_oof_prob) %>% mutate(Cohort = "Krakow nested CV"),
  roc_curve_df(pred_z_final$y_impaired, pred_z_final$probability_impaired) %>% mutate(Cohort = "Zurich external")
)

panel_auc_labels <- c(
  "Krakow nested CV" = paste0(
    "Krakow nested CV (AUC ",
    sprintf("%.2f", panel_auc$AUC[panel_auc$Cohort == "Krakow nested CV"]),
    ", 95% CI ",
    sprintf("%.2f", panel_auc$AUC_low[panel_auc$Cohort == "Krakow nested CV"]),
    "-",
    sprintf("%.2f", panel_auc$AUC_high[panel_auc$Cohort == "Krakow nested CV"]),
    ")"
  ),
  "Zurich external" = paste0(
    "Zurich external (AUC ",
    sprintf("%.2f", panel_auc$AUC[panel_auc$Cohort == "Zurich external"]),
    ", 95% CI ",
    sprintf("%.2f", panel_auc$AUC_low[panel_auc$Cohort == "Zurich external"]),
    "-",
    sprintf("%.2f", panel_auc$AUC_high[panel_auc$Cohort == "Zurich external"]),
    ")"
  )
)

roc_panel_df <- roc_panel_df %>%
  mutate(Label = panel_auc_labels[Cohort])

roc_band_panel_plot <- roc_band_panel %>%
  mutate(Label = panel_auc_labels[Cohort])

panel_curve_cols <- setNames(
  c("#B23A2B", "#2C6E9B"),
  unname(panel_auc_labels[c("Krakow nested CV", "Zurich external")])
)

panel_fill_cols <- setNames(
  c("#E5A49A", "#A8C5D5"),
  unname(panel_auc_labels[c("Krakow nested CV", "Zurich external")])
)

p_panel_roc <- ggplot() +
  geom_ribbon(
    data = roc_band_panel_plot,
    aes(x = FPR, ymin = lower, ymax = upper, fill = Label),
    alpha = 0.25
  ) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey70") +
  geom_step(
    data = roc_panel_df,
    aes(FPR, Sensitivity, color = Label),
    direction = "hv",
    linewidth = 1.0
  ) +
  scale_color_manual(values = panel_curve_cols) +
  scale_fill_manual(values = panel_fill_cols) +
  coord_equal() +
  labs(
    x = "1 - Specificity",
    y = "Sensitivity",
    color = NULL,
    fill = NULL,
    title = "Impaired RV-PA coupling"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom")

p_panel_roc
ggsave(here::here(output_dir_figs, "Figure2D_validation_ROC.png"),
       p_panel_roc, width = 6.5, height = 5.2, dpi = 300)
```
```{r figure2_panel_development, fig.height=9, fig.width=12, dpi=300}
Figure2 <- (p_panel_size | p_stability) / (p_score_three | p_panel_roc) +
  plot_annotation(tag_levels = "A")

Figure2
ggsave(here::here(output_dir_figs, "Figure2_panel.png"),
       Figure2, width = 12, height = 9, dpi = 300)
```

## Individual selected proteins vs continuous TAPSE/sPAP


```{r figure2_alternative_coefficients, fig.height=9, fig.width=12, dpi=300}
Figure2_alternative <- (p_panel_size | p_panel_coefficients) / (p_score_three | p_panel_roc) +
  plot_annotation(tag_levels = "A")

Figure2_alternative
ggsave(here::here(output_dir_figs, "Figure2_alternative_with_coefficients.png"),
       Figure2_alternative, width = 12, height = 9, dpi = 300)
```



## Individual selected proteins vs continuous TAPSE/sPAP

```{r figure3_data}
make_panel_long <- function(expr_mat, clin_df, cohort){
  x <- t(as.matrix(expr_mat[final_panel, clin_df$Barcode, drop = FALSE])) %>% as.data.frame() %>% rownames_to_column("Barcode")
  clin_df %>% select(Barcode, TAPSE_sPAP) %>% left_join(x, by = "Barcode") %>% pivot_longer(all_of(final_panel), names_to = "Protein", values_to = "NPX") %>% mutate(Cohort = cohort)
}
panel_long <- bind_rows(make_panel_long(data_p, p_pah_score, "Krakow"), make_panel_long(data_z, z_val, "Zurich")) %>%
  mutate(Cohort = factor(Cohort, levels = c("Krakow", "Zurich")), Protein = factor(Protein, levels = final_panel))
panel_protein_cor <- panel_long %>% group_by(Cohort, Protein) %>% summarise(
  n = sum(is.finite(TAPSE_sPAP) & is.finite(NPX)),
  rho = cor(TAPSE_sPAP, NPX, method = "spearman", use = "complete.obs"),
  p = suppressWarnings(cor.test(TAPSE_sPAP, NPX, method = "spearman", exact = FALSE)$p.value),
  .groups = "drop"
) %>% mutate(label = paste0("rho = ", round(rho, 2), "\np = ", signif(p, 2)))
panel_protein_cor
write.csv(panel_protein_cor, here::here(output_dir_data, "Figure3_individual_panel_protein_correlations.csv"), row.names = FALSE)
```







```{r figure3_panel, fig.height=6.5, fig.width=13.5, dpi=300}
p_fig3 <- ggplot(panel_long, aes(TAPSE_sPAP, NPX)) +
  geom_point(size = 1.9, alpha = 0.78) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.9, color = "#3568C8") +
  facet_grid(Cohort ~ Protein, scales = "free_y") +
  geom_text(data = panel_protein_cor, aes(x = -Inf, y = Inf, label = label), inherit.aes = FALSE, hjust = -0.03, vjust = 1.08, size = 3.8) +
  labs(x = "TAPSE/sPAP", y = "NPX") +
  theme_classic(base_size = 14) +
  theme(strip.background = element_rect(fill = "white"), strip.text.x = element_text(size = 13), strip.text.y = element_text(size = 13), panel.spacing = unit(0.25, "lines"))
Figure3 <- p_fig3
Figure3
ggsave(here::here(output_dir_figs, "Figure3_individual_panel_proteins.png"), Figure3, width = 13.5, height = 6.5, dpi = 300)
```

## Zurich within-cohort scaling sensitivity

```{r zurich_scaling_sensitivity}
Xz_panel <- t(as.matrix(data_z[final_panel, z_val$Barcode, drop = FALSE]))
mu_z_panel <- colMeans(Xz_panel)
sd_z_panel <- apply(Xz_panel, 2, sd)
sd_z_panel[sd_z_panel == 0 | is.na(sd_z_panel)] <- 1
Zz_panel <- sweep(sweep(Xz_panel, 2, mu_z_panel, "-"), 2, sd_z_panel, "/")
raw_z_within <- as.numeric(cbind(Intercept = 1, Zz_panel) %*% panel_model$beta)
prob_z_within <- plogis(raw_z_within)
roc_z_within <- pROC::roc(z_val$y_impaired, prob_z_within, levels = c(0, 1), direction = "<", quiet = TRUE)
within_scaling_results <- data.frame(Approach = c("Frozen Krakow scaling", "Zurich within-cohort scaling"),
                                     AUC = c(as.numeric(pROC::auc(roc_panel_z)), as.numeric(pROC::auc(roc_z_within))))
within_scaling_results
write.csv(within_scaling_results, here::here(output_dir_data, "Zurich_scaling_sensitivity.csv"), row.names = FALSE)
```



# Clinical associations of the final protein score

## Prepare score and clinical data

```{r clinical_score_data}
score_p <- pred_p_final %>% select(Barcode, panel_score) %>% left_join(p_pah_score, by = "Barcode") %>% mutate(Cohort = "Krakow")
score_z <- pred_z_final %>% select(Barcode, panel_score) %>% left_join(z_val, by = "Barcode") %>% mutate(Cohort = "Zurich")
score_both <- bind_rows(score_p, score_z)
```

## Score associations with clinical variables

```{r clinical_correlations}
clinical_vars <- c("TAPSE_sPAP", "TAPSE", "sPAP", "mPAP", "PVR", "CI", "NTproBNP_common", "SixMWD", "FVC_pred_common", "DLCO_pred_common", "Creatinine_mg_dl", "mRSS_common", "COMPERA_number_common", "Age")
clinical_labels <- c(
  TAPSE_sPAP = "TAPSE/sPAP", TAPSE = "TAPSE", sPAP = "sPAP", mPAP = "mPAP", PVR = "PVR", CI = "Cardiac index",
  NTproBNP_common = "NT-proBNP", SixMWD = "6MWD", FVC_pred_common = "FVC", DLCO_pred_common = "DLCO",
  Creatinine_mg_dl = "Creatinine", mRSS_common = "mRSS", COMPERA_number_common = "COMPERA", Age = "Age"
)

clinical_cor_list <- list()
counter <- 1
for (cohort in c("Krakow", "Zurich")){
  dd <- score_both %>% filter(Cohort == cohort)
  for (v in clinical_vars){
    ok <- is.finite(dd$panel_score) & is.finite(as.numeric(dd[[v]]))
    if (sum(ok) >= 5){
      ct <- suppressWarnings(cor.test(dd$panel_score[ok], as.numeric(dd[[v]])[ok], method = "spearman", exact = FALSE))
      clinical_cor_list[[counter]] <- data.frame(Cohort = cohort, Variable = v, n = sum(ok), rho = unname(ct$estimate), p = ct$p.value)
    } else {
      clinical_cor_list[[counter]] <- data.frame(Cohort = cohort, Variable = v, n = sum(ok), rho = NA, p = NA)
    }
    counter <- counter + 1
  }
}

clinical_cor <- bind_rows(clinical_cor_list) %>% group_by(Cohort) %>% mutate(FDR = p.adjust(p, method = "BH")) %>% ungroup() %>%
  mutate(Label = clinical_labels[Variable])
clinical_cor
write.csv(clinical_cor, here::here(output_dir_data, "Final_score_clinical_correlations.csv"), row.names = FALSE)
```

```{r clinical_heatmap, fig.height=6.5, fig.width=6.2, dpi=300}
p_clinical_heatmap <- clinical_cor %>%
  mutate(Label = factor(Label, levels = rev(unname(clinical_labels)))) %>%
  ggplot(aes(Cohort, Label, fill = rho)) +
  geom_tile(color = "white") +
  geom_text(
    aes(label = ifelse(is.na(rho), "", sprintf("%.2f", rho))),
    size = 4.0
  ) +
  scale_fill_gradient2(
    low = "#2C6E9B",
    mid = "white",
    high = "#B23A2B",
    midpoint = 0,
    limits = c(-1, 1),
    na.value = "grey90"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Spearman rho",
    title = "Clinical associations of the protein score"
  ) +
  theme_classic(base_size = 14)

p_clinical_heatmap
ggsave(here::here(output_dir_figs, "SupplementaryFigure4_clinical_associations.png"),
       p_clinical_heatmap, width = 6.2, height = 6.5, dpi = 300)
```

```{r supplementary_figure4_panel, fig.height=6.5, fig.width=6.2, dpi=300}
SupplementaryFigure4 <- p_clinical_heatmap

SupplementaryFigure4
ggsave(
  here::here(output_dir_figs, "SupplementaryFigure4_panel_clinical_associations.png"),
  SupplementaryFigure4,
  width = 6.2,
  height = 6.5,
  dpi = 300
)
```


## ILD and sex

```{r binary_clinical_associations}
score_binary_assoc <- list()
counter <- 1
for (cohort in c("Krakow", "Zurich")){
  dd <- score_both %>% filter(Cohort == cohort)

  dd_ild <- dd %>% filter(ILD_common %in% c("yes", "no"), !is.na(panel_score))
  if (nrow(dd_ild) >= 5 && length(unique(dd_ild$ILD_common)) == 2){
    wt <- wilcox.test(panel_score ~ ILD_common, data = dd_ild)
    score_binary_assoc[[counter]] <- data.frame(Cohort = cohort, Variable = "ILD", n = nrow(dd_ild), p = wt$p.value)
    counter <- counter + 1
  }

  dd_sex <- dd %>% filter(!is.na(Sex), !is.na(panel_score))
  if (nrow(dd_sex) >= 5 && length(unique(dd_sex$Sex)) == 2){
    wt <- wilcox.test(panel_score ~ Sex, data = dd_sex)
    score_binary_assoc[[counter]] <- data.frame(Cohort = cohort, Variable = "Sex", n = nrow(dd_sex), p = wt$p.value)
    counter <- counter + 1
  }
}
score_binary_assoc <- bind_rows(score_binary_assoc)
score_binary_assoc
write.csv(score_binary_assoc, here::here(output_dir_data, "Final_score_binary_clinical_associations.csv"), row.names = FALSE)
```


## Adjusted association of the protein score with continuous TAPSE/sPAP

The association is tested separately in Krakow and Zurich SSc-PAH patients. TAPSE/sPAP is kept continuous. Model-specific complete cases are used and the available sample size is reported. mRSS is not included in the Krakow model because of missingness.

```{r adjusted_coupling_linear}
make_ild_binary <- function(x){
  xx <- tolower(trimws(as.character(x)))
  case_when(
    xx %in% c("yes", "y", "1", "true", "present") ~ 1,
    xx %in% c("no", "n", "0", "false", "absent") ~ 0,
    TRUE ~ NA_real_
  )
}

adjusted_linear_df <- bind_rows(
  score_p %>% mutate(Cohort = "Krakow"),
  score_z %>% mutate(Cohort = "Zurich")
) %>%
  mutate(
    Cohort = factor(Cohort, levels = c("Krakow", "Zurich")),
    Sex = factor(Sex),
    ILD_binary = make_ild_binary(ILD_common)
  )

fit_score_linear <- function(data, formula, cohort_name, model_name){
  vars <- all.vars(formula)
  dd <- data %>%
    filter(Cohort == cohort_name) %>%
    filter(if_all(all_of(vars), ~ !is.na(.)))

  if (nrow(dd) <= length(vars)){
    return(data.frame(
      Cohort = cohort_name,
      Model = model_name,
      n = nrow(dd),
      Beta = NA_real_,
      CI_low = NA_real_,
      CI_high = NA_real_,
      p = NA_real_,
      Adjusted_R2 = NA_real_
    ))
  }

  fit <- lm(formula, data = dd)
  tt <- broom::tidy(fit, conf.int = TRUE) %>%
    filter(term == "panel_score")

  data.frame(
    Cohort = cohort_name,
    Model = model_name,
    n = nrow(dd),
    Beta = tt$estimate,
    CI_low = tt$conf.low,
    CI_high = tt$conf.high,
    p = tt$p.value,
    Adjusted_R2 = summary(fit)$adj.r.squared
  )
}

adjusted_coupling_results <- bind_rows(
  fit_score_linear(
    adjusted_linear_df,
    TAPSE_sPAP ~ panel_score,
    "Krakow",
    "Unadjusted"
  ),
  fit_score_linear(
    adjusted_linear_df,
    TAPSE_sPAP ~ panel_score + Age + Sex,
    "Krakow",
    "Age + sex"
  ),
  fit_score_linear(
    adjusted_linear_df,
    TAPSE_sPAP ~ panel_score + Age + Sex + ILD_binary + Creatinine_mg_dl,
    "Krakow",
    "Age + sex + ILD + creatinine"
  ),
  fit_score_linear(
    adjusted_linear_df,
    TAPSE_sPAP ~ panel_score,
    "Zurich",
    "Unadjusted"
  ),
  fit_score_linear(
    adjusted_linear_df,
    TAPSE_sPAP ~ panel_score + Age + Sex,
    "Zurich",
    "Age + sex"
  ),
  fit_score_linear(
    adjusted_linear_df,
    TAPSE_sPAP ~ panel_score + Age + Sex + ILD_binary + Creatinine_mg_dl + mRSS_common,
    "Zurich",
    "Age + sex + ILD + creatinine + mRSS"
  )
)

adjusted_coupling_results
write.csv(adjusted_coupling_results,
          here::here(output_dir_data, "TableS4_adjusted_score_continuous_TAPSE_sPAP.csv"),
          row.names = FALSE)
```

# COMPERA and mortality - 3-year follow-up

This section uses administrative censoring at 3 years. Deaths occurring after 3 years are treated as censored at 3 years. Patients who are alive with shorter available follow-up remain censored at their last available follow-up.

```{r compera_3y_settings}
followup_limit_days <- 3 * 365.25
cat("Follow-up horizon:", followup_limit_days, "days (3 years)\n")
```

## Protein score across COMPERA strata

This is a baseline cross-sectional analysis and therefore does not require time restriction.

```{r score_compera_3y, fig.height=5, fig.width=8.5, dpi=300}
score_compera_df <- score_both %>%
  filter(!is.na(COMPERA), !is.na(COMPERA_number_common), !is.na(panel_score))

compera_cor <- score_compera_df %>%
  group_by(Cohort) %>%
  summarise(
    n = n(),
    rho = cor(COMPERA_number_common, panel_score, method = "spearman"),
    p = suppressWarnings(
      cor.test(COMPERA_number_common, panel_score, method = "spearman", exact = FALSE)$p.value
    ),
    .groups = "drop"
  )

compera_cor

p_score_compera <- ggplot(score_compera_df, aes(COMPERA, panel_score, fill = COMPERA)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.75, width = 0.6) +
  geom_jitter(width = 0.10, size = 2.0, alpha = 0.75) +
  facet_wrap(~Cohort, nrow = 1) +
  scale_fill_manual(values = cols_compera, drop = FALSE) +
  labs(
    x = "COMPERA risk stratum",
    y = "RV-PA coupling protein score"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 14),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )

p_score_compera
ggsave(here::here(output_dir_figs, "Figure4A_score_across_COMPERA.png"),
       p_score_compera, width = 8.5, height = 5, dpi = 300)
```

## Zurich 3-year survival data

```{r survival_prepare_3y}
df_surv_z_3y <- score_z %>%
  mutate(
    time_days_original = as.numeric(DeathDAYS),
    event_raw = tolower(trimws(as.character(Death))),
    event_original = case_when(
      event_raw %in% c("yes", "y", "true", "1", "dead", "death") ~ 1,
      event_raw %in% c("no", "n", "false", "0", "alive") ~ 0,
      TRUE ~ NA_real_
    )
  ) %>%
  filter(
    !is.na(time_days_original),
    !is.na(event_original),
    !is.na(COMPERA_number_common),
    !is.na(panel_score)
  ) %>%
  mutate(
    time_days_3y = pmin(time_days_original, followup_limit_days),
    event_3y = ifelse(event_original == 1 & time_days_original <= followup_limit_days, 1, 0)
  )

survival_3y_summary <- df_surv_z_3y %>%
  summarise(
    n = n(),
    deaths_total_followup = sum(event_original),
    deaths_within_3y = sum(event_3y),
    censored_before_3y = sum(event_original == 0 & time_days_original < followup_limit_days),
    followed_at_least_3y = sum(time_days_original >= followup_limit_days)
  )

survival_3y_summary

cat("3-year survival analysis n =", nrow(df_surv_z_3y),
    "deaths within 3 years =", sum(df_surv_z_3y$event_3y), "\n")

write.csv(survival_3y_summary,
          here::here(output_dir_data, "Survival_3year_summary_Zurich.csv"),
          row.names = FALSE)
```

## Cox models - 3 years

```{r survival_cox_3y}
m_compera_3y <- coxph(
  Surv(time_days_3y, event_3y) ~ COMPERA_number_common,
  data = df_surv_z_3y, x = TRUE
)

m_score_3y <- coxph(
  Surv(time_days_3y, event_3y) ~ panel_score,
  data = df_surv_z_3y, x = TRUE
)

m_combined_3y <- coxph(
  Surv(time_days_3y, event_3y) ~ COMPERA_number_common + panel_score,
  data = df_surv_z_3y, x = TRUE
)

c_compera_3y <- unname(summary(m_compera_3y)$concordance[1])
c_score_3y <- unname(summary(m_score_3y)$concordance[1])
c_combined_3y <- unname(summary(m_combined_3y)$concordance[1])

lrt_score_added_3y <- anova(m_compera_3y, m_combined_3y, test = "LRT")
lrt_score_added_3y

cox_table_3y <- bind_rows(
  tidy(m_compera_3y, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model = "COMPERA"),
  tidy(m_score_3y, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model = "Protein score"),
  tidy(m_combined_3y, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model = "COMPERA + protein score")
)

cox_table_3y

write.csv(cox_table_3y,
          here::here(output_dir_data, "Cox_models_Zurich_3year.csv"),
          row.names = FALSE)
```

## Bootstrap C-index and C-index differences - 3 years

```{r survival_bootstrap_cindex_3y}
set.seed(30)

B_surv_3y <- 2000

boot_cindex_3y <- matrix(
  NA_real_,
  nrow = B_surv_3y,
  ncol = 3,
  dimnames = list(NULL, c("COMPERA", "Score", "Combined"))
)

for (b in seq_len(B_surv_3y)){
  idx <- sample(seq_len(nrow(df_surv_z_3y)), replace = TRUE)
  dd <- df_surv_z_3y[idx, , drop = FALSE]

  if (length(unique(dd$event_3y)) < 2) next

  tmp <- tryCatch({
    m0 <- coxph(Surv(time_days_3y, event_3y) ~ COMPERA_number_common, data = dd)
    m1 <- coxph(Surv(time_days_3y, event_3y) ~ panel_score, data = dd)
    m2 <- coxph(Surv(time_days_3y, event_3y) ~ COMPERA_number_common + panel_score, data = dd)

    c(
      unname(summary(m0)$concordance[1]),
      unname(summary(m1)$concordance[1]),
      unname(summary(m2)$concordance[1])
    )
  }, error = function(e) c(NA, NA, NA))

  boot_cindex_3y[b, ] <- tmp
}

cindex_table_3y <- data.frame(
  Model = c("COMPERA", "Protein score", "COMPERA + protein score"),
  C_index = c(c_compera_3y, c_score_3y, c_combined_3y),
  CI_low = apply(boot_cindex_3y, 2, quantile, probs = 0.025, na.rm = TRUE),
  CI_high = apply(boot_cindex_3y, 2, quantile, probs = 0.975, na.rm = TRUE)
)

boot_delta_score_3y <- boot_cindex_3y[, "Score"] - boot_cindex_3y[, "COMPERA"]
boot_delta_combined_3y <- boot_cindex_3y[, "Combined"] - boot_cindex_3y[, "COMPERA"]

delta_cindex_table_3y <- data.frame(
  Comparison = c(
    "Protein score - COMPERA",
    "COMPERA + protein score - COMPERA"
  ),
  Delta_C = c(
    c_score_3y - c_compera_3y,
    c_combined_3y - c_compera_3y
  ),
  CI_low = c(
    quantile(boot_delta_score_3y, 0.025, na.rm = TRUE),
    quantile(boot_delta_combined_3y, 0.025, na.rm = TRUE)
  ),
  CI_high = c(
    quantile(boot_delta_score_3y, 0.975, na.rm = TRUE),
    quantile(boot_delta_combined_3y, 0.975, na.rm = TRUE)
  )
)

cindex_table_3y
delta_cindex_table_3y

write.csv(cindex_table_3y,
          here::here(output_dir_data, "Cindex_models_Zurich_3year.csv"),
          row.names = FALSE)

write.csv(delta_cindex_table_3y,
          here::here(output_dir_data, "Cindex_differences_Zurich_3year.csv"),
          row.names = FALSE)

cox.zph(m_combined_3y)
```

## Prognostic plots - 3 years

```{r cox_forest_3y, fig.height=5, fig.width=7, dpi=300}
forest_df_3y <- bind_rows(
  tidy(m_compera_3y, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model = "COMPERA only"),
  tidy(m_score_3y, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model = "Score only"),
  tidy(m_combined_3y, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model = "Combined")
) %>%
  mutate(
    Label = case_when(
      Model == "COMPERA only" & term == "COMPERA_number_common" ~ "COMPERA",
      Model == "Score only" & term == "panel_score" ~ "Protein score",
      Model == "Combined" & term == "COMPERA_number_common" ~ "COMPERA | combined",
      Model == "Combined" & term == "panel_score" ~ "Protein score | combined",
      TRUE ~ term
    )
  )

p_cox_forest_3y <- ggplot(forest_df_3y, aes(estimate, Label)) +
  geom_vline(xintercept = 1, linetype = 2, color = "grey65") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.18) +
  geom_point(size = 3.0) +
  scale_x_log10() +
  labs(x = "Hazard ratio", y = NULL, title = "3-year all-cause mortality") +
  theme_classic(base_size = 14)

p_cox_forest_3y
ggsave(here::here(output_dir_figs, "Figure4B_Cox_forest_3year.png"),
       p_cox_forest_3y, width = 7, height = 5, dpi = 300)
```


```{r figure4C_cindex, fig.height=4.8, fig.width=6.5, dpi=300}
p_cindex_3y <- cindex_table_3y %>%
  mutate(Model = factor(Model, levels = rev(Model))) %>%
  ggplot(aes(C_index, Model)) +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high), height = 0.15) +
  geom_point(size = 3.2) +
  coord_cartesian(xlim = c(0.5, 1)) +
  labs(x = "C-index", y = NULL, title = "3-year model discrimination") +
  theme_classic(base_size = 14)

p_cindex_3y
ggsave(here::here(output_dir_figs, "Figure4C_Cindex.png"),
       p_cindex_3y, width = 6.5, height = 4.8, dpi = 300)
```

```{r figure4D_delta_cindex, fig.height=4.8, fig.width=7.5, dpi=300}
p_delta_cindex_3y <- delta_cindex_table_3y %>%
  mutate(Comparison = factor(Comparison, levels = rev(Comparison))) %>%
  ggplot(aes(Delta_C, Comparison)) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey65") +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high), height = 0.15) +
  geom_point(size = 3.2) +
  labs(x = "Delta C-index", y = NULL, title = "Difference from COMPERA") +
  theme_classic(base_size = 14)

p_delta_cindex_3y
ggsave(here::here(output_dir_figs, "Figure4D_delta_Cindex.png"),
       p_delta_cindex_3y, width = 7.5, height = 4.8, dpi = 300)
```

```{r figure4_compera_mortality_3y, fig.height=8.8, fig.width=12, dpi=300}
Figure4 <- (p_score_compera | p_cox_forest_3y) /
  (p_cindex_3y | p_delta_cindex_3y) +
  plot_annotation(tag_levels = "A")

Figure4
ggsave(here::here(output_dir_figs, "Figure4_panel_COMPERA_mortality.png"),
       Figure4, width = 12, height = 8.8, dpi = 300)
```

## Kaplan-Meier plot - 3 years

```{r km_score_3y, fig.height=4.8, fig.width=5.8, dpi=300}
km_df_3y <- df_surv_z_3y %>%
  mutate(
    score_group = ifelse(panel_score >= median(panel_score), "High score", "Low score"),
    score_group = factor(score_group, levels = c("Low score", "High score"))
  )

km_fit_3y <- survfit(
  Surv(time_days_3y, event_3y) ~ score_group,
  data = km_df_3y
)

logrank_p_3y <- 1 - pchisq(
  survdiff(Surv(time_days_3y, event_3y) ~ score_group, data = km_df_3y)$chisq,
  df = 1
)

plot(
  km_fit_3y,
  col = c("#004C8C", "#A63A2A"),
  lwd = 2,
  xlim = c(0, followup_limit_days),
  xaxt = "n",
  xlab = "Follow-up (years)",
  ylab = "Survival probability",
  main = "3-year survival by protein score"
)

axis(
  1,
  at = c(0, 365.25, 2 * 365.25, 3 * 365.25),
  labels = c("0", "1", "2", "3")
)

legend(
  "bottomleft",
  legend = levels(km_df_3y$score_group),
  col = c("#004C8C", "#A63A2A"),
  lwd = 2,
  bty = "n"
)

mtext(
  paste0("Log-rank p = ", signif(logrank_p_3y, 3)),
  side = 3, adj = 1, line = 0.2, cex = 0.9
)

png(
  here::here(output_dir_figs, "Supplement_KM_score_Zurich_3year.png"),
  width = 5.8, height = 4.8, units = "in", res = 300
)

plot(
  km_fit_3y,
  col = c("#004C8C", "#A63A2A"),
  lwd = 2,
  xlim = c(0, followup_limit_days),
  xaxt = "n",
  xlab = "Follow-up (years)",
  ylab = "Survival probability",
  main = "3-year survival by protein score"
)

axis(
  1,
  at = c(0, 365.25, 2 * 365.25, 3 * 365.25),
  labels = c("0", "1", "2", "3")
)

legend(
  "bottomleft",
  legend = levels(km_df_3y$score_group),
  col = c("#004C8C", "#A63A2A"),
  lwd = 2,
  bty = "n"
)

mtext(
  paste0("Log-rank p = ", signif(logrank_p_3y, 3)),
  side = 3, adj = 1, line = 0.2, cex = 0.9
)

dev.off()
```

# Supplementary: protein score and NT-proBNP

This section compares the final protein score with NT-proBNP as a marker of impaired RV-PA coupling and explores whether the protein score can replace or complement NT-proBNP within the clinical components used by COMPERA 2.0. These analyses are exploratory and are intended for supplementary material.

## Protein score and NT-proBNP vs TAPSE/sPAP

For direct comparison, the same SSc-PAH patients with available TAPSE/sPAP, protein score and NT-proBNP are used for both biomarkers.

```{r supplementary_ntprobnp_prepare}
rvpa_biomarker_df <- bind_rows(
  score_p %>% transmute(Cohort = "Krakow", Barcode, TAPSE_sPAP, panel_score, NTproBNP_common),
  score_z %>% transmute(Cohort = "Zurich", Barcode, TAPSE_sPAP, panel_score, NTproBNP_common)
) %>%
  filter(!is.na(TAPSE_sPAP), !is.na(panel_score), !is.na(NTproBNP_common), NTproBNP_common > 0) %>%
  mutate(y_impaired = ifelse(TAPSE_sPAP <= 0.32, 1, 0), log2_NTproBNP = log2(NTproBNP_common))

table(rvpa_biomarker_df$Cohort, rvpa_biomarker_df$y_impaired)
```


```{r supplementary_ntprobnp_correlations}
nt_cor_results <- list()

for (cohort in c("Krakow", "Zurich")){
  dd <- rvpa_biomarker_df %>% filter(Cohort == cohort)

  ct <- cor.test(
    dd$panel_score,
    dd$log2_NTproBNP,
    method = "spearman",
    exact = FALSE
  )

  nt_cor_results[[cohort]] <- data.frame(
    Cohort = cohort,
    n = nrow(dd),
    rho = unname(ct$estimate),
    p = ct$p.value
  )
}

nt_cor_results <- bind_rows(nt_cor_results)
nt_cor_results

write.csv(nt_cor_results,
          here::here(output_dir_data, "Supplement_protein_score_NTproBNP_correlations.csv"),
          row.names = FALSE)
```

```{r supplementary_ntprobnp_correlation_plot, fig.width=8.5, fig.height=4.8, dpi=300}
nt_cor_plot_df <- rvpa_biomarker_df %>% mutate(Cohort = factor(Cohort, levels = c("Krakow", "Zurich")))
nt_cor_labels <- nt_cor_results %>% mutate(Cohort = factor(Cohort, levels = c("Krakow", "Zurich")), label = paste0("rho = ", round(rho, 2), "\np = ", signif(p, 2)))
p_nt_cor <- ggplot(nt_cor_plot_df, aes(log2_NTproBNP, panel_score)) +
  geom_point(size = 2.1, alpha = 0.78) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.9, color = "#3568C8") +
  facet_wrap(~Cohort, nrow = 1) +
  geom_text(data = nt_cor_labels, aes(x = -Inf, y = Inf, label = label), inherit.aes = FALSE, hjust = -0.03, vjust = 1.08, size = 4.1) +
  labs(x = "log2 NT-proBNP", y = "RV-PA coupling protein score", title = "Protein score and NT-proBNP") +
  theme_classic(base_size = 14) + theme(strip.background = element_rect(fill = "white"), strip.text = element_text(size = 14))
p_nt_cor
ggsave(here::here(output_dir_figs, "SupplementaryFigure5A_score_NTproBNP_correlation.png"), p_nt_cor, width = 8.5, height = 4.8, dpi = 300)
```

## Discrimination of impaired RV-PA coupling

```{r supplementary_ntprobnp_auc}
nt_auc_results <- list()
nt_delong_results <- list()
nt_roc_score <- list()
nt_roc_nt <- list()

for (cohort in c("Krakow", "Zurich")){
  dd <- rvpa_biomarker_df %>% filter(Cohort == cohort)

  roc_score <- pROC::roc(dd$y_impaired, dd$panel_score, levels = c(0, 1), direction = "<", quiet = TRUE, ci = TRUE)
  roc_nt <- pROC::roc(dd$y_impaired, dd$NTproBNP_common, levels = c(0, 1), direction = "<", quiet = TRUE, ci = TRUE)
  ci_score <- pROC::ci.auc(roc_score)
  ci_nt <- pROC::ci.auc(roc_nt)
  test_auc <- pROC::roc.test(roc_score, roc_nt, paired = TRUE, method = "delong")

  nt_auc_results[[cohort]] <- bind_rows(
    data.frame(Cohort = cohort, Biomarker = "Protein score", n = nrow(dd), AUC = as.numeric(pROC::auc(roc_score)), CI_low = as.numeric(ci_score[1]), CI_high = as.numeric(ci_score[3])),
    data.frame(Cohort = cohort, Biomarker = "NT-proBNP", n = nrow(dd), AUC = as.numeric(pROC::auc(roc_nt)), CI_low = as.numeric(ci_nt[1]), CI_high = as.numeric(ci_nt[3]))
  )

  nt_delong_results[[cohort]] <- data.frame(Cohort = cohort, Delta_AUC = as.numeric(pROC::auc(roc_score) - pROC::auc(roc_nt)), DeLong_p = test_auc$p.value)
  nt_roc_score[[cohort]] <- roc_score
  nt_roc_nt[[cohort]] <- roc_nt
}

nt_auc_results <- bind_rows(nt_auc_results)
nt_delong_results <- bind_rows(nt_delong_results)
nt_auc_results
nt_delong_results
write.csv(nt_auc_results, here::here(output_dir_data, "Supplement_protein_score_vs_NTproBNP_AUC.csv"), row.names = FALSE)
write.csv(nt_delong_results, here::here(output_dir_data, "Supplement_protein_score_vs_NTproBNP_DeLong.csv"), row.names = FALSE)


nt_roc_bands <- bind_rows(
  roc_bootstrap_band(
    rvpa_biomarker_df$y_impaired[rvpa_biomarker_df$Cohort == "Krakow"],
    rvpa_biomarker_df$panel_score[rvpa_biomarker_df$Cohort == "Krakow"],
    B = 2000, seed = 301
  ) %>% mutate(Cohort = "Krakow", Biomarker = "Protein score"),
  roc_bootstrap_band(
    rvpa_biomarker_df$y_impaired[rvpa_biomarker_df$Cohort == "Krakow"],
    rvpa_biomarker_df$NTproBNP_common[rvpa_biomarker_df$Cohort == "Krakow"],
    B = 2000, seed = 302
  ) %>% mutate(Cohort = "Krakow", Biomarker = "NT-proBNP"),
  roc_bootstrap_band(
    rvpa_biomarker_df$y_impaired[rvpa_biomarker_df$Cohort == "Zurich"],
    rvpa_biomarker_df$panel_score[rvpa_biomarker_df$Cohort == "Zurich"],
    B = 2000, seed = 303
  ) %>% mutate(Cohort = "Zurich", Biomarker = "Protein score"),
  roc_bootstrap_band(
    rvpa_biomarker_df$y_impaired[rvpa_biomarker_df$Cohort == "Zurich"],
    rvpa_biomarker_df$NTproBNP_common[rvpa_biomarker_df$Cohort == "Zurich"],
    B = 2000, seed = 304
  ) %>% mutate(Cohort = "Zurich", Biomarker = "NT-proBNP")
)
```

```{r supplementary_ntprobnp_roc, fig.width=9, fig.height=4.8, dpi=300}
nt_roc_df <- bind_rows(
  roc_curve_df(
    rvpa_biomarker_df$y_impaired[rvpa_biomarker_df$Cohort == "Krakow"],
    rvpa_biomarker_df$panel_score[rvpa_biomarker_df$Cohort == "Krakow"]
  ) %>% mutate(Cohort = "Krakow", Biomarker = "Protein score"),
  roc_curve_df(
    rvpa_biomarker_df$y_impaired[rvpa_biomarker_df$Cohort == "Krakow"],
    rvpa_biomarker_df$NTproBNP_common[rvpa_biomarker_df$Cohort == "Krakow"]
  ) %>% mutate(Cohort = "Krakow", Biomarker = "NT-proBNP"),
  roc_curve_df(
    rvpa_biomarker_df$y_impaired[rvpa_biomarker_df$Cohort == "Zurich"],
    rvpa_biomarker_df$panel_score[rvpa_biomarker_df$Cohort == "Zurich"]
  ) %>% mutate(Cohort = "Zurich", Biomarker = "Protein score"),
  roc_curve_df(
    rvpa_biomarker_df$y_impaired[rvpa_biomarker_df$Cohort == "Zurich"],
    rvpa_biomarker_df$NTproBNP_common[rvpa_biomarker_df$Cohort == "Zurich"]
  ) %>% mutate(Cohort = "Zurich", Biomarker = "NT-proBNP")
)

p_nt_roc <- ggplot() +
  geom_ribbon(
    data = nt_roc_bands,
    aes(x = FPR, ymin = lower, ymax = upper, fill = Biomarker),
    alpha = 0.24
  ) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey70") +
  geom_step(
    data = nt_roc_df,
    aes(FPR, Sensitivity, color = Biomarker),
    direction = "hv",
    linewidth = 1.0
  ) +
  facet_wrap(~Cohort, nrow = 1) +
  coord_equal() +
  scale_color_manual(values = c("Protein score" = "#B23A2B", "NT-proBNP" = "#2C6E9B")) +
  scale_fill_manual(values = c("Protein score" = "#E5A49A", "NT-proBNP" = "#A8C5D5")) +
  labs(
    x = "1 - Specificity",
    y = "Sensitivity",
    color = NULL,
    fill = NULL,
    title = "Impaired RV-PA coupling"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 14)
  )

p_nt_roc
ggsave(here::here(output_dir_figs, "SupplementaryFigure5B_NTproBNP_RVPA_ROC.png"),
       p_nt_roc, width = 9, height = 4.8, dpi = 300)
```

```{r supplementary_figure5_ntprobnp, fig.width=10, fig.height=9, dpi=300}
SupplementaryFigure5 <- p_nt_cor / p_nt_roc +
  plot_annotation(tag_levels = "A")

SupplementaryFigure5
ggsave(here::here(output_dir_figs, "SupplementaryFigure5_NTproBNP_comparison.png"),
       SupplementaryFigure5, width = 10, height = 9, dpi = 300)
```

## NT-proBNP replacement within the COMPERA clinical components - 3 years

This repeats the NT-proBNP replacement/complement analysis using the same 3-year mortality horizon.

## Prepare common Zurich cohort

```{r supplementary_compera_components_prepare_3y}
compera_components_z_3y <- score_z %>%
  mutate(
    time_days_original = as.numeric(DeathDAYS),
    event_raw = tolower(trimws(as.character(Death))),
    event_original = case_when(
      event_raw %in% c("yes", "y", "true", "1", "dead", "death") ~ 1,
      event_raw %in% c("no", "n", "false", "0", "alive") ~ 0,
      TRUE ~ NA_real_
    ),
    log2_NTproBNP = ifelse(NTproBNP_common > 0, log2(NTproBNP_common), NA_real_),
    SixMWD_100m = SixMWD / 100
  ) %>%
  filter(
    !is.na(time_days_original),
    !is.na(event_original),
    !is.na(WHO_FC),
    !is.na(SixMWD_100m),
    !is.na(log2_NTproBNP),
    !is.na(panel_score),
    !is.na(COMPERA_number_common)
  ) %>%
  mutate(
    time_days_3y = pmin(time_days_original, followup_limit_days),
    event_3y = ifelse(event_original == 1 & time_days_original <= followup_limit_days, 1, 0)
  )

cat(
  "3-year COMPERA component analysis n =", nrow(compera_components_z_3y),
  "deaths within 3 years =", sum(compera_components_z_3y$event_3y), "\n"
)
```

## COMPERA component models - 3 years

```{r supplementary_compera_components_models_3y}
m_components_nt_3y <- coxph(
  Surv(time_days_3y, event_3y) ~ WHO_FC + SixMWD_100m + log2_NTproBNP,
  data = compera_components_z_3y, x = TRUE
)

m_components_protein_3y <- coxph(
  Surv(time_days_3y, event_3y) ~ WHO_FC + SixMWD_100m + panel_score,
  data = compera_components_z_3y, x = TRUE
)

m_components_both_3y <- coxph(
  Surv(time_days_3y, event_3y) ~ WHO_FC + SixMWD_100m + log2_NTproBNP + panel_score,
  data = compera_components_z_3y, x = TRUE
)

m_compera_same_n_3y <- coxph(
  Surv(time_days_3y, event_3y) ~ COMPERA_number_common,
  data = compera_components_z_3y, x = TRUE
)

c_compera_same_n_3y <- unname(summary(m_compera_same_n_3y)$concordance[1])
c_nt_3y <- unname(summary(m_components_nt_3y)$concordance[1])
c_protein_component_3y <- unname(summary(m_components_protein_3y)$concordance[1])
c_both_component_3y <- unname(summary(m_components_both_3y)$concordance[1])

component_models_3y <- data.frame(
  Model = c(
    "COMPERA 2.0 risk stratum",
    "WHO-FC + 6MWD + NT-proBNP",
    "WHO-FC + 6MWD + protein score",
    "WHO-FC + 6MWD + both biomarkers"
  ),
  C_index = c(
    c_compera_same_n_3y,
    c_nt_3y,
    c_protein_component_3y,
    c_both_component_3y
  )
)

component_models_3y

lrt_add_protein_3y <- anova(
  m_components_nt_3y,
  m_components_both_3y,
  test = "LRT"
)

lrt_add_ntprobnp_3y <- anova(
  m_components_protein_3y,
  m_components_both_3y,
  test = "LRT"
)

lrt_add_protein_3y
lrt_add_ntprobnp_3y

write.csv(
  component_models_3y,
  here::here(output_dir_data, "Supplement_COMPERA_biomarker_component_Cindex_3year.csv"),
  row.names = FALSE
)
```

## Bootstrap COMPERA component C-index - 3 years

```{r supplementary_compera_components_bootstrap_3y}
set.seed(402)

B_components_3y <- 2000

boot_components_3y <- matrix(
  NA_real_,
  nrow = B_components_3y,
  ncol = 4,
  dimnames = list(NULL, c("COMPERA", "NTproBNP", "Protein", "Both"))
)

for (b in seq_len(B_components_3y)){
  idx <- sample(seq_len(nrow(compera_components_z_3y)), replace = TRUE)
  dd <- compera_components_z_3y[idx, , drop = FALSE]

  if (length(unique(dd$event_3y)) < 2) next

  vals <- tryCatch({
    m0 <- coxph(Surv(time_days_3y, event_3y) ~ COMPERA_number_common, data = dd)
    m1 <- coxph(Surv(time_days_3y, event_3y) ~ WHO_FC + SixMWD_100m + log2_NTproBNP, data = dd)
    m2 <- coxph(Surv(time_days_3y, event_3y) ~ WHO_FC + SixMWD_100m + panel_score, data = dd)
    m3 <- coxph(Surv(time_days_3y, event_3y) ~ WHO_FC + SixMWD_100m + log2_NTproBNP + panel_score, data = dd)

    c(
      unname(summary(m0)$concordance[1]),
      unname(summary(m1)$concordance[1]),
      unname(summary(m2)$concordance[1]),
      unname(summary(m3)$concordance[1])
    )
  }, error = function(e) rep(NA_real_, 4))

  boot_components_3y[b, ] <- vals
}

component_cindex_ci_3y <- data.frame(
  Model = c(
    "COMPERA 2.0 risk stratum",
    "WHO-FC + 6MWD + NT-proBNP",
    "WHO-FC + 6MWD + protein score",
    "WHO-FC + 6MWD + both biomarkers"
  ),
  C_index = c(
    c_compera_same_n_3y,
    c_nt_3y,
    c_protein_component_3y,
    c_both_component_3y
  ),
  CI_low = apply(boot_components_3y, 2, quantile, 0.025, na.rm = TRUE),
  CI_high = apply(boot_components_3y, 2, quantile, 0.975, na.rm = TRUE)
)

boot_delta_replace_3y <- boot_components_3y[, "Protein"] - boot_components_3y[, "NTproBNP"]
boot_delta_add_3y <- boot_components_3y[, "Both"] - boot_components_3y[, "NTproBNP"]

component_delta_3y <- data.frame(
  Comparison = c(
    "Protein score instead of NT-proBNP",
    "Protein score added to NT-proBNP"
  ),
  Delta_C = c(
    c_protein_component_3y - c_nt_3y,
    c_both_component_3y - c_nt_3y
  ),
  CI_low = c(
    quantile(boot_delta_replace_3y, 0.025, na.rm = TRUE),
    quantile(boot_delta_add_3y, 0.025, na.rm = TRUE)
  ),
  CI_high = c(
    quantile(boot_delta_replace_3y, 0.975, na.rm = TRUE),
    quantile(boot_delta_add_3y, 0.975, na.rm = TRUE)
  )
)

component_cindex_ci_3y
component_delta_3y

write.csv(
  component_cindex_ci_3y,
  here::here(output_dir_data, "Supplement_COMPERA_biomarker_component_Cindex_bootstrap_3year.csv"),
  row.names = FALSE
)

write.csv(
  component_delta_3y,
  here::here(output_dir_data, "Supplement_COMPERA_biomarker_component_delta_Cindex_3year.csv"),
  row.names = FALSE
)
```

## COMPERA component plots - 3 years



# Secondary analysis: PAH diagnosis and DETECT

## SSc-PAH vs SSc-noPAH - Zurich

```{r de_group_zurich}
meta_z_group <- z_all %>%
  filter(!is.na(Group), !is.na(Age), !is.na(Sex)) %>%
  mutate(
    Group = factor(Group, levels = c("SSc-noPAH", "SSc-PAH")),
    Age_z = as.numeric(scale(Age))
  ) %>%
  column_to_rownames("Barcode")

expr_z_group <- as.matrix(data_z[, rownames(meta_z_group), drop = FALSE])
design_z_group <- model.matrix(~ Group + Age_z + Sex, data = meta_z_group)
fit_z_group <- eBayes(lmFit(expr_z_group, design_z_group))

DE_group_z <- topTable(
  fit_z_group, coef = "GroupSSc-PAH",
  number = Inf, adjust.method = "BH"
) %>% rownames_to_column("Protein")

write.csv(DE_group_z, here::here(output_dir_data, "DE_SScPAH_vs_noPAH_Zurich.csv"), row.names = FALSE)
```

## SSc-PAH vs SSc-noPAH - Krakow

```{r de_group_poland}
meta_p_group <- p_all %>%
  filter(!is.na(Group), !is.na(Age), !is.na(Sex)) %>%
  mutate(
    Group = factor(Group, levels = c("SSc-noPAH", "SSc-PAH")),
    Age_z = as.numeric(scale(Age))
  ) %>%
  column_to_rownames("Barcode")

expr_p_group <- as.matrix(data_p[, rownames(meta_p_group), drop = FALSE])
design_p_group <- model.matrix(~ Group + Age_z + Sex, data = meta_p_group)
fit_p_group <- eBayes(lmFit(expr_p_group, design_p_group))

DE_group_p <- topTable(
  fit_p_group, coef = "GroupSSc-PAH",
  number = Inf, adjust.method = "BH"
) %>% rownames_to_column("Protein")

write.csv(DE_group_p, here::here(output_dir_data, "DE_SScPAH_vs_noPAH_Krakow.csv"), row.names = FALSE)
```

```{r detect_plot_helpers}
make_pah_volcano <- function(res, title){
  keyvals <- rep("grey75", nrow(res))
  names(keyvals) <- rep("Not significant", nrow(res))
  keyvals[res$adj.P.Val < 0.1 & res$logFC > 0] <- "#B23A2B"
  keyvals[res$adj.P.Val < 0.1 & res$logFC < 0] <- "#2C6E9B"
  names(keyvals)[keyvals == "#B23A2B"] <- "Higher in SSc-PAH"
  names(keyvals)[keyvals == "#2C6E9B"] <- "Lower in SSc-PAH"

  EnhancedVolcano(
    res,
    lab = res$Protein,
    x = "logFC",
    y = "adj.P.Val",
    labSize = 0,
    pointSize = 1.5,
    pCutoff = 0.1,
    FCcutoff = 0,
    colCustom = keyvals,
    colAlpha = 1,
    subtitle = NULL,
    title = title,
    xlab = "NPX difference (SSc-PAH - SSc-noPAH)"
  ) +
    theme(
      legend.position = "none",
      text = element_text(size = 14),
      plot.title = element_text(size = 16),
      axis.title = element_text(size = 14)
    )
}

concordance_group <- inner_join(
  DE_group_z %>% select(Protein, effect_z = logFC),
  DE_group_p %>% select(Protein, effect_p = logFC),
  by = "Protein"
)

pear_group <- cor.test(
  concordance_group$effect_z,
  concordance_group$effect_p,
  method = "pearson"
)

spear_group <- cor.test(
  concordance_group$effect_z,
  concordance_group$effect_p,
  method = "spearman",
  exact = FALSE
)
```

```{r figure5B_zurich_volcano, fig.height=5, fig.width=5.8, dpi=300}
p_volcano_group_z <- make_pah_volcano(DE_group_z, "Zurich")
p_volcano_group_z
ggsave(here::here(output_dir_figs, "Figure5B_Zurich_PAH_noPAH_volcano.png"),
       p_volcano_group_z, width = 5.8, height = 5, dpi = 300)
```

```{r supplementary6A_krakow_volcano, fig.height=5, fig.width=5.8, dpi=300}
p_volcano_group_p <- make_pah_volcano(DE_group_p, "Krakow")
p_volcano_group_p
ggsave(here::here(output_dir_figs, "SupplementaryFigure6A_Krakow_PAH_noPAH_volcano.png"),
       p_volcano_group_p, width = 5.8, height = 5, dpi = 300)
```

```{r supplementary6B_pah_concordance, fig.height=5, fig.width=5.8, dpi=300}
p_concordance_group <- ggplot(concordance_group, aes(effect_z, effect_p)) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "grey75") +
  geom_vline(xintercept = 0, linewidth = 0.4, color = "grey75") +
  geom_point(size = 1.6, alpha = 0.6) +
  geom_smooth(method = MASS::rlm, formula = y ~ x, se = FALSE, linewidth = 0.9, color = "grey30") +
  labs(
    x = "Zurich effect",
    y = "Krakow effect",
    title = paste0(
      "SSc-PAH vs SSc-noPAH\n",
      "Pearson r = ", round(pear_group$estimate, 2),
      "; Spearman rho = ", round(spear_group$estimate, 2)
    )
  ) +
  theme_classic(base_size = 14)

p_concordance_group
ggsave(here::here(output_dir_figs, "SupplementaryFigure6B_PAH_noPAH_concordance.png"),
       p_concordance_group, width = 5.8, height = 5, dpi = 300)
```



## DETECT improvement - nested-CV elastic net

No differential-abundance preselection is used for this analysis. Proteomics models use all proteins measured in both cohorts. Alpha and lambda are tuned in the inner cross-validation. For the combined model, the DETECT score is forced into the model and only protein coefficients are penalized.

```{r detect_helpers}
stratified_folds_binary <- function(y, k = 5){
  folds <- integer(length(y))
  for (lev in sort(unique(y))){
    idx <- sample(which(y == lev))
    folds[idx] <- rep(seq_len(k), length.out = length(idx))
  }
  folds
}

specificity_at_sensitivity <- function(y, score, target_sensitivity = 0.95){
  ok <- is.finite(y) & is.finite(score)
  y <- y[ok]
  score <- score[ok]

  thresholds <- sort(unique(c(Inf, score, -Inf)), decreasing = TRUE)
  out <- lapply(thresholds, function(th){
    pred <- ifelse(score >= th, 1, 0)
    sens <- sum(pred == 1 & y == 1) / sum(y == 1)
    spec <- sum(pred == 0 & y == 0) / sum(y == 0)
    c(threshold = th, sensitivity = sens, specificity = spec)
  }) %>% do.call(rbind, .) %>% as.data.frame()

  out <- out[out$sensitivity >= target_sensitivity, , drop = FALSE]
  if (nrow(out) == 0) return(c(threshold = NA, sensitivity = NA, specificity = NA))

  best <- out[out$specificity == max(out$specificity), , drop = FALSE]
  best <- best[which.max(best$threshold), , drop = FALSE]
  unlist(best[1, ])
}

calc_auc <- function(y, score){
  as.numeric(pROC::auc(
    pROC::roc(y, score, levels = c(0, 1), quiet = TRUE, direction = "<")
  ))
}

fit_elastic_net_inner <- function(X, y, foldid, alphas = c(0, 0.25, 0.5, 0.75, 1), penalty.factor = NULL){
  if (is.null(penalty.factor)){
    penalty.factor <- rep(1, ncol(X))
  }

  fits <- vector("list", length(alphas))
  cv_auc <- rep(NA_real_, length(alphas))

  for (i in seq_along(alphas)){
    fits[[i]] <- try(
      cv.glmnet(
        X, y,
        family = "binomial",
        alpha = alphas[i],
        foldid = foldid,
        type.measure = "auc",
        standardize = TRUE,
        nlambda = 100,
        penalty.factor = penalty.factor
      ),
      silent = TRUE
    )

    if (!inherits(fits[[i]], "try-error") && any(is.finite(fits[[i]]$cvm))){
      cv_auc[i] <- max(fits[[i]]$cvm, na.rm = TRUE)
    }
  }

  if (all(!is.finite(cv_auc))) return(NULL)

  best_i <- which.max(cv_auc)
  best_fit <- fits[[best_i]]
  best_lambda_i <- which.max(best_fit$cvm)

  list(
    fit = best_fit,
    alpha = alphas[best_i],
    lambda = best_fit$lambda[best_lambda_i],
    cv_auc = best_fit$cvm[best_lambda_i]
  )
}
```

```{r detect_crossvalidation}
detect_df <- z_all %>%
  filter(
    Group %in% c("SSc-noPAH", "SSc-PAH"),
    !is.na(DETECT_score),
    Barcode %in% colnames(data_z)
  ) %>%
  mutate(y = ifelse(Group == "SSc-PAH", 1, 0))

X_detect_prot <- t(as.matrix(data_z[common_proteins, detect_df$Barcode, drop = FALSE]))
y_detect <- detect_df$y
detect_score <- detect_df$DETECT_score

set.seed(1)
R_detect <- 50
K_detect <- 5
alpha_grid <- c(0, 0.25, 0.5, 0.75, 1)

pred_sum_detect <- matrix(
  0, nrow = nrow(detect_df), ncol = 3,
  dimnames = list(
    detect_df$Barcode,
    c("DETECT", "Proteomics_elastic_net", "DETECT_plus_proteomics_elastic_net")
  )
)

pred_n_detect <- matrix(
  0, nrow = nrow(detect_df), ncol = 3,
  dimnames = dimnames(pred_sum_detect)
)

detect_tuning_list <- list()
counter_detect <- 1

for (r in seq_len(R_detect)){
  outer_folds <- stratified_folds_binary(y_detect, K_detect)

  for (f in seq_len(K_detect)){
    tr <- which(outer_folds != f)
    te <- which(outer_folds == f)

    p_detect <- detect_score[te]

    inner_k <- min(5, min(table(y_detect[tr])))
    if (inner_k < 3) next
    inner_fold <- stratified_folds_binary(y_detect[tr], inner_k)

    prot_fit <- fit_elastic_net_inner(
      X_detect_prot[tr, , drop = FALSE],
      y_detect[tr],
      foldid = inner_fold,
      alphas = alpha_grid
    )

    X_comb_tr <- cbind(DETECT = detect_score[tr], X_detect_prot[tr, , drop = FALSE])
    X_comb_te <- cbind(DETECT = detect_score[te], X_detect_prot[te, , drop = FALSE])
    penalty_comb <- c(0, rep(1, ncol(X_detect_prot)))

    comb_fit <- fit_elastic_net_inner(
      X_comb_tr,
      y_detect[tr],
      foldid = inner_fold,
      alphas = alpha_grid,
      penalty.factor = penalty_comb
    )

    if (is.null(prot_fit) || is.null(comb_fit)) next

    p_prot <- as.numeric(
      predict(prot_fit$fit$glmnet.fit, newx = X_detect_prot[te, , drop = FALSE],
              s = prot_fit$lambda, type = "response")
    )

    p_comb <- as.numeric(
      predict(comb_fit$fit$glmnet.fit, newx = X_comb_te,
              s = comb_fit$lambda, type = "response")
    )

    pred_sum_detect[te, "DETECT"] <- pred_sum_detect[te, "DETECT"] + p_detect
    pred_sum_detect[te, "Proteomics_elastic_net"] <- pred_sum_detect[te, "Proteomics_elastic_net"] + p_prot
    pred_sum_detect[te, "DETECT_plus_proteomics_elastic_net"] <- pred_sum_detect[te, "DETECT_plus_proteomics_elastic_net"] + p_comb
    pred_n_detect[te, ] <- pred_n_detect[te, ] + 1

    beta_prot <- as.matrix(coef(prot_fit$fit$glmnet.fit, s = prot_fit$lambda))[-1, 1]
    beta_comb <- as.matrix(coef(comb_fit$fit$glmnet.fit, s = comb_fit$lambda))[-1, 1]
    beta_comb_proteins <- beta_comb[-1]

    detect_tuning_list[[counter_detect]] <- data.frame(
      rep_id = r,
      fold = f,
      proteomics_alpha = prot_fit$alpha,
      proteomics_inner_AUC = prot_fit$cv_auc,
      proteomics_nonzero = sum(beta_prot != 0),
      combined_alpha = comb_fit$alpha,
      combined_inner_AUC = comb_fit$cv_auc,
      combined_nonzero_proteins = sum(beta_comb_proteins != 0)
    )
    counter_detect <- counter_detect + 1
  }
}

detect_oof <- pred_sum_detect / pred_n_detect
detect_tuning <- bind_rows(detect_tuning_list)

detect_pred_df <- data.frame(
  Barcode = detect_df$Barcode,
  y = y_detect,
  DETECT = detect_oof[, "DETECT"],
  Proteomics = detect_oof[, "Proteomics_elastic_net"],
  DETECT_plus_proteomics = detect_oof[, "DETECT_plus_proteomics_elastic_net"]
)

write.csv(detect_pred_df, here::here(output_dir_data, "DETECT_repeated_nestedCV_predictions.csv"), row.names = FALSE)
write.csv(detect_tuning, here::here(output_dir_data, "DETECT_elastic_net_tuning.csv"), row.names = FALSE)

detect_tuning %>%
  summarise(
    proteomics_alpha_median = median(proteomics_alpha),
    proteomics_nonzero_median = median(proteomics_nonzero),
    combined_alpha_median = median(combined_alpha),
    combined_nonzero_proteins_median = median(combined_nonzero_proteins)
  )
```

```{r detect_performance}
get_detect_perf <- function(score, model_name){
  ss <- specificity_at_sensitivity(y_detect, score, 0.95)
  data.frame(
    Model = model_name,
    AUC = calc_auc(y_detect, score),
    Sensitivity = unname(ss["sensitivity"]),
    Specificity_at_95_sensitivity = unname(ss["specificity"]),
    Threshold = unname(ss["threshold"]),
    Negative_RHCs_avoided = round(unname(ss["specificity"]) * sum(y_detect == 0))
  )
}

detect_perf <- bind_rows(
  get_detect_perf(detect_oof[, "DETECT"], "DETECT"),
  get_detect_perf(detect_oof[, "Proteomics_elastic_net"], "Proteomics elastic net"),
  get_detect_perf(detect_oof[, "DETECT_plus_proteomics_elastic_net"], "DETECT + proteomics elastic net")
)

set.seed(2)
B_detect <- 2000
boot_detect <- vector("list", B_detect)

for (b in seq_len(B_detect)){
  idx <- sample(seq_len(nrow(detect_df)), replace = TRUE)
  yb <- y_detect[idx]
  if (length(unique(yb)) < 2) next

  vals <- lapply(
    c("DETECT", "Proteomics_elastic_net", "DETECT_plus_proteomics_elastic_net"),
    function(m){
      sc <- detect_oof[idx, m]
      ss <- specificity_at_sensitivity(yb, sc, 0.95)
      c(AUC = calc_auc(yb, sc), Spec95 = unname(ss["specificity"]))
    }
  )

  boot_detect[[b]] <- data.frame(
    DETECT_AUC = vals[[1]]["AUC"],
    Proteomics_AUC = vals[[2]]["AUC"],
    Combined_AUC = vals[[3]]["AUC"],
    DETECT_Spec95 = vals[[1]]["Spec95"],
    Proteomics_Spec95 = vals[[2]]["Spec95"],
    Combined_Spec95 = vals[[3]]["Spec95"]
  )
}

boot_detect <- bind_rows(boot_detect)

detect_ci <- data.frame(
  Model = c("DETECT", "Proteomics elastic net", "DETECT + proteomics elastic net"),
  AUC_low = c(
    quantile(boot_detect$DETECT_AUC, 0.025, na.rm = TRUE),
    quantile(boot_detect$Proteomics_AUC, 0.025, na.rm = TRUE),
    quantile(boot_detect$Combined_AUC, 0.025, na.rm = TRUE)
  ),
  AUC_high = c(
    quantile(boot_detect$DETECT_AUC, 0.975, na.rm = TRUE),
    quantile(boot_detect$Proteomics_AUC, 0.975, na.rm = TRUE),
    quantile(boot_detect$Combined_AUC, 0.975, na.rm = TRUE)
  ),
  Spec95_low = c(
    quantile(boot_detect$DETECT_Spec95, 0.025, na.rm = TRUE),
    quantile(boot_detect$Proteomics_Spec95, 0.025, na.rm = TRUE),
    quantile(boot_detect$Combined_Spec95, 0.025, na.rm = TRUE)
  ),
  Spec95_high = c(
    quantile(boot_detect$DETECT_Spec95, 0.975, na.rm = TRUE),
    quantile(boot_detect$Proteomics_Spec95, 0.975, na.rm = TRUE),
    quantile(boot_detect$Combined_Spec95, 0.975, na.rm = TRUE)
  )
)

detect_perf <- left_join(detect_perf, detect_ci, by = "Model")

detect_delta <- data.frame(
  Comparison = c(
    "Proteomics elastic net - DETECT",
    "DETECT + proteomics elastic net - DETECT"
  ),
  Delta_AUC = c(
    mean(boot_detect$Proteomics_AUC - boot_detect$DETECT_AUC, na.rm = TRUE),
    mean(boot_detect$Combined_AUC - boot_detect$DETECT_AUC, na.rm = TRUE)
  ),
  AUC_low = c(
    quantile(boot_detect$Proteomics_AUC - boot_detect$DETECT_AUC, 0.025, na.rm = TRUE),
    quantile(boot_detect$Combined_AUC - boot_detect$DETECT_AUC, 0.025, na.rm = TRUE)
  ),
  AUC_high = c(
    quantile(boot_detect$Proteomics_AUC - boot_detect$DETECT_AUC, 0.975, na.rm = TRUE),
    quantile(boot_detect$Combined_AUC - boot_detect$DETECT_AUC, 0.975, na.rm = TRUE)
  ),
  Delta_Spec95 = c(
    mean(boot_detect$Proteomics_Spec95 - boot_detect$DETECT_Spec95, na.rm = TRUE),
    mean(boot_detect$Combined_Spec95 - boot_detect$DETECT_Spec95, na.rm = TRUE)
  ),
  Spec95_low = c(
    quantile(boot_detect$Proteomics_Spec95 - boot_detect$DETECT_Spec95, 0.025, na.rm = TRUE),
    quantile(boot_detect$Combined_Spec95 - boot_detect$DETECT_Spec95, 0.025, na.rm = TRUE)
  ),
  Spec95_high = c(
    quantile(boot_detect$Proteomics_Spec95 - boot_detect$DETECT_Spec95, 0.975, na.rm = TRUE),
    quantile(boot_detect$Combined_Spec95 - boot_detect$DETECT_Spec95, 0.975, na.rm = TRUE)
  )
)

detect_perf
detect_delta

write.csv(detect_perf, here::here(output_dir_data, "TableS7_DETECT_model_performance.csv"), row.names = FALSE)
write.csv(detect_delta, here::here(output_dir_data, "TableS7_DETECT_incremental_performance.csv"), row.names = FALSE)


detect_roc_bands <- bind_rows(
  roc_bootstrap_band(
    y_detect,
    detect_oof[, "DETECT"],
    B = 2000,
    seed = 501
  ) %>% mutate(Model = "DETECT"),
  roc_bootstrap_band(
    y_detect,
    detect_oof[, "Proteomics_elastic_net"],
    B = 2000,
    seed = 502
  ) %>% mutate(Model = "Proteomics elastic net"),
  roc_bootstrap_band(
    y_detect,
    detect_oof[, "DETECT_plus_proteomics_elastic_net"],
    B = 2000,
    seed = 503
  ) %>% mutate(Model = "DETECT + proteomics elastic net")
)
```

```{r detect_roc_plot, fig.height=5.5, fig.width=9.5, dpi=300}
roc_detect_df <- bind_rows(
  roc_curve_df(y_detect, detect_oof[, "DETECT"]) %>% mutate(Model = "DETECT"),
  roc_curve_df(y_detect, detect_oof[, "Proteomics_elastic_net"]) %>% mutate(Model = "Proteomics elastic net"),
  roc_curve_df(y_detect, detect_oof[, "DETECT_plus_proteomics_elastic_net"]) %>% mutate(Model = "DETECT + proteomics elastic net")
)
auc_labels <- detect_perf %>% mutate(Label = paste0(Model, "\nAUC ", sprintf("%.2f", AUC), " (95% CI ", sprintf("%.2f", AUC_low), "-", sprintf("%.2f", AUC_high), ")")) %>% select(Model, Label)
roc_detect_df <- roc_detect_df %>% left_join(auc_labels, by = "Model")
detect_roc_bands_plot <- detect_roc_bands %>% left_join(auc_labels, by = "Model")
roc_cols <- c("DETECT" = "grey35", "Proteomics elastic net" = "#2C6E9B", "DETECT + proteomics elastic net" = "#B23A2B")
roc_fills <- c("DETECT" = "grey75", "Proteomics elastic net" = "#A8C5D5", "DETECT + proteomics elastic net" = "#E5A49A")
line_values <- setNames(roc_cols[auc_labels$Model], auc_labels$Label)
fill_values <- setNames(roc_fills[auc_labels$Model], auc_labels$Label)
p_detect_roc <- ggplot() +
  geom_ribbon(data = detect_roc_bands_plot, aes(x = FPR, ymin = lower, ymax = upper, fill = Label), alpha = 0.20) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey70") +
  geom_step(data = roc_detect_df, aes(FPR, Sensitivity, color = Label), direction = "hv", linewidth = 1.0) +
  scale_color_manual(values = line_values) + scale_fill_manual(values = fill_values) + coord_equal() +
  labs(x = "1 - Specificity", y = "Sensitivity", color = NULL, fill = NULL, title = "RHC-confirmed PAH") +
  theme_classic(base_size = 14) +
  theme(legend.position = "right", legend.direction = "vertical", legend.text = element_text(size = 11), legend.key.width = unit(1.2, "cm"), legend.spacing.y = unit(0.35, "cm")) +
  guides(color = guide_legend(ncol = 1, byrow = TRUE), fill = guide_legend(ncol = 1, byrow = TRUE))
p_detect_roc
ggsave(here::here(output_dir_figs, "Figure5C_DETECT_ROC.png"), p_detect_roc, width = 9.5, height = 5.5, dpi = 300)
```

```{r detect_scheme, fig.height=3.8, fig.width=10, dpi=300}
detect_scheme_df <- data.frame(
  x = c(1.2, 4.7, 8.2),
  y = c(1, 1, 1),
  label = c(
    "DETECT score\n(no refitting)",
    paste0(length(common_proteins), " proteins\nnested-CV elastic net"),
    paste0(
      "DETECT score + ", length(common_proteins),
      " proteins\nDETECT forced; proteins penalized"
    )
  )
)

p_detect_scheme <- ggplot(detect_scheme_df, aes(x, y, label = label)) +
  geom_label(
    size = 4.1,
    fill = "white",
    label.size = 0.35,
    label.padding = unit(0.5, "lines")
  ) +
  annotate(
    "text",
    x = c(1.2, 4.7, 8.2),
    y = 0.45,
    label = c("Model 1", "Model 2", "Model 3"),
    size = 4.0
  ) +
  xlim(0, 9.5) +
  ylim(0.15, 1.6) +
  theme_void()

p_detect_scheme
ggsave(here::here(output_dir_figs, "Figure5A_DETECT_scheme.png"),
       p_detect_scheme, width = 10, height = 3.8, dpi = 300)
```

```{r figure5_detect, fig.height=8.5, fig.width=13, dpi=300}
Figure5 <- p_detect_scheme / (p_volcano_group_z | p_detect_roc) + plot_layout(heights = c(0.62, 1.38), widths = c(0.85, 1.15)) + plot_annotation(tag_levels = "A")
Figure5
ggsave(here::here(output_dir_figs, "Figure5_panel_DETECT.png"), Figure5, width = 13, height = 8.5, dpi = 300)
```


```{r supplementary6C_elastic_net_alpha, fig.height=5, fig.width=6.2, dpi=300}
detect_alpha_long <- bind_rows(
  detect_tuning %>% transmute(Model = "Proteomics", alpha = proteomics_alpha),
  detect_tuning %>% transmute(Model = "DETECT + proteomics", alpha = combined_alpha)
)

p_detect_alpha <- ggplot(detect_alpha_long, aes(factor(alpha), fill = Model)) +
  geom_bar(position = "dodge") +
  labs(
    x = "Selected elastic-net alpha",
    y = "Outer CV folds",
    fill = NULL,
    title = "Elastic-net tuning"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom")

p_detect_alpha
ggsave(here::here(output_dir_figs, "SupplementaryFigure6C_elastic_net_alpha.png"),
       p_detect_alpha, width = 6.2, height = 5, dpi = 300)
```

```{r supplementary_figure6_panel, fig.height=5, fig.width=13, dpi=300}
SupplementaryFigure6 <- p_volcano_group_p | p_concordance_group | p_detect_alpha
SupplementaryFigure6 <- SupplementaryFigure6 + plot_annotation(tag_levels = "A")

SupplementaryFigure6
ggsave(here::here(output_dir_figs, "SupplementaryFigure6_DETECT_supporting.png"),
       SupplementaryFigure6, width = 13, height = 5, dpi = 300)
```



# Supplementary manuscript tables

## Table S1. Final five-protein panel

```{r tableS1_final_panel}
TableS1_final_panel <- panel_coefficients %>%
  transmute(
    Protein,
    `Krakow logFC` = round(logFC, 3),
    `Moderated t` = round(t, 3),
    `P value` = signif(P.Value, 3),
    FDR = signif(adj.P.Val, 3),
    `Final coefficient` = round(Coefficient, 3),
    `OR per 1-SD NPX` = round(OR_per_1SD_NPX, 3),
    `Bootstrap selection frequency` = round(Selection_frequency, 3)
  )

knitr::kable(
  TableS1_final_panel,
  align = c("l", rep("c", ncol(TableS1_final_panel) - 1)),
  caption = "Table S1. Final five-protein RV-PA coupling panel."
)

write.csv(TableS1_final_panel,
          here::here(output_dir_data, "TableS1_final_five_protein_panel.csv"),
          row.names = FALSE)
```

## Table S2. Panel discrimination and scaling sensitivity

```{r tableS2_panel_performance}
TableS2_panel_performance <- panel_auc %>%
  mutate(
    AUC = round(AUC, 3),
    AUC_low = round(AUC_low, 3),
    AUC_high = round(AUC_high, 3)
  )

knitr::kable(
  TableS2_panel_performance,
  align = c("l", "c", "c", "c"),
  caption = "Table S2a. Discrimination of impaired RV-PA coupling."
)

TableS2_scaling <- within_scaling_results %>%
  mutate(AUC = round(AUC, 3))

knitr::kable(
  TableS2_scaling,
  align = c("l", "c"),
  caption = "Table S2b. Zurich scaling sensitivity."
)

write.csv(TableS2_panel_performance,
          here::here(output_dir_data, "TableS2_panel_performance.csv"),
          row.names = FALSE)

write.csv(TableS2_scaling,
          here::here(output_dir_data, "TableS2b_Zurich_scaling_sensitivity.csv"),
          row.names = FALSE)
```

## Table S3. Clinical correlations of the protein score

```{r tableS3_clinical_correlations}
TableS3_clinical_correlations <- clinical_cor %>%
  transmute(
    Cohort,
    Variable = Label,
    n,
    `Spearman rho` = round(rho, 3),
    `P value` = signif(p, 3),
    FDR = signif(FDR, 3)
  )

knitr::kable(
  TableS3_clinical_correlations,
  align = c("l", "l", "c", "c", "c", "c"),
  caption = "Table S3. Clinical correlations of the RV-PA coupling protein score."
)

write.csv(TableS3_clinical_correlations,
          here::here(output_dir_data, "TableS3_clinical_correlations.csv"),
          row.names = FALSE)
```

## Table S4. Adjusted association with continuous TAPSE/sPAP

```{r tableS4_adjusted_linear}
TableS4_adjusted_score_models <- adjusted_coupling_results %>%
  transmute(
    Cohort,
    Model,
    n,
    `Protein-score beta` = round(Beta, 4),
    `95% CI lower` = round(CI_low, 4),
    `95% CI upper` = round(CI_high, 4),
    `P value` = signif(p, 3),
    `Adjusted R2` = round(Adjusted_R2, 3)
  )

knitr::kable(
  TableS4_adjusted_score_models,
  align = c("l", "l", rep("c", 6)),
  caption = "Table S4. Linear regression of continuous TAPSE/sPAP on the protein score, analysed separately in Krakow and Zurich."
)

write.csv(TableS4_adjusted_score_models,
          here::here(output_dir_data, "TableS4_adjusted_score_continuous_TAPSE_sPAP.csv"),
          row.names = FALSE)
```

## Table S5. Three-year mortality analyses

```{r tableS5_mortality}
TableS5_cox <- cox_table_3y %>%
  transmute(
    Model,
    Term = term,
    HR = round(estimate, 3),
    `95% CI lower` = round(conf.low, 3),
    `95% CI upper` = round(conf.high, 3),
    `P value` = signif(p.value, 3)
  )

knitr::kable(
  TableS5_cox,
  caption = "Table S5a. Cox proportional-hazards models for 3-year all-cause mortality."
)

TableS5_cindex <- cindex_table_3y %>%
  mutate(across(c(C_index, CI_low, CI_high), ~ round(.x, 3)))

knitr::kable(
  TableS5_cindex,
  caption = "Table S5b. Three-year C-index estimates."
)

TableS5_delta <- delta_cindex_table_3y %>%
  mutate(across(c(Delta_C, CI_low, CI_high), ~ round(.x, 3)))

knitr::kable(
  TableS5_delta,
  caption = "Table S5c. Difference in C-index relative to COMPERA."
)
```

## Table S6. Protein score and NT-proBNP

```{r tableS6_ntprobnp}
TableS6_cor <- nt_cor_results %>%
  transmute(
    Cohort,
    n,
    `Spearman rho` = round(rho, 3),
    `P value` = signif(p, 3)
  )

knitr::kable(
  TableS6_cor,
  caption = "Table S6a. Correlation between the protein score and NT-proBNP."
)

TableS6_auc <- nt_auc_results %>%
  mutate(across(c(AUC, CI_low, CI_high), ~ round(.x, 3)))

knitr::kable(
  TableS6_auc,
  caption = "Table S6b. Discrimination of impaired RV-PA coupling by the protein score and NT-proBNP."
)

TableS6_delong <- nt_delong_results %>%
  mutate(
    Delta_AUC = round(Delta_AUC, 3),
    DeLong_p = signif(DeLong_p, 3)
  )

knitr::kable(
  TableS6_delong,
  caption = "Table S6c. Paired DeLong comparison of protein-score and NT-proBNP AUCs."
)

TableS6_components <- component_cindex_ci_3y %>%
  mutate(across(c(C_index, CI_low, CI_high), ~ round(.x, 3)))

knitr::kable(
  TableS6_components,
  caption = "Table S6d. Exploratory 3-year COMPERA biomarker-component analysis."
)

TableS6_component_delta <- component_delta_3y %>%
  mutate(across(c(Delta_C, CI_low, CI_high), ~ round(.x, 3)))

knitr::kable(
  TableS6_component_delta,
  caption = "Table S6e. Exploratory C-index differences in the COMPERA biomarker-component analysis."
)
```

## Table S7. DETECT and proteomics

```{r tableS7_detect}
TableS7_performance <- detect_perf %>%
  mutate(
    across(c(AUC, AUC_low, AUC_high, Sensitivity,
             Specificity_at_95_sensitivity, Spec95_low, Spec95_high),
           ~ round(.x, 3))
  )

knitr::kable(
  TableS7_performance,
  caption = "Table S7a. Repeated nested-CV performance for DETECT, proteomics, and DETECT plus proteomics."
)

TableS7_delta <- detect_delta %>%
  mutate(
    across(c(Delta_AUC, AUC_low, AUC_high, Delta_Spec95, Spec95_low, Spec95_high),
           ~ round(.x, 3))
  )

knitr::kable(
  TableS7_delta,
  caption = "Table S7b. Incremental performance relative to DETECT."
)

write.csv(TableS7_performance,
          here::here(output_dir_data, "TableS7_DETECT_model_performance.csv"),
          row.names = FALSE)

write.csv(TableS7_delta,
          here::here(output_dir_data, "TableS7_DETECT_incremental_performance.csv"),
          row.names = FALSE)
```

# Save important objects

```{r save_objects}
qs_save(list(
  data_z = data_z,
  data_p = data_p,
  clin_z = clin_z,
  clin_p = clin_p,
  DE_coupling_z = DE_coupling_z,
  DE_coupling_p = DE_coupling_p,
  DE_group_z = DE_group_z,
  DE_group_p = DE_group_p,
  p_dev = p_dev,
  z_val = z_val,
  candidate_ranking_p = candidate_ranking_p,
  candidate_proteins = candidate_proteins,
  final_panel = final_panel,
  final_panel_choice = final_panel_choice,
  stability50 = stability50,
  panel_model = panel_model,
  pred_p_dev = pred_p_dev,
  pred_p_final = pred_p_final,
  pred_z_final = pred_z_final,
  lasso_cv_summary = lasso_cv$summary,
  lasso_delta_summary = lasso_delta_summary,
  panel_auc = panel_auc,
  panel_protein_cor = panel_protein_cor,
  clinical_cor = clinical_cor,
  adjusted_coupling_results = adjusted_coupling_results,
  df_surv_z_3y = df_surv_z_3y,
  cox_table_3y = cox_table_3y,
  cindex_table_3y = cindex_table_3y,
  delta_cindex_table_3y = delta_cindex_table_3y,
  lrt_score_added_3y = lrt_score_added_3y,
  nt_cor_results = nt_cor_results,
  nt_auc_results = nt_auc_results,
  nt_delong_results = nt_delong_results,
  component_cindex_ci_3y = component_cindex_ci_3y,
  component_delta_3y = component_delta_3y,
  detect_perf = detect_perf,
  detect_delta = detect_delta,
  detect_tuning = detect_tuning
), here::here(output_dir_data, "SSc_PAH_serum_proteomics_FINAL_main_objects.qs2"))
```
